---
title: "Python Simulation Results, R > 1"
subtitle: "Doc.2 Python simulation results - evolution of Mpox to high transmissibility when spillover R > 1"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)

# colors for plotting
num_cols = 20
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The objective of this document is to process and analyze the output of a python simulation of the multi-step branching process of the adaptation of a virus with a near critical (though slightly greater than 1) $R_0$ out of a zoonotic source. 

# Loading simulation results

The output of the python simulation is stored in `python simulations/output.csv`. This is a `.csv` file containing a dataframe of statistics calculated from 1000 iterations of each of 200 parameter combinations. These parameter regimes simulated the branching process with different combinations of the spillover R, $Rwt$, of the disease and the step increase in transmissibility with a given mutation, $deltaR$. We assumed the final R of highest transmissibility would be 3, and that a beneficial mutation would emerge and transmit with a probability of 1/1000 infections ($\mu = 10^{-3}$). 

```{r load-sim-results}
results = read.csv(here("code", "project", "R greater than 1", "python simulations", "output2.csv"))

```

Now lets go ahead and plot the probability of evolution to high transmissibility.

```{r plot-pHigh}
# plot probability of evolution to high R
plot = ggplot() +
  geom_line(data = results %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

plot
```
Here we can compare this to the probability that the process does not go extinct before the second type emerges (the extinction probability for a GW process with lambda = Rwt)

```{r non-extinction probability}
#nonlinear equation solver to solve for fixed point of the generating function for the branching process

## define branching process generating function for offspring distribution
bp_singletype = function(x) {y = exp(Rwt*(x-1)) - x} #fixed point solution will be this function set to 0

## define initial conditions 
init = 1 #start with single spillover infection meep meep

## start fixed point guess
xstart = 0.5 #50 percent will go extinct so lets see

## solve for non extinction probability (1-Pextinction)
non_extinction = function(Rwt) {
  Rwt = Rwt #set Rwt
  return(1 - nleqslv(xstart, bp_singletype, method="Newton", global="none", control=list(trace=1,stepmax=2))$x) #solve for nonextinction probability
  }

# solve to plot
rwt = seq(1.01, 2, by = 0.01)
pNE = NaN
for(i in 1:length(rwt)){
  Rwt = rwt[i]
  pNE[i] = non_extinction(Rwt)
}

#solve 1-1/r0 
ap = 1-(1/rwt)

```

Now we can plot our predictions based on the non-extinction probability of a branching process starting with a single case with the spillover Rwt. 

```{r plot-pHigh-preds}
# plot probability of evolution to high R
plot = ggplot() +
  geom_line(data = results %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR)), alpha = 0.5) +
  geom_line(aes(x = rwt, y = pNE)) +
  geom_line(aes(x=rwt, y = ap), col = "red") +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

plot
```


And lets plot the number of generations at 1) emergence and 2) extinction. 

```{r plot-generations}
plot2 = ggplot() +
  geom_line(data = results, aes(x = Rwt, y = Emergent_generation, group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Generation of emergence", col = "Increase with Mutation", title = "Mean generation of emergence, spillover R > 1")

plot3 = ggplot() +
  geom_line(data = results, aes(x = Rwt, y = Extinct_generation, group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Generation of extinction", col = "Increase with Mutation", title = "Mean generation of extinction, spillover R > 1")

grid.arrange(plot2,plot3)
```

Finally, lets plot the size of the extinction events. Unfortunately I was dumb and didn't take into account that calculating the mean emergent size with NaNs requires an explicit ignorance of NaNs in python. Ugh. 

```{r plot-size}
plot4 = ggplot() +
  geom_line(data = results, aes(x = Rwt, y = log10(Extinct_size), group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Size at extinction", col = "Increase with Mutation", title = "Mean size at extinction, spillover R > 1")

plot4
```


I reran and was less stupid so here's emergence.

```{r plot-size-emergence}
plot5 = ggplot() +
  geom_point(data = results, aes(x = Rwt, y = log10(Emergent_size), group = factor(deltaR), col = factor(deltaR))) +
  geom_line(data = results, aes(x = Rwt, y = log10(Emergent_size), group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Size at emergence", col = "Increase with Mutation", title = "Mean size at emergence, spillover R > 1")

plot5

```

# Greater than but close to one

I reran the simulations but choosing the wildtype R to be closer to 1. I'll load those results here: 

```{r load-sim-2-results}
results2 = read.csv(here("code", "project", "R greater than 1", "python simulations", "output3.csv"))

```

And plot how the probability of emergence changes as a function of the number of mutations and wildtype R: 

```{r plot-pHigh-2}
# plot probability of evolution to high R
plot = ggplot() +
  geom_line(data = results2 %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

plot
```

and finally see how this compares with the approximation beign the non-extinction probability of the first type: 

```{r non-extinction-probability-2}
#nonlinear equation solver to solve for fixed point of the generating function for the branching process

## define branching process generating function for offspring distribution
bp_singletype = function(x) {y = exp(Rwt*(x-1)) - x} #fixed point solution will be this function set to 0

## define initial conditions 
init = 1 #start with single spillover infection meep meep

## start fixed point guess
xstart = 0.5 #50 percent will go extinct so lets see

## solve for non extinction probability (1-Pextinction)
non_extinction = function(Rwt) {
  Rwt = Rwt #set Rwt
  return(1 - nleqslv(xstart, bp_singletype, method="Newton", global="none", control=list(trace=1,stepmax=2))$x) #solve for nonextinction probability
  }

# solve to plot
rwt = seq(range(results2$Rwt)[1], range(results2$Rwt)[2], by = diff(range(results2$Rwt))/nrow(results2))
pNE = NaN
for(i in 1:length(rwt)){
  Rwt = rwt[i]
  pNE[i] = non_extinction(Rwt)
}

```

```{r plot-pHigh-preds}
# plot probability of evolution to high R
plot = ggplot() +
  geom_line(data = results2 %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR)), alpha = 0.5) +
  geom_point(data = results2 %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR)), alpha = 0.5) +
  geom_line(aes(x = rwt, y = pNE)) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

#save
file = here("results", "analysis", "R greater than 1", "fig_P_evolution_smallR.png")
ggsave(filename = file, plot = plot)
```

So really, the divergence is happening for R wt between 1 and 1.05. I think? that is okay?

Let's look at the other stats?

```{r plot-final-size-emergence-smallR}
plot = ggplot() + 
  geom_line(data = results2 %>% filter(deltaR != 1), aes(x = Rwt, y = Emergent_size, group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "final size when emergent", col = "Increase with Mutation", title = "Total number of cases when highly transmissible variant emerges, spillover R > 1")

plot
```

This one makes sense, because a lot of the sizes get too big and go extinct or don't get big at all, so the "sample size" when taking the mean changes. 

```{r plot-generation-emergence-smallR}


```