---
title: "Probability of Evolution out of Source for Near Critical R"
subtitle: "Probability of Mpox evolving high transmissibility when R is near 1"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)

# colors for plotting
num_cols = 20
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Plotting the probability of evolving sustained transmission when $R<1$

These models were developed in `code\project\R less than 1\AdaptiveEvolutionModel.qmd` and the function defining the analytic solution to the branching process model is stored in `code\functions\functions.R`

The following code loads the function: 

```{r load-function}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

The following figure plots the probability that a virus with some wild-type R after spillover into the human population evolves sufficient transmissibility ($R>1$0 to potentiate sustained human-human transmission.

Here, I am plotting for a range of $\Delta R$ jump sizes in transmissibility with each mutation and assuming that the rate per infection of evolving a higher transmissibility is 1/1000 infections ($10^{-3}$)

```{r subcritical}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(.05,.1,.25,.5,1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

# plot
ggplot() + 
  geom_line(data = probs.emergence.deltaR, aes(x = R0, y = P.Emergence, group = delta_R, col = delta_R)) + #plot p emergence as function of R_wt for each m
  theme_bw() + #white background with grey gridlines
  labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "probability of emergence with delta R", col = expression(Delta*R)) + #axis and title labels
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(trans = "log10", limits = c(10^-12,1)) + #log10 y scale
  scale_color_manual(values = my.colors, breaks = as.character(round(delta_Rs, digits = 2))) #because pretty

```

# Plotting the probability of evolving high transmissibility when $R>1$

The stochastic models were developed in `code\project\R greater than 1\BP_evolution_greater_than_1.qmd` and again in python in `code\project\R greater than 1\python simulations`. The function defining the approximate analytic solution to the branching process model is first introduced in `code\project\R greater than 1\python simulations\BP_evolution_greater_than_1_2.qmd` and is stored in `code\functions\functions.R`.
 