---
title: "Comparing Adaptive Evolution Models"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(nleqslv) #for solving nonlinear systems of equations
library(gridExtra) #for arrangeing plots
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
source(here("code", "functions", "archive", "functions.R")) #for previous version of adaptive mutation function

```

We set out to develop a model of adaptive evolution wherein a pathogen's basic reproductive number increases by a fixed step, $\Delta R$, with each mutation. As a result, we wanted pathogens that started with higher $R_0$ to require fewer mutations to emerge.

Previously (see "Archive" files), we formatted our adaptive evolution model to match the structure presented in @AntiaNature2003. In this model, $R_0$ between 0 and 1 was divided into $m$ intervals; the intial $R_0$ of the spillover pathogen determined how many mutations (intervals) away the pathogen was to adapatation to human-human transmission ($R_0 > 1$.) We wanted to explicitly define the number of "types" in our branching process model, which is the biological equivalent to saying there are $m$ possible phenotypes, where the $R_0$ of the $ith$ phenotype lies in the $ith$ interval. 

The function itself would define the $\Delta R$ gained with each mutation as the size of each interval, and the branching process would start with 1 infection of the type into which the initial $R_0$ fell. If this is confusing, yes, it is. That is why we changed it :)

Below is a plot of the old model, where we varied parameter $m$ to indicate the number of possible phenotypes with $R_0$ between 0 and above 1. The original code for this model can be found in `\code\functions\archive\functions.R\`. I have rewritten it here in terms of $R_wt$ and $R_adapted$ in place of $R0_1$ and $Rfinal$, respectively, but it is the same model. 

```{r archived-model, results = "hide"}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
ms = c(2,5, 10,20) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence = matrix(nrow = length(R_wts), ncol = length(ms)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(ms)){
  for(i in 1:length(R_wts)){
    probs.emergence[i,j] = pEmergence_old(m = ms[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each m,R_wt combination
  }
}

## format matrix for plotting
probs.emergence = cbind(R_wts, probs.emergence) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence) = c("R0", ms) #name the columns for easy referencing
probs.emergence = probs.emergence %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "m", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

## plot

# colors for plotting
my.colors = c("black", "red", "forestgreen", "blue", "orange", "purple") #because pretty
# plot
ggplot() + 
  geom_line(data = probs.emergence, aes(x = R0, y = P.Emergence, group = m, col = m)) + #plot p emergence as function of R_wt for each m
  theme_bw() + #white background with grey gridlines
  labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "probability of emergence with m phenotypes") + #axis and title labels
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(trans = "log10", limits = c(10^-12,1)) + #log10 y scale
  scale_color_manual(values = my.colors, breaks = c("2", "5", "10", "20")) #because pretty




```

Here, we plot the same model, but now written in terms of the step increase in $R_0$, $\Delta R$, with each mutation. 

```{r delta-r-model, results = "hide"}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = 1/(c(2,5, 10,20) - 1)#how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

## plot

# colors for plotting
my.colors = c("black", "red", "forestgreen", "blue", "orange", "purple") #because pretty
# plot
ggplot() + 
  geom_line(data = probs.emergence.deltaR, aes(x = R0, y = P.Emergence, group = delta_R, col = delta_R)) + #plot p emergence as function of R_wt for each m
  theme_bw() + #white background with grey gridlines
  labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "probability of emergence with delta R", col = expression(Delta*R)) + #axis and title labels
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(trans = "log10", limits = c(10^-12,1)) + #log10 y scale
  scale_color_manual(values = my.colors, breaks = as.character(round(delta_Rs, digits = 2))) #because pretty

```

Below, I plot the results of the model in terms of $\Delta R$ with orange dashed lines, and the corresponding curves from the old model in terms of "m" in solid colors: 

```{r Comparison, results = "hide"}
# colors for plotting
my.colors = c("black", "red", "forestgreen", "blue", "orange", "purple") #because pretty
# plot
ggplot() + 
  geom_line(data = probs.emergence, aes(x = R0, y = P.Emergence, group = m, col = m)) +
  geom_line(data = probs.emergence.deltaR, aes(x = R0, y = P.Emergence, group = delta_R, col = "new_model"), linetype = "dashed") + #plot p emergence as function of R_wt for each m
  theme_bw() + #white background with grey gridlines
  labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "comparing adpative evolution model functions") + #axis and title labels
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(trans = "log10", limits = c(10^-12,1)) + #log10 y scale
  scale_color_manual(values = my.colors, breaks = c("2", "5", "10", "20", "new_model")) #because pretty


```

Comparing the plots, we see that they are the same function, just written in terms of different parameters! Cheers! 