---
title: "Waning population immunity projections for Smallpox in the Congo, 2023-2050"
subtitle: "Doc.2 Modeling the decay of immunity to smallpox at the population level after ceasing vaccination"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure

#load functions
source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function
source(here("code", "functions", "matrix_model.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to project the birth and death rates in the DRC from 2024-2080 and predict the change in portion of individuals born during or before 1980.

# Trends in birth rates and death rates from 1980-2023

First, we load the fecundities and mortalities from `data\congo\formatted data\`. Later, we will also need the population size by age in 1980 to initialize our population vector, so I will load that, too. 

```{r load-data}
#birth rates
births = as.matrix(read_csv(here("data", "congo", "formatted data", "birth_rates.csv"))[,-1])

#death rates
deaths = as.matrix(read_csv(here("data", "congo","formatted data", "death_rates.csv"))[,-1])

#population init
init_pop = as.vector(read_csv(here("data", "congo", "formatted data", "pop_1980.csv"))$x) #note that csv file reads in as tibble, so values must be extracted from column x and saved as vector

```

For easier manipulation, I want to reformat the matrices as long form data frames: 

```{r format-demographics}
#format birth rates
births_lf = as.data.frame(births) %>% #add column for year, then pivot to long form with column indicating age and column indicating birth rates
  mutate("Year" = seq(1980:2023)) %>%
  pivot_longer(
    cols = !Year,
    names_to = "Age",
    values_to = "birth_rate"
  )

#format death rates
deaths_lf = as.data.frame(deaths) %>% #add column for year, then pivot to long form with column indicating age and column indicating death rates
  mutate("Year" = seq(1980:2023)) %>%
  pivot_longer(
    cols = !Year,
    names_to = "Age",
    values_to = "death_rate"
  )

```

Now, lets plot birth and death rates over time by age:

```{r plot-demographics}
#births
fecundity_trend = ggplot() +
  geom_line(data = births_lf, aes(x = Year, y = log10(birth_rate), group = as.factor(as.numeric(Age)), col = as.factor(as.numeric(Age)))) +
  labs(x = "Year", y = "log10 Birth Rate", title = "Fecundity by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

#deaths
mortality_trend = ggplot() +
  geom_line(data = deaths_lf, aes(x = Year, y = log10(death_rate), group = as.factor(as.numeric(Age)), col = as.factor(as.numeric(Age)))) +
  labs(x = "Year", y = "log10 Death Rate", title = "Death by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

#plot together
grid.arrange(fecundity_trend, mortality_trend)


```

While birth and death rates seem relatively stable through time, there is a slight downward trend for death rates in children and younger adults we may want to consider. 

A naive approach (but the approach I'm taking for now) is to make predictions for future age structures based on the birth and death rates from the last documented year, 2023. We can revisit this. Based on the consistency in age structure through time (see `code\data analysis\waning_population_immunity.qmd`), estimating loss of population immunity assuming consistent birth/death rates over the next 60 years may have some relevance. 

# Future predictions from matrix model 

Below, we use the `matrix_-_model` function from `code/functions/matrix_model.R` to predict the proportion of the population immune to Smallpox from 1980 - 2080. The leslie matrix from 2023 onward will be identical and populated with demographics from 2023. 

`matrix_model` takes 4 required arguments:

1) `births` - a matrix of birth rates with dimensions `nxm`, where `n` is the number of time steps to simulate over (in our case, years) and `m` is the number of age groups (for us, 101)
2) `deaths` - a matrix of death rates with dimensions `nxm` where `n` is the number of time steps to simulate over (in our case, years) and `m` is the number of age groups (for us, 101)
3) `init_pop` - a vector of size `m` containing the initial number of individuals in each of the `m` age classes, and 
4) `gens` - the `n` number of time steps to simulate over

The function returns a matrix of dimensions `mxn` with each column corresponding to the age structure of a given year. 

Our loaded demographics matrices `births` and `deaths` only contain the measured birth and death rates by age from 1980-2023. To make our projections for population size from 2024-2080 assuming constant birth and death rates, we need to add `2080-2023=57` replicate rows of the 2023 demographics for each age. We do so in the code chunk below. 

```{r matrix-setup}
year.final = 2080 #year to simulate to

# how many years (rows) must be added with estimates from 2023?
pred.years =  year.final - 2023

# replicate final row - 2023 demographics - pred.years times for births matrix
births = rbind(births, matrix(rep(births[nrow(births),], pred.years), ncol = ncol(births), byrow = T)) #copy final row of births to make the the next pred.years rows 

#replicate final row - 2023 demographics - pred.years times for deaths matrix
deaths = rbind(deaths, matrix(rep(deaths[nrow(deaths),], pred.years), ncol = ncol(deaths), byrow = T)) #copy final row of births to make the the next pred.years rows 

```

Now, we can run our matrix model to predict the change in population from 2024-2080. 

```{r run-matrix-model}
sim.years = year.final-1980
m = matrix_model(births = births, deaths = deaths, init_pop = init_pop, gens = sim.years)

```

To double-check that the predicted exponential increase from consistent birth and death rates is observed, we plot the total number of individuals across all ages each year below: 

```{r plot-total}
population_size = apply(m, 2, sum)
plot(x = 1980:(1980+sim.years), y = log10(population_size), xlab = "year", ylab = "log10 population size", bg = "black", type = "o", pch = 21)
```

That looks nearly exponential to me!

# Predicted portion of pop immune

Now, we can estimate the change in proportion of the population born during or before 1980 over time. This will be the proportion of the population that we consider immune to smallpox, either due to having received the smallpox vaccine, or having been naturally infected with smallpox during their lifetime. 

In the code chunk below, we process the output of the matrix model as a long-form data frame, `sim_population`, with each observation corresponding to an Age and Number of individuals of that age for a given year (`Age`, `Size`, `Year`.)

```{r process-population-matrix}
#formatting population matrix as age dataframe
sim_pop = cbind(1980:year.final, t(m)) %>% as.data.frame() 
colnames(sim_pop) = c("Year", as.character(0:100))


# extracting population size through time and formatting in long-form
sim_population = sim_pop %>% 
  pivot_longer( #longform for easier manipulation 
    cols = !Year,
    names_to = "Age", 
    values_to = "Size"
    ) %>% 
  as.data.frame() # make data frame, not tibble

summary(sim_population) #see variable types and information

# mutate age feature to numeric
sim_population = sim_population %>% mutate("Age" = as.numeric(Age))
```

In the code chunk below, we calculate the fraction of individuals old enough to have been alive in 1980 over the total population size each year. 

This is the proportion immune to smallpox each year, and is stored in the data frame `sim_vacc_frac` which contains three variables: `Year`; `num_immune`, the number of individuals immune; `total`, the total population size that year, and; `prop_immune`, the proportion of immune individuals in the total population. 

```{r proportion-immune-calculation}
# adding new feature to population dataframe - vacced
# binary, true if older than year-1980 and thus had received vaccine or natural infection; false if younger and not immune.
# indicated whether immune
sim_population = sim_population %>%
  mutate("Vacced" = case_when(Age >= Year - 1980 ~ TRUE, .default = FALSE))

# sum number category for all vacced within each year
sim_vacc_frac = sim_population %>%
  group_by(Year) %>% #perform functions within groups
  filter(Vacced == T) %>% #filter for only vaccinated ages
  summarise(num_immune = sum(Size)) #sum across all ages 

#add total population size each year to data frame
sim_vacc_frac$total = sim_population %>%
  group_by(Year) %>%
  summarise(total = sum(Size)) %>%
  pull("total")

#calculate fraction immune
sim_vacc_frac = sim_vacc_frac %>%
  mutate("prop_immune" = num_immune/total) #yahoo

```

Finally, we plot the estimated proportion immune given a matrix model simulation from 1980-2023 estimates of birth and death rates in black, and the projected proportion immune given a matrix model simulation from 2023 estimates of birth and death rates in red. 

```{r plot-predictions}
ggplot() + 
  geom_point(data = sim_vacc_frac %>% filter(Year <= 2023), aes(x = Year, y = (prop_immune)), col = "black") + geom_line(data = sim_vacc_frac, aes(x = Year, y = (prop_immune)), col = "black") + 
  geom_point(data = sim_vacc_frac %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") + geom_line(data = sim_vacc_frac %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") +
  labs(x = "Year", y = "Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023 and projected to 2080") +
  theme_bw()  

```

# Saving output

Finally, I am going to save the output of this model to  `results\data analysis\congo`. I will save the output of the matrix model simulation as a .csv file, and the formatted dataframe containing the proportion of the population immune through time as a .csv file. 

```{r save-results}
## filenames
file1 = here("results", "data analysis", "congo", "population_matrix.csv") #path for matrix model output
file2 = here("results", "data analysis", "congo", "proportion_immune.csv") #path for calculated fractions immune

## save
write.csv(m, file = file1) #REMINDER that this is an mxn matrix, with each row corresponding to an age and each column corresponding to a year
write.csv(sim_vacc_frac, file = file2) #in long form



```