---
title: "Exploring waning population immunity to Smallpox in the Congo"
subtitle: "Doc.1 How to best model the decay of immunity to smallpox at the population level after ceasing vaccination"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure

#load source code
source(here("code", "functions", "analysis_functions.R")) #for formatting multiple excel sheets

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to explore models of the decay of immunity to smallpox and related pox viruses at the population level after vaccination ceased (1980) in the Democratic Republic of the Congo. 

Two methods will be 1) predicting fecundities and mortality from trends in demographics and using an age-structured matrix model to estimate the decay in the proportion of individuals immune, and 2) fitting a decay model to population size through time. 

# Approach 

We use a Leslie-matrix age-structured model to capture the birth of susceptible individuals and loss of immune individuals in the DRC from 1980-2023. 

We assume that all individuals in all age classes were vaccinated against or immune to smallpox by the time of smallpox eradication (and the subsequent cessation of vaccination campaigns.)  

Fecundity and survivorship estimates by single-year age are taken from the WHO [Department of Economic and Social Affairs, Population Division](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=Population).

# Loading data

The datasets for this analysis can be found in the excel file `data/congo/congo_demographics_1980-2023.xlsx`. This file contains three sheets: 

1) the number of individuals of each age in 1980 (census),
2) mortality by age from 1980-2023, and
3) fecundity by age from 1980-2023.

Note that fecundity is given as the average number of children per 1000 women. We will necessarily need to transform these estimates to return the number of fertile offspring (female) per 1 woman. We will return to this. 

The following code chunk imports these three sheets as 3 independent data frames in a single list: 

```{r load-data}
congo_demographics = multiplesheets(here("data", "congo", "congo_demographics_1980-2023.xlsx"))

```

Here, we take two processing steps. 

First, mortality is reported in the number of deaths (in thousands) per age. For death rate in each age group, I need to divide the number of deaths by the total number in that age group in the population. As both the number of individuals at each and the number of deaths at each age are given per 1000, I can simply take the ratio of dead:population. 

```{r death-rates}
# extract mortality dataframe from congo_demographics
deaths= congo_demographics$mortality %>% #longform for easier manipulation
  pivot_longer(
    cols = !Year,
    names_to = "Age", 
    values_to = "Deaths"
  ) %>%
  mutate("Age" = case_when(Age == "100+" ~ "100", .default = Age)) #rename 100+ age to 100 

# extract total number of individuals at each age from population dataframe in congo_demographics
population = congo_demographics$population %>%
  pivot_longer(
    cols = !Year, 
    names_to = "Age",
    values_to = "Size"
  ) 

# combine data frames by Age and Year 
deaths = inner_join(deaths,population, by = c("Year", "Age"))

# use nrows to check all rows are kept
deaths = deaths %>%
  mutate("mortality" = Deaths/Size)

ggplot(data = deaths) +
  geom_line(aes(x = Year, y = log10(mortality), color = as.factor(as.numeric(Age)), group = as.factor(as.numeric(Age)))) + 
  labs(x = "Year", y = "log10 Mortality Rate", title = "Mortality by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()
```

We see some decline in mortality over time, which is a good sign that we've got these right. For late projections, we might use this information to make predictions about mortality rates in the future. 

Fecundity is given as the average number of children per 1000 women. We will necessarily need to transform these estimates to return the number of fertile offspring (female) per 1 woman; we can accomplish this by dividing by $2 \times 1000$.


```{r birth-rates}
# extract fecundity dataframe from congo_demographics
births = congo_demographics$fertility
# find the minimum and maximum ages 
min_age = min(as.numeric(colnames(births)), na.rm = T)
max_age = max(as.numeric(colnames(births)), na.rm = T)
add_ages = as.character(c(0:(min_age-1), (max_age+1):100))

# add columns of zero fecundity for non-fertile ages
births[add_ages] = 0


# format df to lonform
births = births %>% pivot_longer(
    cols = !Year,
    names_to = "Age", 
    values_to = "Births"
  )

# mutate births to obtain fertile offspring per woman 
births = births %>%
  mutate("Births" = Births/(2*1000))

#plot
ggplot(data = births) +
  geom_line(aes(x = Year, y = (Births), color = as.factor(as.numeric(Age)), group = as.factor(as.numeric(Age)))) + 
  labs(x = "Year", y = "log10 Birth Rate", title = "Fecundity by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

```

Note that birth rate will be considered 0 for ages 0-14 and 50-100, as is assumed in the report. 
Also note that both the population size, death rate, and fecundity dataframes have 4444 rows, corresponding to each Age-Year combination. 

For our leslie matrix, we will sadly want to collapse these back to long form. 

```{r formating-parameter-matrices}
# matrix of birth rates. Rows are year, columns are age
births = births %>% pivot_wider(
  names_from = Age,
  values_from = Births
)
#order columns by increasing age and save as matrix
births = births[,as.character(0:100)] %>% as.matrix() #ensures the correct order of columns

# matrix of death rates. Rows are year, columns are age
deaths = deaths %>% 
  select(Year, Age, mortality) %>% pivot_wider(
  names_from = Age, 
  values_from = mortality
)
#order columns by increasing age and save as matrix
deaths = deaths[, as.character(0:100)] %>% as.matrix() #ensures the correct order of columns by age

# isolate initial population structure in 1980 as vector
init_pop = population %>% filter(Year == 1980) %>% pull(Size) #be sure that pop is in order of ages. Sizes are in thousands. 


#save these to formatted data
write.csv(births, file = here("data", "congo", "formatted data", "birth_rates.csv"))
write.csv(deaths, file = here("data", "congo", "formatted data", "death_rates.csv"))
write.csv(init_pop, file = here("data", "congo", "formatted data","pop_1980.csv"))
```


# Age structure in 1980, 2000, and 2023

```{r age-structure-plots}
### create df of population age distribution in 1980, 2000, and 2023
age_structure = data.frame("Age" = 0:100, "pop_1980" = m[, (1980-1980)+1], "pop_2000" = m[,(2000-1980)+1], "pop_2023" = m[,(2023-1980)+1])
### note that I don't have male and female ratios here, so I'll assume 50/50 
p_1980 = ggplot() + # 1980
  geom_col(data = age_structure, aes(x = pop_1980/2, y = factor(Age), fill = "male")) +
  geom_col(data = age_structure, aes(x = -pop_1980/2, y = factor(Age), fill = "female")) +
  scale_fill_manual(values = c("red", "blue")) + 
  labs(x = "Population size (in thousands)", y = "Age", fill = "Sex", title = "Age Structure, 1980") +
  scale_y_discrete(breaks = seq(0,100, 10), labels = seq(0, 100, 10)) + 
  scale_x_continuous(breaks=c(-600,-300, 0 , 300, 600), labels=c("600", "300", "0", "300", "600")) +
  theme_bw()

p_2000 = ggplot() + # 2000
  geom_col(data = age_structure, aes(x = pop_2000/2, y = factor(Age), fill = "male")) +
  geom_col(data = age_structure, aes(x = -pop_2000/2, y = factor(Age), fill = "female")) +
  scale_fill_manual(values = c("red", "blue")) + 
  labs(x = "Population size (in thousands)", y = "Age", fill = "Sex", title = "Age Structure, 2000") +
  scale_y_discrete(breaks = seq(0,100, 10), labels = seq(0, 100, 10)) + 
  scale_x_continuous(breaks=c(-900, -600,-300, 0 , 300, 600, 900), labels=c("900", "600", "300", "0", "300", "600", "900")) +
  theme_bw()

p_2023 = ggplot() + # 2023
  geom_col(data = age_structure, aes(x = pop_2023/2, y = factor(Age), fill = "male")) +
  geom_col(data = age_structure, aes(x = -pop_2023/2, y = factor(Age), fill = "female")) +
  scale_fill_manual(values = c("red", "blue")) + 
  labs(x = "Population size (in thousands)", y = "Age", fill = "Sex", title = "Age Structure, 2000") +
  scale_y_discrete(breaks = seq(0,100, 10), labels = seq(0, 100, 10)) + 
  scale_x_continuous(breaks=c(-2000, -1000, -500, -100, 0 , 100, 500, 1000, 2000), labels=c("2000", "1000", "500", "100", "0", "100", "500", "1000", "2000")) +
  theme_bw()

# plot together
grid.arrange(p_1980, p_2000, p_2023, nrow = 3)

```

From the age structure, it looks like the Congo has maintained rapid growth from 1980 - present, suggesting that the population has maintained high fecundity. 

