---
title: "heatmaps"
format: html
editor: visual
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv) #for solving nonlinear equations

# colors for plotting
num_cols = 30
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

## plotting the total probability of emergence

The total probability of emergence (of a highly transmissible disease, $R \approx 3$) in a given year is a function of the probability of emergence per spillover ($p$) and the rate of spillovers per year ($S$). The number of spillovers per year is the rate of spillovers multiplied by the unit time, 1 year, such that $n = S(1) = S$. Treating disease emergence as a bernoullie process, the total probability of emergence in a year can be calculated as:

$$ 
P = 1 - (1-p)^{n}
$$

$p$ is a function of the initial reproductive number after spillover, $R_{spill}$; the mate at which adaptive mutations are aquired per-infection, $\mu$; and, importantly, the proportional increase in fitness conferred by a beneficial mutation, $\sigma$. We note that changes in the reproductive number of a variant due to non-pharameutical interventions that change contact rates (such as social distancing) can affect both $S$ and $\sigma$. We thus plot a series of heatmaps showing the total probability of emergence over the course of a year given a range of mutation rates and spillover R for various combinations of $S$ and $\sigma$.

## load functions 

We will need two functions: a function to calculate the probability of emergence per spillover ($p$) and a function that calculates the total probability of emergence in a given interval of time, $P$. 

We load these two functions from the file `\code\functions\functions.R`.

```{r load-functions}
source(here("code", "functions", "functions.R"))

```

# setting our range of parameters

```{r params}
## probability per spillover parameters
R_wts = seq(0.5, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
sigmas = c(0.02, 0.05, 0.1)
mus = 10^seq(-6,-2, by = 0.1) # and mu

## total emergence parameters
Ss = c(100, 1000) #spillovers/year
years = 1 #time interval in which to calculate probability

```

In the code chunk below, we generate `r length(Ss)*length(sigmas)` different heatmaps that plot the total probability of emergence over the course of `r years` years given a range of $R_{spill}$ and $\mu$. Each of the 6 plots corresponds with one combination of $\sigma$ and $S$.


```{r generate-plots}
plots = lapply(sigmas, function(sigma){
         lapply(Ss, function(S){
          # set parms for total p
            S = S #spillover rate
          
            ## matrix of ps
            ps = sapply(mus, function(mu){ #iterates over all spillover R and mu combinations - outermost sapply is columns
                    sapply(R_wts, function(R_spill){
                    p = pEmergence_supercrit_deltaR(delta_R = sigma, mu = mu, R_wt = R_spill, R_adapted = R_adapted) # generates a matrix containing probabilities of emergence per spillover
                  })
                })
          # solve for total emergence probability 
            P = apply(ps, c(1,2), function(p) {total_P(p = p, S = S, years = years)}) %>% as.data.frame() # generates a matrix containing Total emergence probabilities from total_P calculation, converts to data frame
            colnames(P) = mus # columns are mus
            
          # create a plot
            ## transforms data frame to long form
            P_long = P %>% pivot_longer(cols = colnames(P),
                                        names_to = "mu",
                                        values_to = "P") %>%
              cbind("Rspill" = R_wts, .) %>%
              as.data.frame()
            
            ## plots long form df
            plot = ggplot() + geom_tile(data = P_long, aes(x = Rspill, y = log10(as.numeric(mu)), fill = log10(P))) +
                      labs(x = "R spill", y = "mu", fill = "P", title = paste0("sigma is ", sigma, " and S is ", S)) + 
                      scale_x_continuous(breaks = seq(0.5, 1, by = 1)) +
                      scale_y_continuous(breaks = seq(-6, -2, by = 1)) +
                      scale_fill_gradientn(colours = my.colors, na.value = "cadetblue") +
                      geom_contour(data = P_long, aes(x = Rspill, y = as.numeric(log10(as.numeric(mu))), z = log10(P)), breaks = c(-2,-1), col = 'maroon')
                      theme_classic()
            
            return(plot)
            
        })
  })

#returns a list of length(S) lists containing length(sigmas) elements
```

```{r plot-plots}
plots_list = do.call(c, plots) #concatenates list of lists into a single list
do.call(grid.arrange, plots_list) #plots all lists together 
```

