---
title: "Probability of Evolution out of Source for Near Critical R"
subtitle: "Probability of Mpox evolving high transmissibility when R is near 1"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)

# colors for plotting
num_cols = 20
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Plotting the probability of evolving sustained transmission when $R<1$

These models were developed in `code\project\R less than 1\AdaptiveEvolutionModel.qmd` and the function defining the analytic solution to the branching process model is stored in `code\functions\functions.R`

The following code loads the function: 

```{r load-function}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

The following figure plots the probability that a virus with some wild-type R after spillover into the human population evolves sufficient transmissibility ($R>1$0 to potentiate sustained human-human transmission.

Here, I am plotting for a range of $\Delta R$ jump sizes in transmissibility with each mutation and assuming that the rate per infection of evolving a higher transmissibility is 1/1000 infections ($10^{-3}$)

```{r subcritical}
## parameters
R_wts = seq(0.01, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(.02,.05,.1,.25, 1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

# plot
plot = ggplot() + 
          geom_line(data = probs.emergence.deltaR, aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R), linewidth = 2) + #plot p emergence as function of R_wt for each m
          theme_bw() + #white background with grey gridlines
          labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "probability of emergence with delta R", col = expression(sigma)) + #axis and title labels
          scale_x_continuous(breaks = seq(0.5, 1, by = 0.1)) +
          coord_cartesian(ylim = c(-8,0), xlim = c(0.6, 1)) +
          scale_color_manual(values = my.colors, breaks = as.character(round(delta_Rs, digits = 2))) #because pretty

plot
```

# Plotting the probability of evolving high transmissibility when $R>1$

The approximation asserts that the non-extinction probability of a multi-type BP starting with a wildtype R and evolving towards an $ R >= 3 $, where each intermediate increases in transmissibility by some factor $\Delta R$ and where the final type has a negligibly small extinction probability. (we can talk about the time it would take to reach that transmissibility later).

The stochastic models were developed in `code\project\R greater than 1\BP_evolution_greater_than_1.qmd` and again in python in `code\project\R greater than 1\python simulations`. The function defining the approximate analytic solution to the branching process model is first introduced in `code\project\R greater than 1\python simulations\BP_evolution_greater_than_1_2.qmd` and is stored in `code\functions\functions.R`.
 
The function we need is also loaded from `code\functions\functions.R`. 

```{r supercritical}
# solve to plot
rwt = seq(1.00, 2, by = 0.01)
delta_Rs = c(.02,.05,.1,.25, 1) 

### solve for non-extinction probabilities
pNE_multitype = sapply(delta_Rs, function(y) {
  sapply(rwt, function(x) {pEmergence_supercrit_deltaR(delta_R = y, mu = 10^-3, R_wt = x, R_adapted = 10000)})
  }
) %>%
  as.data.frame()

# add wildtype Rs and change the column names
pNE_multitype = cbind(rwt, pNE_multitype) %>% as.data.frame() #add a column for wildtype Rs
colnames(pNE_multitype) = c("R_wt", as.character(delta_Rs)) # add column names for referencing

# finally pivot to longform
pNE_multitype = pNE_multitype %>% pivot_longer(
  cols = !R_wt,
  names_to = "delta_R",
  values_to = "Psubcritical_evolution"
)

colnames(pNE_multitype)[1] = "R0"

```

# plot together: 

The probability of emergence of a highly transmissible virus is approximately the non-extinction probability of a single-type BP starting with a single infection of the first type (wildtype) R when the wildtype R is greater than 1. 

The probability of emergence of a highly transmissible virus when the wildtype R is slightly subcritical requires that the probability of emergence be multiplied by the non-extinction probability of the first supercritical variant. 

I am working on that function here: 

```{r probs-emergence}
## parameters
R_wts = seq(0.75, 1, by = 0.001) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(.02,.05,.1,.25, 1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", round(delta_Rs, digits = 2)) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

probs.emergence.deltaR$delta_R = as.numeric(probs.emergence.deltaR$delta_R)


ggplot() +
  geom_line(data =probs.emergence.deltaR, aes(x = R0, y = log10(P.Emergence), group = factor(delta_R), col = factor(delta_R)))
```

The code chunk below uses the `supercrit_R()` function loaded from `functions.R` to calculate the first supercritical R for a lineage starting with some wildtype R and evolving by some $\Delta R$ with each mutation. 

The probability of evolution to a highly transmissible R is then calculated for each supercritical R by the `non_extinction()` approximation in `functions.R`. 

```{r probs-evolution}
# add first supercritical R to DF
pHigh.Transmission = probs.emergence.deltaR %>%
  mutate("supercrit.R" = supercrit_R(Rwt=R0, deltaR=delta_R))

# calculate non-extinction probability of BP starting with that R
pHigh.Transmission$pNE = apply(pHigh.Transmission, MARGIN = 1,FUN = function(x) { pEmergence_supercrit_deltaR(delta_R = x["delta_R"], mu = 10^-3, R_wt = x["supercrit.R"], R_adapted = 10000)})

```

We can multiply the probability of a subcritical R emerging with a single infection of the first supercritical type by adaptive evolution by the probability that a single-type BP starting witht he first supercritical R evolves high transmissibility. This will give the probability that an infection with a subcritical R evolves high transmissibility.

```{r subcritical-high-transmission}
## calculate subcritical evolution probability
pHigh.Transmission = pHigh.Transmission %>%
  mutate("Psubcritical_evolution" = pNE * P.Emergence)

###add r greater than 1
pEvolution = rbind(pHigh.Transmission %>% select("R0", "delta_R", "Psubcritical_evolution"), pNE_multitype) %>% as.data.frame()

```

Finally, we can put together our plot with both subcritical and supercritical Rs, showing the probability that an infection with some near-critical R evolves high transmissibility:

```{r total-probs}
### 
plot = ggplot() +
  geom_line(data = pEvolution, aes(x = R0, y = log10(Psubcritical_evolution), group = factor(delta_R), col = factor(delta_R))) +
  scale_y_continuous(limits = c(-5, 0)) +
  scale_x_continuous(limits = c(.7,1.25)) +
  labs(x = "Rwt", y = "Probability of High Transmissibility", title = paste("Mu", mu)) +
  scale_color_manual(values = my.colors) +
  #geom_vline(xintercept = 1) +
  theme_classic()


plot
```

# Single function

Now, we would like to create a single function that would return this figure...

I think (?) I can use the same BP model ... let me try. 

```{r single-function-attempt}
rwt = seq(0.5, 1.5, by = 0.05)
delta_Rs = c(0.02, 0.05, 0.1, 0.25, 1)
  
  ### solve for non-extinction probabilities
pNE_multitype = sapply(delta_Rs, function(y) {
  sapply(rwt, function(x) {pEmergence_supercrit_deltaR(delta_R = y, mu = 10^-3, R_wt = x, R_adapted = 10000)})
  }
) %>%
  as.data.frame()

# add wildtype Rs and change the column names
pNE_multitype = cbind(rwt, pNE_multitype) %>% as.data.frame() #add a column for wildtype Rs
colnames(pNE_multitype) = c("R_wt", as.character(delta_Rs)) # add column names for referencing

# finally pivot to longform
pNE_multitype = pNE_multitype %>% pivot_longer(
  cols = !R_wt,
  names_to = "delta_R",
  values_to = "pNE_mtbp"
)

```

and we can plot these curves below. The solid lines are the branching process solutions, and the dotted lines are the probability (R < 1 evolves to be R > 1) x probability (R > 1 evolves to be R > 3). Note the slight discrepancy.

```{r plot-prob-evolve-high-transmission}
plot = ggplot() + #plotting the results for less than 1 with the results for greater than 1
  geom_line(data = pNE_multitype, aes(x = R_wt, y = log10(pNE_mtbp), group = delta_R, col = delta_R), linewidth = 1.5) +
  #geom_line(data = pEvolution %>% filter(R0 <= 1), aes(x = R0, y = log10(Psubcritical_evolution), group = factor(delta_R), col = factor(delta_R)), linetype = "dotted") +
  theme_bw() +
  coord_cartesian(ylim = c(-6, 0)) +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = expression(sigma), title = "Probability of evolving high transmissibility")+
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        legend.title = element_text(size = 16), # Legend title font size
        legend.text = element_text(size = 14),  # Legend key labels font size
        plot.title = element_text(size = 20, face="bold"))    # Plot title font size)

#file = here("results", "analysis", "fig_p_evolution_Rs.jpg")
#ggsave(filename = file, plot = plot)
```

