---
title: "Probability of Evolution out of Source for Near Critical R"
subtitle: "Probability of Mpox evolving high transmissibility when R is near 1"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv) #for solving nonlinear equations

# colors for plotting
num_cols = 30
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to plot the threshold transmissibility at which viruses with different characteristic wildtype transmissibilities, $Rwt$, and number of mutations to reach supercriticality, set by the increase in transmission with each mutation $deltaR$, are unlikely to emerge. We call this $R_{threshold}$, or $R_T$. 

We would then like to plot how the fraction of the population which must be vaccinated to reduce the probability of emergence below this threshold varries as a function of $Rwt$ and $deltaR$. 

# Threshold R 

For a given increase in transmissibility with mutation and mutation rate, $\DeltaR$ and $\mu$, we can plot the probability that the virus achieves sustained human-human transmission (i.e $R>1$) through progressive mutations as a function of the initial spillover $R$. 

We developed a function to return the non-extinction probability of this branching process in the file `code\project\R less than 1\AdaptiveEvolutionModel.qmd` and the function defining the analytic solution to the branching process model is stored in `code\functions\functions.R`.

We load this functions file below: 

```{r load-functions}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

The following figure plots the probability that a virus with some wild-type R after spillover into the human population evolves sufficient transmissibility ($R>1$0 to potentiate sustained human-human transmission.

Here, I am plotting for a range of $\Delta R$ jump sizes in transmissibility with each mutation and assuming that the rate per infection of evolving a higher transmissibility is 1/1000 infections ($10^{-3}$) or 1/10000 infections ($10^{-4}$.

```{r plot-subcritical}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(1, 0.25, 0.1, 0.05)
mus = c(10^-3, 10^-4) #mutation rate

emergence_curves = list() #storage list

for(k in 1:length(mus)){ #creating a seperate dataframe, probs.emergence, for each mu
  mu = mus[k] #set mu
  ## matrix to store emergence probabilities 
  probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

  ## simulate
  for(j in 1:length(delta_Rs)){
    for(i in 1:length(R_wts)){
      probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
    }
  }
  
  ## format matrix for plotting
  probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
  colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
  emergence_curves[[k]] = probs.emergence.deltaR %>% #pivot to long-form
    pivot_longer(cols = !R0, #keep R_wt column
                 names_to = "delta_R", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

}

## plot
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[1]))) + #plot p emergence as function of R_wt for each delta R
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(Delta*R), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    coord_cartesian(ylim = c(-8, 0))+
    #scale_y_continuous(limits = c(-13,0)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))

for(i in 2:length(emergence_curves)){
  plot = plot +  geom_line(data = emergence_curves[[i]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[i])), lwd = 1, alpha = 0.75)  #plot p emergence as function of R_wt for each delta r and mu
}

#print plot and save to results 
print(plot)

# save 
file = here("results", "analysis", "R less than 1", "fig_scheme_threshold_R.jpg")
ggsave(filename = file, plot = plot)
  

```

The sharp discontinuities in the probability of emergence are introduced when the R of the wild-type spillover infection is large enough that one less mutation is required for emergence, dramatically increasing the probability of emergence. 

Vaccination at the population level reduces the number of susceptibles that can be infected, effectively reducing the reproductive number of an infeciton. Our goal is to see how much vaccination would be required to reduce the probability of emergence to something below 1/10000 or 1/100000 introductions (below, dotted line). 

```{r plot-plan}
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[1]))) + #plot p emergence as function of R_wt for each delta R
    geom_segment(aes(y = -5, yend = -9, 
                 x = emergence_curves[[1]] %>% filter(delta_R == 0.25, signif(P.Emergence,1) == 10^-5) %>% slice(1) 
                 %>% pull(R0) %>% as.numeric(), 
                 xend = emergence_curves[[1]] %>% filter(delta_R == 0.25, signif(P.Emergence,1) == 10^-5) %>% slice(1) 
                 %>% pull(R0) %>% as.numeric()), 
                 linetype = "dotted")  +
    geom_segment(aes(y = -5, yend = -5, 
                 x = 0, 
                 xend = emergence_curves[[1]] %>% filter(delta_R == 0.25, signif(P.Emergence,1) == 10^-5) %>% slice(1) 
                 %>% pull(R0) %>% as.numeric()), 
                 linetype = "dotted")  +
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(Delta*R), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    coord_cartesian(ylim = c(-8, 0)) +
    #scale_y_continuous(limits = c(-13,0)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))


#print plot and save to results 
print(plot)

signif(emergence_curves[[1]]$P.Emergence,1) == 10^-5
```

Below, we can plot how this threshold R changes as a function of the two parameters defining the probability of emergence per R, $\Delta R$ and $\mu$. 

```{r prep-Rt-heatmap}
### define parms
deltaRs = seq(0.05, 0.15, by = 0.01) # define vectors of deltaR 
mus = 10^seq(-6,-2, by = 0.1) # and mu

Rwts = seq(0, 1, by = 0.01) # define vector of Rwts for candidate RTs
R_adapted = 1000 # define final R s.t. verry low probability of extinction if virus emerges

threshold = 10^-5 # define threshold prob of emergence that we tolerate

### define empty matrix in which to store solutions
RTs = matrix(nrow = length(deltaRs), ncol = length(mus))

### solve for approximate RTs
for(i in 1:length(deltaRs)){ #iterate over deltaRs
  for(j in 1:length(mus)){ #iterate over mus
    Pemergence = sapply(Rwts, function(x) pEmergence_deltaR(delta_R = deltaRs[i], mu = mus[j], R_wt = x, R_adapted = R_adapted)) #solve for emergence probabilities for each WT R
    RT = Rwts[which.min(abs(Pemergence-threshold))] #find R whos Pemergence is closest to threshold
    RTs[i,j] = RT #store in DF
  } #might think about splitting this one across cores
}

## make longform 
RTs.long = as.data.frame(cbind(deltaRs, RTs)) #bind deltaRs as a column for feature referencing
colnames(RTs.long) = c("deltaR", mus) #name the columns so referencing is easier

RTs.long = RTs.long %>% #pivot to long form
  pivot_longer(cols = !deltaR,
               names_to = "mu", 
               values_to = "RT")
#matrix still saved as RTs
```

```{r plot-Rt-heatmap}
### plot
plot = ggplot() + 
  geom_tile(data = RTs.long, aes(x = log10(as.numeric(mu)), y = deltaR, fill = RT)) +
  scale_x_continuous(breaks = seq(-6, -2, by = 1)) +
  labs(x = "log10 mu", y = "delta R", fill = "RT", title = "Threshold R to prevent emergence by evolution") +
  scale_fill_gradientn(colours = my.colors) +
  theme_classic()

# save 
file = here("results", "analysis", "R less than 1", "fig_threshold_R.jpg")
ggsave(filename = file, plot = plot)
  
```

man these colors are bad but we'll stick with it for now because i can distinguish things well enough.

# Fraction to vaccinate

The function to calculate the fraction to vaccinate to reduce R to some threshold below 1 is given in `code\functions\functions.R`.

```{r calculate-vacc-frac}
# define Reffectives
Reffs = c(0.95,1.05,1.15,3) #if R of mpox was one of these, how much do we need to vaccinate given a range of characteristics of the virus to reduce R below the Rt given by those characteristics?

# return a list of matrices with the fv to reduce wt R in Reffs below threshold RTs given in RTs
list.fvs = lapply(Reffs, function(x) apply(RTs, MARGIN = c(1,2), FUN = function(i) {frac_vacc(Rt = i, Reff = x)})) ## calls fraction to vaccinate function from "functions"

# lets turn each of these into a long-form data frame and use grid.arrange to plot them
list.fvs.long = lapply(list.fvs, function(x) {fvs.long = as.data.frame(cbind(deltaRs, x)) #bind deltaRs as a column for feature referencing
      colnames(fvs.long) = c("deltaR", mus) #name the columns so referencing is easier
      
      fvs.long = fvs.long %>% #pivot to long form
        pivot_longer(cols = !deltaR,
                     names_to = "mu", 
                     values_to = "fv")
      return(fvs.long)
      }
      )
names(list.fvs.long) = as.character(Reffs) #renames the elements of list of plots
```

```{r plot-vacc-frac}
# greats a list of grobs for ggplot objects     
list.fvs.heatmaps = lapply(list.fvs.long, function(df) {
  plot_title = names(list.fvs.long)[which(lapply(list.fvs.long, function(x) identical(x, df)) == T)] #function after title figures out the index of the df in the list of dfs and sets the title to the title of that df in the list
  plot = ggplot() + geom_tile(data = df, aes(x = log10(as.numeric(mu)), y = deltaR, fill = fv)) + 
                labs(x = "mu", y = "deltaR", fill = "fraction", title = plot_title) + 
                scale_x_continuous(breaks = seq(-6, -2, by = 1)) +
                scale_fill_gradientn(colours = my.colors, limits = c(0,1), na.value = "cadetblue") +
                theme_classic()
  return(plot)
})

#arrange all plots in grid
plot = do.call(grid.arrange, c(list.fvs.heatmaps, top = "Fraction to vaccinate"))

# save 
file = here("results", "analysis", "R less than 1", "fig_threshold_vacc.jpg")
ggsave(filename = file, plot = plot)
  

```
