---
title: "Probability of Evolution out of Source for Near Critical R"
subtitle: "Probability of Mpox evolving high transmissibility when R is near 1"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)

# colors for plotting
num_cols = 20
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Plotting the probability of evolving sustained transmission when $R<1$

These models were developed in `code\project\R less than 1\AdaptiveEvolutionModel.qmd` and the function defining the analytic solution to the branching process model is stored in `code\functions\functions.R`

The following code loads the function: 

```{r load-function}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

The following figure plots the probability that a virus with some wild-type R after spillover into the human population evolves sufficient transmissibility ($R>1$0 to potentiate sustained human-human transmission.

Here, I am plotting for a range of $\Delta R$ jump sizes in transmissibility with each mutation and assuming that the rate per infection of evolving a higher transmissibility is 1/1000 infections ($10^{-3}$)

```{r subcritical}
## parameters
R_wts = seq(0.0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(.02,.05,.1,.25, 1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

# plot
plot = ggplot() + 
          geom_line(data = probs.emergence.deltaR, aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R)) + #plot p emergence as function of R_wt for each m
          theme_bw() + #white background with grey gridlines
          labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "probability of emergence with delta R", col = expression(Delta*R)) + #axis and title labels
          scale_x_continuous(breaks = seq(0.5, 1, by = 0.1)) +
          coord_cartesian(ylim = c(-8,0), xlim = c(0.5, 1)) +
          scale_color_manual(values = my.colors, breaks = as.character(round(delta_Rs, digits = 2))) #because pretty

plot
```

# Plotting the probability of evolving high transmissibility when $R>1$

The approximation asserts that the non-extinction probability of a single-type BP starting with a wildtype R of the first type will be the probability of emergence of the highly transmissible type (we can talk about the time it would take to reach that transmissibility later).

The stochastic models were developed in `code\project\R greater than 1\BP_evolution_greater_than_1.qmd` and again in python in `code\project\R greater than 1\python simulations`. The function defining the approximate analytic solution to the branching process model is first introduced in `code\project\R greater than 1\python simulations\BP_evolution_greater_than_1_2.qmd` and is stored in `code\functions\functions.R`.
 
The function we need is also loaded from `code\functions\functions.R`. 

```{r supercritical}
# solve to plot
rwt = seq(1.01, 2, by = 0.01)
pNE = NaN #vector to store nonextinction probs in 
for(i in 1:length(rwt)){
  pNE[i] = non_extinction(Rwt = rwt[i])
}


#plot
# plot probability of evolution to high R / nonextinction probability for supercritical R singletype
plot = ggplot() +
  geom_line(aes(x = rwt, y = pNE)) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

plot

```

# plot together: 

The probability of emergence of a highly transmissible virus is approximately the non-extinction probability of a single-type BP starting with a single infection of the first type (wildtype) R when the wildtype R is greater than 1. 

The probability of emergence of a highly transmissible virus when the wildtype R is slightly subcritical requires that the probability of emergence be multiplied by the non-extinction probability of the first supercritical variant. 

I am working on that function here: 

```{r probs-emergence}
## parameters
R_wts = seq(0.02, 1, by = 0.02) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(0.1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", round(delta_Rs, digits = 2)) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

probs.emergence.deltaR$delta_R = as.numeric(probs.emergence.deltaR$delta_R)

ggplot() + geom_point(data = probs.emergence.deltaR, aes(x = R0, y = log10(P.Emergence)))

```

```{r probs-evoltuion}
# calculate first supercritical R for each starting type 
supercrit_R = function(Rwt, deltaR){
    m = ceiling(abs(log10(Rwt + 10^-12))/log10(1+deltaR)) #number of types in the branching process #note that we need a slight overshoot of R0 = 1 to get supercritical
   ### define R0 vector 
   super_R = Rwt*(1+deltaR)^(m)
  return(super_R) # add the changes from wtR0 to give R0 of first supercritical variant
}

# add first supercritical R to DF
pHigh.Transmission = probs.emergence.deltaR %>%
  mutate("supercrit.R" = supercrit_R(Rwt=R0, deltaR=delta_R))

# calculate non-extinction probability of BP starting with that R
pNEs = NaN
for(i in 1:nrow(pHigh.Transmission)) {
  pNEs[i] = non_extinction(Rwt = pHigh.Transmission$supercrit.R[i])
}

# add to df 
pHigh.Transmission = pHigh.Transmission %>%
  mutate("pNE" = pNEs)

ggplot() + geom_line(data = pHigh.Transmission, aes(x = R0, y = log10(pNE))) + 
  geom_line(data = pHigh.Transmission,aes(x = R0, y = log10(P.Emergence)))

```

We can multiply the probability of a subcritical R emerging with a single infection of the first supercritical type by adaptive evolution by the probability that a single-type BP starting witht he first supercritical R evolves high transmissibility. This will give the probability that an infection with a subcritical R evolves high transmissibility.

```{r subcritical-high-transmission}
## calculate subcritical evolution probability
pHigh.Transmission = pHigh.Transmission %>%
  mutate("Psubcritical_evolution" = pNE * P.Emergence)

###add r greater than 1
pEvolution = rbind(pHigh.Transmission %>% select("R0", "Psubcritical_evolution"), data.frame("R0" = rwt, "Psubcritical_evolution" = pNE)) %>% as.data.frame()

```

Finally, we can put together our plot with both subcritical and supercritical Rs, showing the probability that an infection with some near-critical R evolves high transmissibility:

```{r total-probs}
### 
plot = ggplot() +
  geom_line(data = pEvolution, aes(x = R0, y = log10(Psubcritical_evolution))) +
  scale_y_continuous(limits = c(-18, -.1)) +
  scale_x_continuous(limits = c(.7,1.25)) +
  labs(x = "Rwt", y = "Probability of High Transmissibility", title = paste("Mu", mu, "deltaR", delta_Rs[1])) +
  theme_classic()


plot
```

# Single function

Now, we would like to create a single function that would return this figure...