---
title: "Poster Figures"
subtitle: "EEID Notre Dame 2025"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)

# colors for plotting
num_cols = 20
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to combine work developed in `code/data analysis` and `code/project` to generate a cohesive picture of the probability of virus evolution out of zoonotic reservoirs and into new species populations. In particular, we are interested in the emergence of zoonoses closely related to viruses that have been eradicated from a population and for which population-level immunity is decreasing.

This document will be divided into three sections: 

1) Visualizing the loss of population-level immunity to smallpox over time and consequences for R0
2) Demonstrating the probability that a virus can evolve high transmissibility when near critical, as the true basic reproductive number of Mpox is unknown but likely close to 1. 
3) Showing the effect of vaccination on the probability of emergence of highly transmissible disease.

All functions for branching processes and their non-extinction probabilities can be found in the file `code/functions/functions.R` and are loaded below. 

```{r load-analytical-functions}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

All functions required for data analysis and interpretation can be found in the file `code/functions/analysis_functions.R` and are loaded below. 

```{r load-data-analysis-functions}
source(here("code", "functions", "analysis_functions.R"))

```
# Visualizing the loss of population-level immunity to Smallpox and its consequences 

After the eradication of Smallpox and the end of smallpox vaccinations in 1980, the proportion of the population with immunity to smallpox has been slowly decaying. We used a matrix model parameterized by human demographics data reported by the World Health Organization to capture the loss of smallpox immunity at the population level in the Democratic Republic of the Congo from 1980-2023. We then observed how the effective reproductive number of Mpox could be expected to increase as the cross-protection conferred by smallpox immunity is lost. 

## Loss of population level immunity to Smallpox

We observed that the demographics of the Congo have shown little change since 1980, and the country is still undergoing exponential growth. We predicted the continued loss of population level immunity to Smallpox using the matrix model parameterized by the demographics from 2023. 

In the code chunk below, we load the model estimates and projections. These were originally developed in `code/data analysis/congo/matrix_model_projections_2.qmd` and are saved to `results/data analysis/congo/proportion_immune.csv`

```{r load-decaying-immunity}
proportion_immune = read.csv(here("results", "data analysis", "congo", "proportion_immune.csv"))
```

Plotted below are the model estimates of the fraction of the population of the DRC with smallpox immunity from 1980-2023 (black) and projected until 2080 (red.)

```{r plot-decaying-immunity}
plot = ggplot() + 
            geom_point(data = proportion_immune %>% filter(Year <= 2023), aes(x = Year, y = (prop_immune)), col = "black") + geom_line(data = proportion_immune, aes(x = Year, y = (prop_immune)), col = "black") + 
            geom_point(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") + geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") +
            labs(x = "Year", y = "Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023 \nand projected to 2080") +
            theme_bw()  
# save 
file = here("results", "poster figs", "fig_decaying_immunity.png")
ggsave(filename = file, plot = plot)
```

## The consequences of decaying Smallpox immunity at the population level for the transmissibility of Mpox

One concern that arises from this loss of population immunity is the potential for increased transmission in closely related zoonoses that have spilled into the human population. One such related zoonotic virus is Mpox. While the basic reproductive number of Mpox is unknown (difficulty in estimating it arises from obstacles such as heterogeneity in transmission between subpopulations or between outbreaks, remaining population immunity from Smallpox history, and unknown cross-protection against transmission conferred by mpox), we can plot the effective reproductive number of mpox over time relative to it's basic reproductive number, that is, $\frac{R_e}{R_0}$. 

The increase in $\frac{R_e}{R_0}$ as the fraction of the population with smallpox immunity $f_i$ decays is given as $\frac{R_e}{R_0} = 1-cf_I$, where $c$ represents the cross-protection against mpox transmission conferred by smallpox immunity. 

Below, calculate $\frac{R_e}{R_0}$ for Mpox considering different degrees of cross-protection conferred by Smallpox immunity. 

```{r calculate-relative-Re}
#cross-protection
cs = seq(0.5, 1, by = 0.1) #half protected to completely protected

#calculate
##create a matrix with year and predicted Re_r0 for each c. 
relative.effective.R = sapply(cs, FUN = function(j) { #columns will be cs, rows will be year
  sapply(proportion_immune$prop_immune, FUN = function(i) {
    #set cross-reactive fraction immune equal to c*f
    Re_R0_ratio(j, i)})}) #applying re_r0_ratio function with each c to proportion immune

#add year to resulting matrix, format as dataframe, and rename features
relative.effective.R = cbind(proportion_immune$Year, relative.effective.R) %>% as.data.frame()
colnames(relative.effective.R) = c("Year", cs)

#make longform for ease of plotting 
relative.effective.R = relative.effective.R %>% 
  pivot_longer(cols = !Year, 
               names_to = "cross.immunity", 
               values_to = "relative.Re")
```

Now, we plot $\frac{R_e}{R_0}$ as a function of time for different degrees of cross-protection conferred by Smallpox immunity. 

```{r plot-relative-Re}
plot = ggplot() + 
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
  scale_color_manual(values = my.colors) +
  labs(x = "Year", y = "relative effective R, Re/R0", title = "increase in Mpox reproductive number") +
  theme_bw()

#save 
file = here("results", "poster figs", "fig_increase_relative_R.png")
ggsave(filename = file, plot = plot)

```

Finally, we might be interested in plotting both together: 

```{r plot-Re-f-together}
plot = ggplot() + 
  #plotting relative effective R
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
  #plotting proportion immune
  geom_point(data = proportion_immune %>% filter(Year <= 2023), aes(x = Year, y = (prop_immune)), col = "black") + geom_line(data = proportion_immune, aes(x = Year, y = (prop_immune)), col = " black") + 
            geom_point(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") + geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") +
  scale_color_manual(values = my.colors) +
  scale_y_continuous(name = "fraction of population immune to Smallpox", sec.axis = sec_axis(~., name = "relative effective R, Re/R0")) +
  labs(x = "Year", title = "increase in Mpox reproductive number") +
  theme_bw()

#save
file = here("results", "poster figs", "fig_increase_R_with_f.png")
ggsave(filename = file, plot = plot)
```


# Probability to evolution of high transmissibility

We consider the probability that a highly virulent pathogen ($R_e = 3$ can evolve during tranmission after a spillover Mpox infection with some near critical $R_e$.

We consider 3 probabilities: 

1) The probability that a **highly transmissible** variant emerges from a spillover infection with a **slightly supercritical** infection; 
2) the probability that a **slightly supercritical** infection emerges from a **slightly subcritical** infection, and;
3) that a **slightly subcritical** infection gives rise to a slightly supercritical infectiona and subsequently evolves to **high transmissibility**. 

## Highly transmissible evolution from slightly supercritical spillover infection

We ran stochastic simulations of a branching process model of virus evolution from one slighlty supercritical infection stepwise to an infection with a highly transmissible reproductive number (these simulations can be found in `code\project\R greater than 1\python simulations`). 

After analysing this data in `code\project\R greater than 1\EvolutionRGreaterThan1.qmd`, we find that the nonextinction probability of the first infection is a good approximation for the probability of evolution to high transmissibility - except for very small R ($ R_E < 1.05$, where the approximation underestimates emergence probability on account of the fact that ultimate extinction may occur for the first type but not before the second type emerges...could it be the sum of the non-extinction probabilities for each type... or something like this?)

We load the python simulation results below: 
```{r load-supercritical-evolution}
results = read.csv(here("code", "project", "R greater than 1", "python simulations", "output3.csv"))
```

and calculate the approximate solutions: 

```{r calculate-supercritical-evolution}
# sequnce of simulated Rwt
Rwt = seq(range(results$Rwt)[1], range(results$Rwt)[2], unique(results$Rwt)[2]-unique(results$Rwt)[1])

# calculate nonextinction probabilities
pNE = sapply(Rwt, non_extinction)
```

Below, we plot the non-extinction probabilities for the supercritical wild-type Rs with the probabilities of evolving high transmissibility from each wild-type R observed from the simulated branching processes: 

```{r plot-supercritical-evolution}
# plot probability of evolution to high R
plot = ggplot() +
  geom_line(data = results %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR)), alpha = 0.5) +
  geom_point(data = results %>% filter(deltaR != 1), aes(x = Rwt, y = pEmergent, group = factor(deltaR), col = factor(deltaR)), alpha = 0.5) +
  geom_line(aes(x = Rwt, y = pNE)) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

#save
file = here("results", "poster figs", "fig_P_evolution_supercritR.png")
ggsave(filename = file, plot = plot)
```

## Evolution from slightly subcritical spillover infection to supercritical (emergence of sustained transmission)

Below, we analytically solve for the non-extinction probability of a branching process starting with one infection with an $R_e$ less than 1 and evolving stepwise to an $R_e$ greater than 1, at which point the transmissibility is high enough to consider persistence inevitable. With each mutation, the $R_e$ increases by the percentage indicated by $\DeltaR$. 

This model was developed initially in the archived file `code\project\R less than 1\AdaptiveEvolutionModelDevelopment.qmd`, and further explored in `code\project\R less than 1\ Evolution_near_critical_R.qmd`. 


relative.effective.R = sapply(cs, FUN = function(j) { #columns will be cs, rows will be year
  sapply(proportion_immune$prop_immune, FUN = function(i) {
    #set cross-reactive fraction immune equal to c*f
    Re_R0_ratio(j, i)})}) #applying re_r0_ratio function with each c to proportion immune


```{r calculate-subcritical-emergence}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(1, 0.1, 0.05, 0.02)
mus = c(10^-3,10^-4) #mutation rate

#applying the pEmergence_deltaR function loaded from functions.R to each parameter combiantion
emergence_curves = lapply(mus, function(mu) { #creating a seperate dataframe, probs.emergence, for each mu, and storing in list emergence curves
  ### simulate 
  probs.emergence.deltaR = sapply(delta_Rs, function(deltaR) { #delta Rs are columns 
    sapply(R_wts, function(Rwt) { #R_wts are rows
      pEmergence_deltaR(delta_R = deltaR, mu = mu, R_wt = Rwt, R_adapted = R_adapted)
    })
  })

  ## format matrix for plotting
  probs.emergence.deltaR = cbind(rep(mu, nrow(probs.emergence.deltaR)), R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
  colnames(probs.emergence.deltaR) = c("mu", "R0", round(delta_Rs, digits = 2)) #name the columns for easy referencing
  emergence_curves = probs.emergence.deltaR %>% #pivot to long-form
    pivot_longer(cols = !c(R0,mu), #keep R_wt column
                 names_to = "delta_R", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column
  }
) # output of nested applies will be length(mus) dataframes, each giving the emergence probabilities of each Rwt and delta R combination for some mu
```


```{r plot-subcritical-emergence}
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[1])), linewidth = 2) + #plot p emergence as function of R_wt for each delta R
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(sigma), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    coord_cartesian(ylim = c(-8, 0), xlim = c(0.6,1))+
    #scale_y_continuous(limits = c(-13,0)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))

for(i in 2:length(emergence_curves)){
  plot = plot +  geom_line(data = emergence_curves[[i]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[i])), lwd = 1, alpha = 0.75)  #plot p emergence as function of R_wt for each delta r and mu
}

plot
# save
#file = here("results", "poster figs", "fig_P_emergence_subcritical.png")
#ggsave(filename = file, plot = plot)

```

## Evolution from slightly subcritical or supercritical spillover infection to highly transmissible

Now, we plot the probability that a subcritical spillover infection not only becomes supercritical, but subsequently evolves to be highly transmissible. Thus, the probability that a subcritical spillover evolves to become highly transmissible during human-human transmission is the product of the probabilities that a supercritical infection arises from the subcritical infection chain and the probablity that this first supercritical variant does not go extinct before evolving high transmissibility. This is plotted with the probability that a spillover variant that is slighlty supercritical evolves to be highly transmissible. 

We do this for multiple $\DeltaRs$, on which the supercritical variants aren't dependent but the subcritical are. 

In the code chunk below, we calculate the probability of 

```{r calculate-subcritical-to-transmissible}
##### calculating subcritical to highly transmissible evolution
## parameters
R_wts = seq(0.1, 0.98, by = 0.02) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(0.5, 0.1, 0.05, 0.02)
mu = 10^-3

#applying the pEmergence_deltaR function loaded from functions.R to each parameter combiantion
### calculate subcritical to supercritical emergence
probs.emergence.deltaR = sapply(delta_Rs, function(deltaR) { #delta Rs are columns 
  sapply(R_wts, function(Rwt) { #R_wts are rows
    pEmergence_deltaR(delta_R = deltaR, mu = mu, R_wt = Rwt, R_adapted = R_adapted)
  })
})
 

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", round(delta_Rs, digits = 2)) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

#### calculating supercritical to highly transmissible probability
# add first supercritical R to DF
probs.emergence.deltaR = probs.emergence.deltaR %>%
    mutate("supercrit.R" = supercrit_R(Rwt = R0, deltaR = as.numeric(delta_R)))

# calculate non-extinction probability of BP starting with that R
pNEs = NaN
for(i in 1:nrow(probs.emergence.deltaR)) {
  pNEs[i] = non_extinction(Rwt = probs.emergence.deltaR$supercrit.R[i])
}

# add to df 
probs.emergence.deltaR = probs.emergence.deltaR %>%
  mutate("pNE" = pNEs)

#finally, calculate subcritical to highly transmissble emergence probability
probs.emergence.deltaR = probs.emergence.deltaR %>% 
  mutate("Psubcritical_evolution" = pNE * P.Emergence)
# output of nested applies will be length(mus) dataframes, each giving the emergence probabilities of each Rwt and delta R c

######################### add r greater than 1
# solve to plot
rwt = seq(1.01, 2, by = 0.01)
pNE = NaN #vector to store nonextinction probs in 
for(i in 1:length(rwt)){
  pNE[i] = non_extinction(Rwt = rwt[i])
}

#format in long form for plotting
supercrit = cbind(rep(rwt, length(delta_Rs)), rep(delta_Rs, each = length(rwt)), rep(pNE, length(delta_Rs))) %>% as.data.frame()

#add colnames
colnames(supercrit) = c("R0", "delta_R", "Psubcritical_evolution")

# add plotting information to new df
pEvolution = rbind(probs.emergence.deltaR %>% select("R0", "delta_R", "Psubcritical_evolution"), supercrit) %>% as.data.frame()

```

And now we plot the probability of evolving high transmissibility in the two regimes together: 

```{r plot-subcritical-to-transmissible}
### plot together
plot = ggplot() + 
    geom_line(data = pEvolution %>% filter(R0 <= 1.01), aes(x = R0, y = log10(Psubcritical_evolution), group = delta_R, col = delta_R, linetype = as.character(mus[1]))) +  #plot p emergence as function of R_wt for each delta R
    theme_bw() + #white background with grey gridlines
  geom_line(data = pEvolution %>% filter(R0 >= 1.01), aes(x = R0, y = log10(Psubcritical_evolution), group = delta_R, linetype = as.character(mus[1])), col = "black") +
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(sigma), linetype = expression(mu)) + #axis and title labels
    coord_cartesian(ylim = c(-8, 0), xlim = c(0.5,1.25))+
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))

plot

```

# Revist last section if you can analytically calculate emergence probability for the supercritical regime

# The effect of vaccination on preventing the emergence of highly transmissible disease

The sharp decreases in the probability of evolution to high transmission as more mutations are required for emergence suggests that small amounts of vaccination can significantly reduce the probability of disease emergence. 

If we consider a 1 in 100,000 (10^-5) probability of emergence tolerable, then we can calculate the reproductive number for each parameter regime below which emergence is less than this threshold. We will call this the "threshold R", or $R_T$. 


```{r plot-plan}
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[1]))) + #plot p emergence as function of R_wt for each delta R
    geom_segment(aes(y = -5, yend = -9, 
                 x = emergence_curves[[1]] %>% filter(delta_R == 0.25, signif(P.Emergence,1) == 10^-5) %>% slice(1) 
                 %>% pull(R0) %>% as.numeric(), 
                 xend = emergence_curves[[1]] %>% filter(delta_R == 0.25, signif(P.Emergence,1) == 10^-5) %>% slice(1) 
                 %>% pull(R0) %>% as.numeric()), 
                 linetype = "dotted")  +
    geom_segment(aes(y = -5, yend = -5, 
                 x = 0, 
                 xend = emergence_curves[[1]] %>% filter(delta_R == 0.25, signif(P.Emergence,1) == 10^-5) %>% slice(1) 
                 %>% pull(R0) %>% as.numeric()), 
                 linetype = "dotted")  +
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(Delta*R), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    coord_cartesian(ylim = c(-8, 0)) +
    #scale_y_continuous(limits = c(-13,0)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))


#print plot and save to results 
print(plot)

signif(emergence_curves[[1]]$P.Emergence,1) == 10^-5

##save scheme
file = here("results", "poster figs", "fig_RT_scheme.png")
ggsave(filename = file, plot = plot)
```

Plotted below is a heatmap of $R_T$ as a function of the mutation rate and the increase in $\Delta R$ with each mutation: 

```{r caculate-Rt-heatmap}
### define parms
deltaRs = seq(0.05, 0.15, by = 0.01) # define vectors of deltaR 
mus = 10^seq(-6,-2, by = 0.1) # and mu

Rwts = seq(0, 1, by = 0.01) # define vector of Rwts for candidate RTs
R_adapted = 1000 # define final R s.t. verry low probability of extinction if virus emerges

threshold = 10^-5 # define threshold prob of emergence that we tolerate

### define empty matrix in which to store solutions
RTs = matrix(nrow = length(deltaRs), ncol = length(mus))

### solve for approximate RTs
for(i in 1:length(deltaRs)){ #iterate over deltaRs
  for(j in 1:length(mus)){ #iterate over mus
    Pemergence = sapply(Rwts, function(x) pEmergence_deltaR(delta_R = deltaRs[i], mu = mus[j], R_wt = x, R_adapted = R_adapted)) #solve for emergence probabilities for each WT R
    RT = Rwts[which.min(abs(Pemergence-threshold))] #find R whos Pemergence is closest to threshold
    RTs[i,j] = RT #store in DF
  } #might think about splitting this one across cores
}

## make longform 
RTs.long = as.data.frame(cbind(deltaRs, RTs)) #bind deltaRs as a column for feature referencing
colnames(RTs.long) = c("deltaR", mus) #name the columns so referencing is easier

RTs.long = RTs.long %>% #pivot to long form
  pivot_longer(cols = !deltaR,
               names_to = "mu", 
               values_to = "RT")
#matrix still saved as RTs
```

```{r plot-Rt-heatmap}
### plot
plot = ggplot() + 
  geom_tile(data = RTs.long, aes(x = log10(as.numeric(mu)), y = deltaR, fill = RT)) +
  scale_x_continuous(breaks = seq(-6, -2, by = 1)) +
  labs(x = "log10 mu", y = "delta R", fill = "RT", title = "Threshold R to prevent emergence by evolution") +
  scale_fill_gradientn(colours = my.colors) +
  theme_classic()

# save 
file = here("results", "poster figs", "fig_RT.jpg")
ggsave(filename = file, plot = plot)
  
```

In much the same way as a heuristic can inform the extent of vaccination required to eradicate an endemic disease ($f_{eradicate} = 1-\frac{1}{R_0}$), we can observe that vaccination to prevent the emergence of a zoonotic disease with some $R_0$ in the human population requires an additional portion, to prevent the emergence during stochastic chains with subcritical R. This heuristic is given as: 

$f_{p} = 1 - \frac{R_T}{R_0}$

Below, we plot $f_p$ for each parameter combination and considering whether the $R_0$ of the zoonotic disease is slightly subcritical, slightly supercritical, or has evolved to high transmissibility in the human population: 

The function to calculate the fraction to vaccinate to reduce R to some threshold below 1 is given in `code\functions\functions.R`.

```{r calculate-vacc-frac}
# define Reffectives
Reffs = c(0.95,1.05,1.15,3) #if R of mpox was one of these, how much do we need to vaccinate given a range of characteristics of the virus to reduce R below the Rt given by those characteristics?

# return a list of matrices with the fv to reduce wt R in Reffs below threshold RTs given in RTs
list.fvs = lapply(Reffs, function(x) apply(RTs, MARGIN = c(1,2), FUN = function(i) {frac_vacc(Rt = i, Reff = x)})) ## calls fraction to vaccinate function from "functions"

# lets turn each of these into a long-form data frame and use grid.arrange to plot them
list.fvs.long = lapply(list.fvs, function(x) {fvs.long = as.data.frame(cbind(deltaRs, x)) #bind deltaRs as a column for feature referencing
      colnames(fvs.long) = c("deltaR", mus) #name the columns so referencing is easier
      
      fvs.long = fvs.long %>% #pivot to long form
        pivot_longer(cols = !deltaR,
                     names_to = "mu", 
                     values_to = "fv")
      return(fvs.long)
      }
      )
names(list.fvs.long) = as.character(Reffs) #renames the elements of list of plots
```

```{r plot-vacc-frac}
# greats a list of grobs for ggplot objects     
list.fvs.heatmaps = lapply(list.fvs.long, function(df) {
  plot_title = names(list.fvs.long)[which(lapply(list.fvs.long, function(x) identical(x, df)) == T)] #function after title figures out the index of the df in the list of dfs and sets the title to the title of that df in the list
  plot = ggplot() + geom_tile(data = df, aes(x = log10(as.numeric(mu)), y = deltaR, fill = fv)) + 
                labs(x = "mu", y = "deltaR", fill = "fraction", title = plot_title) + 
                scale_x_continuous(breaks = seq(-6, -2, by = 1)) +
                scale_fill_gradientn(colours = my.colors, limits = c(0,1), na.value = "cadetblue") +
                theme_classic()
  return(plot)
})

#arrange all plots in grid
plot = do.call(grid.arrange, c(list.fvs.heatmaps, top = "Fraction to vaccinate to prevent emergence by evolution"))

# save 
file = here("results", "poster figs", "fig_f_p.jpg")
ggsave(filename = file, plot = plot)
  

```

I would like to plot a 4x4 figure with the heatmap for $R_T$ in the upper left, then three $f_p$ figures: one for slightly subcritical, one for slightly supercritical, and one for highly transmissible.




```{r calculate-conclusion-fig}
##################################### calculate RTs
### define parms
deltaRs = seq(0.05, 0.15, by = 0.01) # define vectors of deltaR 
mus = 10^seq(-6,-2, by = 0.1) # and mu

Rwts = seq(0, 1, by = 0.01) # define vector of Rwts for candidate RTs
R_adapted = 1000 # define final R s.t. verry low probability of extinction if virus emerges

threshold = 10^-5 # define threshold prob of emergence that we tolerate

### define empty matrix in which to store solutions
RTs = matrix(nrow = length(deltaRs), ncol = length(mus))

### solve for approximate RTs
for(i in 1:length(deltaRs)){ #iterate over deltaRs
  for(j in 1:length(mus)){ #iterate over mus
    Pemergence = sapply(Rwts, function(x) pEmergence_deltaR(delta_R = deltaRs[i], mu = mus[j], R_wt = x, R_adapted = R_adapted)) #solve for emergence probabilities for each WT R
    RT = Rwts[which.min(abs(Pemergence-threshold))] #find R whos Pemergence is closest to threshold
    RTs[i,j] = RT #store in DF
  } #might think about splitting this one across cores
}

## make longform 
RTs.long = as.data.frame(cbind(deltaRs, RTs)) #bind deltaRs as a column for feature referencing
colnames(RTs.long) = c("deltaR", mus) #name the columns so referencing is easier

RTs.long = RTs.long %>% #pivot to long form
  pivot_longer(cols = !deltaR,
               names_to = "mu", 
               values_to = "RT")
#matrix still saved as RTs


############################# Calculate FVs
# define Reffectives
Reffs = c(0.95,1.1,3) #if R of mpox was one of these, how much do we need to vaccinate given a range of characteristics of the virus to reduce R below the Rt given by those characteristics?

# return a list of matrices with the fv to reduce wt R in Reffs below threshold RTs given in RTs
list.fvs = lapply(Reffs, function(x) apply(RTs, MARGIN = c(1,2), FUN = function(i) {frac_vacc(Rt = i, Reff = x)})) ## calls fraction to vaccinate function from "functions"

# lets turn each of these into a long-form data frame and use grid.arrange to plot them
list.fvs.long = lapply(list.fvs, function(x) {fvs.long = as.data.frame(cbind(deltaRs, x)) #bind deltaRs as a column for feature referencing
      colnames(fvs.long) = c("deltaR", mus) #name the columns so referencing is easier
      
      fvs.long = fvs.long %>% #pivot to long form
        pivot_longer(cols = !deltaR,
                     names_to = "mu", 
                     values_to = "fv")
      return(fvs.long)
      }
      )
names(list.fvs.long) = as.character(Reffs) #renames the elements of list of plots

```

```{r plot-conclusion-fig}
### plot RTs
plot1 = ggplot() + 
  geom_tile(data = RTs.long, aes(x = log10(as.numeric(mu)), y = deltaR, fill = RT)) +
  scale_x_continuous(breaks = seq(-6, -2, by = 1)) +
  labs(x = "log10 mu", y = "delta R", fill = "RT", title = "Threshold R to prevent emergence by evolution") +
  scale_fill_gradientn(colours = my.colors) +
  theme_classic()

### plot FVs
# greats a list of grobs for ggplot objects     
list.fvs.heatmaps = lapply(list.fvs.long, function(df) {
  plot_title = names(list.fvs.long)[which(lapply(list.fvs.long, function(x) identical(x, df)) == T)] #function after title figures out the index of the df in the list of dfs and sets the title to the title of that df in the list
  plot = ggplot() + geom_tile(data = df, aes(x = log10(as.numeric(mu)), y = deltaR, fill = fv)) + 
                labs(x = "mu", y = "deltaR", fill = "fraction", title = plot_title) + 
                scale_x_continuous(breaks = seq(-6, -2, by = 1)) +
                scale_fill_gradientn(colours = my.colors, limits = c(0,1), na.value = "cadetblue") +
                theme_classic()
  return(plot)
})

#add Rts to list of grobs
list.fvs.heatmaps = append(list.fvs.heatmaps, list(plot1), after = 0)

#arrange all plots in grid
plot = do.call(grid.arrange, c(list.fvs.heatmaps, top = "Fraction to vaccinate to prevent emergence by evolution"))


## save
file = here("results", "poster figs", "fig_conclusion.png")
ggsave(filename = file, plot = plot)

```
