---
title: "Exploring waning population immunity to Smallpox in the Congo"
subtitle: "Doc.1 How to best model the decay of immunity to smallpox at the population level after ceasing vaccination"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure

source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The objective of this document is to develop a function to simulate the multi-step branching process of the adaptation of a virus with a near critical $R_0$ out of a zoonotic source. 

# Model 


```{r bp-testfunction}
### define function
bp_testdeltaR = function(Rwt, deltaR, Rfinal, mu, gens){
  ##### calculate parameters
  # caluclate number of types
   # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    print("extinct")
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    print("emergent") 
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if emerges at high R
  }
  return(X)
}


```

```{r run-sim}
# parms
Rwt = 1.5
deltaR = 0.3 
Rfinal = 3
mu = 10^-3
gens = 1000


##### calculate parameters
  # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    print("extinct")
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    print("emergent") 
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if emerges at high R
  }

### cool 
```


```{r run-testfunction}
# parms
Rwt = 1.3
deltaR = 0.3
Rfinal = 3
mu = 10^-3
gens = 100

#run sim
N = bp_deltaR(Rwt = Rwt, deltaR = deltaR, Rfinal = Rfinal, mu = mu, gens = 100)


```

# Probably and time to emergence *per intro*

I will run the simulation 100 and calculate the fraction of emergent versus extinct spillover events. 
I will keep track of the total size of the outrbeak (sum of all infections) and the number of generations before the emergence of the highly-transmissible variant. 

```{r }
### define function
bp_deltaR = function(Rwt, deltaR, Rfinal, mu, gens){
  ##### calculate parameters
  # caluclate number of types
   # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    extinct = extinct + 1
    extinction_size = c(extinction_size, sum(X))
    extinction_gen = c(extinction_gen, gen)
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    emerged = emerged + 1
    emergence_size = c(emergence_size, sum(X))
    emergence_gen = c(emergence_gen, gen)
    break} #break if emerges at high R
  }
  return(X)
}
```

```{r sim-setup}
# Parms
spills = 1000 #number of spillovers 

Rwt = 1.3 #wt r
deltaR = 0.3 #change in r with mut  
Rfinal = 3 #r of final type 
mu = 10^-3 #mutation rate
gens = 100 #number of generations to run each sim

# Storage
## emergence
emerged = 0 #counter to keep track of emergence events
emergence_size = c(0) # vector to keep track of final sizes of emergence events
emergence_gens = c(0) # vector to keep track of generation at which final variant emerges
## extinction
extinct = 0 #counter to keep track of extinction events
extinction_size = c(0) # vector to keep track of final sizes of extinction events
extinction_gens = c(0) # vector to keep track of generation at which extinction occurs


```

```{r run-sim-perintro}
c = 1
while(c<=spills){
  #parms
   # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    extinct = extinct + 1
    extinction_size = c(extinction_size, sum(X, na.rm = T))
    extinction_gens = c(extinction_gens, gen)
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    emerged = emerged + 1
    emergence_size = c(emergence_size, sum(X, na.rm = T))
    emergence_gens = c(emergence_gens, gen)
    break} #break if emerges at high R
  }
  if(c %% 10 == 0){
  print(paste("done", c))
  }
  c = c+1
  
}

```

```{r analysis}
#### fraction emerged and extinct
p_emerge = emerged/spills
p_extinct = extinct/spills

#average number of gens to and size at emergence 
mean_emergence_gens = mean(emergence_gens)
mean_emergence_size = mean(emergence_size)

#average number of gens to and size after extinction
mean_extinction_gens = mean(extinction_gens)
mean_extinction_size = mean(extinction_size)
```


Next steps - 
run simulation for varying wild type and delta R. 

Then add 1 or 2 other mus. 

Figuring out how to do this timely is not trivial! I bet I could get away with 100 spillovers...

```{}
