---
title: "Poster Figures"
subtitle: "Summer 2025"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)

# colors for plotting
num_cols = 20
my.colors = brewer.pal(n = 8, name = "Set1")
my.colors.many = colorRampPalette(my.colors)(num_cols)
```

# Objective

The goal of this document is to combine work developed in `code/data analysis` and `code/project` to generate a cohesive understanding of the factors that most greatly contribute to the probability of virus evolution out of zoonotic reservoirs and into new species populations. In particular, we are interested in the emergence of zoonoses closely related to viruses that have been eradicated from a population and for which population-level immunity is decreasing.

This document will be divided into three sections: 

1) Visualizing the loss of population-level immunity to smallpox over time and consequences for R
2) Demonstrating and contextualizing how viruses with low transmissibility can evolve high transmissibility, and will given a significant rate of spillover into the new population. 
3) Showing the sensitivity of the probability of emergence of highly transmissible disease to reductions in the transmissibility of spillover infections, particularly due to vaccination

All functions for branching processes and their non-extinction probabilities can be found in the file `code/functions/functions.R` and are loaded below. 

```{r load-analytical-functions}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

All functions required for data analysis and interpretation can be found in the file `code/functions/analysis_functions.R` and are loaded below. 

```{r load-data-analysis-functions}
source(here("code", "functions", "analysis_functions.R"))
```

# Visualizing the loss of population-level immunity to Smallpox and its consequences 

After the eradication of Smallpox and the end of smallpox vaccinations in 1980, the proportion of the population with immunity to smallpox has been slowly decaying. We used a matrix model parameterized by human demographics data reported by the World Health Organization to capture the loss of smallpox immunity at the population level in the Democratic Republic of the Congo from 1980-2023. We then observed how the effective reproductive number of Mpox could be expected to increase as the cross-protection conferred by smallpox immunity is lost. 

## Loss of population level immunity to Smallpox

We observed that the demographics of the Congo have shown little change since 1980, and the country is still undergoing exponential growth. We predicted the continued loss of population level immunity to Smallpox using the matrix model parameterized by the demographics from 2023. 

In the code chunk below, we load the model estimates and projections. These were originally developed in `code/data analysis/congo/matrix_model_projections_2.qmd` and are saved to `results/data analysis/congo/proportion_immune.csv`

```{r load-decaying-immunity}
proportion_immune = read.csv(here("results", "data analysis", "congo", "proportion_immune.csv"))
```

Plotted below are the model estimates of the fraction of the population of the DRC with smallpox immunity from 1980-2023 (black) and projected until 2080 (red.)

```{r plot-decaying-immunity}
plot = ggplot() + 
            geom_line(data = proportion_immune %>% filter(Year <= 2024), aes(x = Year, y = (prop_immune)), col = "black", linewidth = 2) +
            geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "black", alpha = 0.5, linewidth = 2) +
            labs(x = "Year", y = "Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023 \nand projected to 2080") +
            theme(axis.text = element_text(size = 20),
                  axis.title = element_text(size = 20), 
                  legend.title = element_text(size = 20),
                  legend.text = element_text(size = 20)) +
            theme_bw()  
```

One concern that arises from this loss of population immunity is the potential for increased transmission in closely related zoonoses that have spilled into the human population. One such related zoonotic virus is Mpox. While the basic reproductive number of Mpox is unknown (difficulty in estimating it arises from obstacles such as heterogeneity in transmission between subpopulations or between outbreaks, remaining population immunity from Smallpox history, and unknown cross-protection against transmission conferred by mpox), we can plot the effective reproductive number of mpox over time relative to it's basic reproductive number, that is, $\frac{R_e}{R_0}$. 

The increase in $\frac{R_e}{R_0}$ as the fraction of the population with smallpox immunity $F_i$ decays is given as $\frac{R_e}{R_0} = 1-cF_I$, where $c$ represents the cross-protection against mpox transmission conferred by smallpox immunity. 

Below, we calculate $\frac{R_e}{R_0}$ for Mpox considering different degrees of cross-protection conferred by Smallpox immunity. 

```{r calculate-relative-Re}
#cross-protection
cs = seq(0.5, 1, by = 0.1) #half protected to completely protected

#calculate
##create a matrix with year and predicted Re_r0 for each c. 
relative.effective.R = sapply(cs, FUN = function(j) { #columns will be cs, rows will be year
  sapply(proportion_immune$prop_immune, FUN = function(i) {
    #set cross-reactive fraction immune equal to c*f
    Re_R0_ratio(j, i)})}) #applying re_r0_ratio function with each c to proportion immune

#add year to resulting matrix, format as dataframe, and rename features
relative.effective.R = cbind(proportion_immune$Year, relative.effective.R) %>% as.data.frame()
colnames(relative.effective.R) = c("Year", cs)

#make longform for ease of plotting 
relative.effective.R = relative.effective.R %>% 
  pivot_longer(cols = !Year, 
               names_to = "cross.immunity", 
               values_to = "relative.Re")
```

Now, we plot $\frac{R_e}{R_0}$ as a function of time for different degrees of cross-protection conferred by Smallpox immunity. 

```{r plot-relative-Re}
plot = ggplot() + 
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
  scale_color_manual(values = my.colors) +
  labs(x = "Year", y = "relative effective R, Re/R0", title = "increase in Mpox reproductive number") +
  theme_bw()

```

We plot the loss of smallpox immunity on the left axis and related increase in reproductive number on the right axis. This figure will be saved in `/results/paper figs`.

```{r plot-Re-f-together}
plot = ggplot() + 
  #plotting relative effective R
  geom_line(data = relative.effective.R %>% filter(Year <= 2024), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), linewidth = 2) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), linewidth = 2, alpha = 0.5) +
  #plotting proportion immune
  geom_line(data = proportion_immune %>% filter(Year <= 2024), aes(x = Year, y = (prop_immune)), col = "black", linewidth = 2) +
  geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "black", alpha = 0.5, linewidth = 2) +
  #adding labels
  scale_y_continuous(name = expression(paste("Fraction immune to Smallpox ", italic("(F)"))), sec.axis = sec_axis(~., name = expression(paste("Relative effective R ", italic((R["e"]/R["0"])))), breaks = c(0, 0.5, 1)), breaks = c(0, 0.5, 1)) +
  scale_x_continuous(breaks = c(1980,2000,2020,2040,2060,2080)) +
  labs(x = "Year", color = "Cross immunity") +
  # plot aesthetics
  theme_bw() + 
  theme(axis.text = element_text(size = 25),
        axis.text.y.right = element_text(size = 25, color = "royalblue4"),
        axis.title = element_text(size = 25),
        axis.title.y.right = element_text(size = 25, color = "royalblue4"),
        legend.title = element_text(size = 25),
        legend.text = element_text(size = 25)) +
  scale_color_manual(values = brewer.pal(n = length(cs), name = "Blues")) 


print(plot)
#save
file = here("results", "paper figs", "fig_increase_R_with_f.png")
ggsave(filename = file, plot = plot)

```


## Increase in the rate of spillovers

As immunity wanes at the population level, the number of spillovers also increases proportional to the density of susceptible individuals (as the number of susceptible human-animal interactions will decrease.) The proportion of successful of spillovers in the presence of population immunity compared to a population with no immunity would follow the same pattern as the reproductive number, that is, the rate at which the number of successful spillovers approaches the number that would occur in a completely naive population would follow the relative reproductive number in the population: 

$spill \ rate = spill \ rate_0 (1-cf_v)$

I make a simple adjustment to the previous figure to acknowledge this in case we want it later.

```{r plot-Re-f-together}
plot = ggplot() + 
  #plotting relative effective R
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
  #plotting proportion immune
  geom_point(data = proportion_immune %>% filter(Year <= 2023), aes(x = Year, y = (prop_immune)), col = "black") + geom_line(data = proportion_immune, aes(x = Year, y = (prop_immune)), col = " black") + 
            geom_point(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") + geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "red") +
  scale_color_manual(values = my.colors) +
  scale_y_continuous(name = "Fraction of population immune to Smallpox, FI", sec.axis = sec_axis(~., name = "relative effective R, R/R0 \n and relative spillover rate")) +
  labs(x = "Year", title = "increase in Mpox reproductive number and spillover rate", color = "Cross immunity, c") +
  theme_bw()

```

# Small changes in $R_{spill}$ can dramatically increase the probability of emergence of epidemic disease

The probability that a disease emerges over some unit time $t$ depends both on the rate at which the infections spill over into the new host population, $S$, and the probability that the disease emerges after a spillover, $p$, and is given as $P = p(1-p)^{n-1}$, where $n$ is the number of spillovers that occur during some unit time $t$ given the spillover rate $S$ such that $n = St$. 

Canonically, an infection with a small reproductive number after spilling over into the host population ($R_{spill} \leq 1$) may cause a few infections but will ultimately go extinct in the population. In reality, a mutation gained by the pathogen during infection may increase its reproductive number and transmit to a new host, increasing the expected number of subsequent infections and thus the probability of gaining more mutations that increase transmissibility. 

We use branching process theory to explore this process of mutation during transmission chains. In this section, we plot the probabilities that the pathogen emerges in the population with an $R > 1$ and subsequently evolves an $R /geq 3$, the estimated reproductive number of smallpox. 

Below, we plot the probability that the pathogen evolves an $R > 1$ after spilloverfor different mus: 

```{r plot-subcritical-to-supercritical-mus}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(1, 0.25, 0.1, 0.05)
mus = c(10^-3, 10^-4) #mutation rate

emergence_curves = list() #storage list

for(k in 1:length(mus)){ #creating a seperate dataframe, probs.emergence, for each mu
  mu = mus[k] #set mu
  ## matrix to store emergence probabilities 
  probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

  ## simulate
  for(j in 1:length(delta_Rs)){
    for(i in 1:length(R_wts)){
      probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
    }
  }
  
  ## format matrix for plotting
  probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
  colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
  emergence_curves[[k]] = probs.emergence.deltaR %>% #pivot to long-form
    pivot_longer(cols = !R0, #keep R_wt column
                 names_to = "delta_R", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

}

## plot
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[1]))) + #plot p emergence as function of R_wt for each delta R
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(sigma), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    coord_cartesian(ylim = c(-8, 0), xlim = c(0.2,1))+
    #scale_y_continuous(limits = c(-13,0)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))

for(i in 2:length(emergence_curves)){
  plot = plot +  geom_line(data = emergence_curves[[i]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[i])), lwd = 1, alpha = 0.75)  #plot p emergence as function of R_wt for each delta r and mu
}

#print plot and save to results 
print(plot)

# save 
file = here("results", "paper figs", "fig_subcritical_to_critical_mus.jpg")
ggsave(filename = file, plot = plot)

```

and below we plot the curves for a single mu: 

```{r plot-subcritical-to-supercritical}
## parameters
R_wts = seq(0.01, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(.02,.05,.1,.25, 1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

# plot
plot = ggplot() + 
          geom_line(data = probs.emergence.deltaR, aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R), linewidth = 2) + #plot p emergence as function of R_wt for each m
          theme_bw() + #white background with grey gridlines
          labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "probability of emergence with delta R", col = expression(sigma)) + #axis and title labels
          scale_x_continuous(breaks = seq(0.5, 1, by = 0.1)) +
          coord_cartesian(ylim = c(-8,0), xlim = c(0.6, 1)) +
          scale_color_manual(values = my.colors, breaks = as.character(round(delta_Rs, digits = 2))) + #because pretty
  # aesthetics
          theme_bw() + 
          theme(axis.text = element_text(size = 25),
                axis.title = element_text(size = 25),
                legend.title = element_text(size = 25),
                legend.text = element_text(size = 25))

# save 
file = here("results", "paper figs", "fig_subcritical_to_critical.jpg")
ggsave(filename = file, plot = plot)


```

# Emergence probability per spillover

Disease that become supercritical may nevertheless go extinct stochastically after emergence. Here, we plot the probability that a pathogen continues to spread in a population and evolves $R \geq 3$, the approximate reproductive number of smallpox. 


below we plot the probability that a pathogen with a subcritical reproductive number evolves a hihgly infectious (R = 3) reproductive number after spillover: 

```{r subcritical-to-highly-infectious}
rwt = seq(0.5, 1.5, by = 0.05)
delta_Rs = c(0.02, 0.05, 0.1, 0.25, 1)
  
  ### solve for non-extinction probabilities
pNE_multitype = sapply(delta_Rs, function(y) {
  sapply(rwt, function(x) {pEmergence_supercrit_deltaR(delta_R = y, mu = 10^-3, R_wt = x, R_adapted = 10000)})
  }
) %>%
  as.data.frame()

# add wildtype Rs and change the column names
pNE_multitype = cbind(rwt, pNE_multitype) %>% as.data.frame() #add a column for wildtype Rs
colnames(pNE_multitype) = c("R_wt", as.character(delta_Rs)) # add column names for referencing

# finally pivot to longform
pNE_multitype = pNE_multitype %>% pivot_longer(
  cols = !R_wt,
  names_to = "delta_R",
  values_to = "pNE_mtbp"
)

# plot
plot = ggplot() + #plotting the results for less than 1 with the results for greater than 1
  geom_line(data = pNE_multitype, aes(x = R_wt, y = log10(pNE_mtbp), group = delta_R, col = delta_R), linewidth = 1.5) +
  #axes and titles
  scale_x_continuous(breaks = c(0.5,1,1.5)) +
  coord_cartesian(ylim = c(-6, 0)) +
  scale_color_manual(values = my.colors) +
  labs(x = expression(paste(R[e], " of introduced pathogen")), y = "log P(evolving high R)", col = expression(sigma)) +
  #aesthetics
  theme_bw() +
  theme(axis.text=element_text(size=25),
        axis.title=element_text(size=24),
        legend.title = element_text(size = 25), # Legend title font size
        legend.text = element_text(size = 25)  # Legend key labels font size
        )    
#save 
file = here("results", "paper figs", "fig_subcritical_to_high_R.png")
ggsave(filename = file, plot = plot)



```

And now we show how this probability changes with mu: 

```{r subcrictical-to-highly-infections-mus}
rwt = seq(0.5, 1.5, by = 0.05)
mus = 10^seq(-5, -1, by = 1)
sigma = 0.1
  
  ### solve for non-extinction probabilities
pNE_multitype = sapply(mus, function(y) {
  sapply(rwt, function(x) {pEmergence_supercrit_deltaR(delta_R = sigma, mu = y, R_wt = x, R_adapted = 10000)})
  }
) %>%
  as.data.frame()

# add wildtype Rs and change the column names
pNE_multitype = cbind(rwt, pNE_multitype) %>% as.data.frame() #add a column for wildtype Rs
colnames(pNE_multitype) = c("R_wt", as.character(mus)) # add column names for referencing

# finally pivot to longform
pNE_multitype = pNE_multitype %>% pivot_longer(
  cols = !R_wt,
  names_to = "mu",
  values_to = "pNE_mtbp"
)

# plot
plot = ggplot() + #plotting the results for less than 1 with the results for greater than 1
  geom_line(data = pNE_multitype, aes(x = R_wt, y = log10(pNE_mtbp), group = factor(as.numeric(mu)), alpha = factor(as.numeric(mu))), color = "#4DAF4AFF", linewidth = 1.5) +
  #axes and titles
  scale_x_continuous(breaks = c(0.5,1,1.5)) +
  coord_cartesian(ylim = c(-6, 0)) +
  scale_alpha_manual(values = c(0.2, 0.4, 0.6, 0.8, 1 ), labels = c(expression(10^-5), expression(10^-4), expression(10^-3), expression(10^-2), expression(10^-1))) +
  labs(x = expression(paste(R[e], " of introduced pathogen")), y = "log P(evolving high R)", alpha = expression(mu)) +
  #aesthetics
  theme_bw() +
  theme(axis.text=element_text(size=25),
        axis.title=element_text(size=25),
        legend.title = element_text(size = 25), # Legend title font size
        legend.text = element_text(size = 25)  # Legend key labels font size
        )    
#save 
file = here("results", "paper figs", "fig_subcritical_to_high_R_mus.png")
ggsave(filename = file, plot = plot)




```

# Pandemic potential 

The total probability of emergence (of a highly transmissible disease, $R \approx 3$) in a given year is a function of the probability of emergence per spillover ($p$) and the rate of spillovers per year ($S$). The number of spillovers per year is the rate of spillovers multiplied by the unit time, 1 year, such that $n = S(1) = S$. Treating disease emergence as a bernoullie process, the total probability of emergence in a year can be calculated as:

$$ 
P = 1 - (1-p)^{n}
$$

$p$ is a function of the initial reproductive number after spillover, $R_{spill}$; the mate at which adaptive mutations are aquired per-infection, $\mu$; and, importantly, the proportional increase in fitness conferred by a beneficial mutation, $\sigma$. We note that changes in the reproductive number of a variant due to non-pharameutical interventions that change contact rates (such as social distancing) can affect both $S$ and $R_{spill$. We thus plot a series of heatmaps showing the total probability of emergence over the course of a year given a range of mutation rates and $\sigma$s for various combinations of $S$ and $R_{spill$.

# How does emergence chang with $S$ and $R$?

Below, we plot how the probability of emergence of highly transmissible disease changes as a function of $R_{spill}$ and $S$ for a range of $\sigma$ and $\mu$.

```{r heatmaps-S-Rspill}

#function pEmergence_supercrit_deltaR calculates the probability of evolution to R = 3 default
#function total_P calculates the probability of emergence of a length of time years given spillover rate per year and proabability of emergence per spillover. 

### define parm ranges
R_adapted = 100000
mus = c(10^-4, 10^-3) #rows of grid
sigmas = c(0.05, 0.1) #columns of grid

R_wts = seq(0.75, 1, by = 0.01) #wt R0s to test, x axis of plots
Ss = 10^seq(0,5, by = 0.1) #spillovers/year, y axis of plots 
#Ss = seq(1, 10^5, 1000) #for linear y

#generate list of plots to format as grid
plots = lapply(mus, function(mu){ #rows of grid
          lapply(sigmas, function(sigma){ #columns of grid
            ## set simulation parms for fig
            mu = mu # set mu for this plot
            sigma = sigma # set sigma for this plot
            
            ## vector of ps
            ps = sapply(R_wts, function(R_spill){ # iterates over all spillover R for a given mu and sigma 
                    p = pEmergence_supercrit_deltaR(delta_R = sigma, mu = mu, R_wt = R_spill, R_adapted = R_adapted) }) # generates a vector containing probabilities of emergence per spillover, withhighest R set to 3 as default
          
            ## calculate total P for each ps and S combination
            Ps = sapply(Ss, function(S){ #outside is columns
              sapply(ps, function(p){ #inside is rows
                P = total_P(p, S, years=1) #probability of emergence over a single year
              })
            }) %>% cbind(R_wts, .) %>% as.data.frame()
            
            # format df for plotting
            colnames(Ps) = c("Rspill", Ss) #name columns
            Ps_long = Ps %>% pivot_longer(cols = !Rspill,
                                     names_to = "S",
                                     values_to = "P") %>% as.data.frame()

            ## Plot
            plot = ggplot() + geom_tile(data = Ps_long, aes(x = Rspill, y = log10(as.numeric(S)), fill = log10(P))) +
              labs(x = expression(paste(R[e], " of introduced pathogen")), y = expression(paste("Log spillover rate ", year^-1)), title = paste0("mu is ", mu, " sigma is ", sigma), legend = "log P") +
              scale_fill_gradient2(low = ("blue"), mid = "white", high = ("red"), midpoint = -5, limits = c(-10, 0), na.value = "lightgrey") +
              geom_contour(data = Ps_long, aes(x = Rspill, y = log10(as.numeric(S)), z = log10(P)), breaks = c(-3,-1), col = 'black',linewidth = 1.5) +
              #aesthetics
              theme_classic() +
              theme(axis.text=element_text(size=24),
                    axis.title=element_text(size=24),
                    legend.title = element_text(size = 24), # Legend title font size
                    legend.text = element_text(size = 16)  # Legend key labels font size
                    )    
              
  
            return(plot)
   })
})

#plot
plots = do.call(c, (plots))
do.call(grid.arrange, c(plots, nrow = 2))

#save 
file = here("results", "paper figs", "fig_pandemic_potential.jpg")
ggsave(filename = file, plot = do.call(grid.arrange, c(plots, nrow = 2)))
```

# Change in mpox $P_E$ post-Smallpox eradication 

Below, we plot sample trajectories of $R_e$ and $S$ from 1980 as population immunity is lost. We assume complete (100%) vaccination coverage in 1980 and a reproductive number of $0.32$ in 1984 (Based on Blumberg and JLS estimation from 1980-1984 paper on fitting stuttering chains to $R_e$). $R_e$ and $S$ take the following forms: 

$R_e = R_0(1-cF)$ and
$S = S_0(1-cF)$ 

where $F$ is the fraction of the population with immunity to smallpox, $c$ is the degree of cross-protection conferred by smallpox immunity agains mpox, $R_0$ is the basic reproductive number, and $S_0$ is the spillover rate in a completely unvaccinated population. 

We can calculate $c$ by assuming different $R_0$ of mpox. We vary $R_0$ across figures and plot multiple sample trajectors of $S$ and $R_e$ for per figure varying $S_0$. 

```{r define-re-S-functions}
# functions 
calc_c = function(f, R0, Re) {(1/f)*(1-(Re/R0))} #function to calculate c given re of a certain year, f of that year, and R0 
Re_calc = function(f,R0,c) {R0*(1-c*f)}
S_calc = function(f,S0,c) {S0*(1-c*f)}

```

```{r define-parms}
# PE heatmap parms
### define parm ranges
R_adapted = 100000
mu = 10^-3 #rows of grid
sigma = 0.1 #columns of grid

R_wts = seq(0.1, 1, by = 0.01) #wt R0s to test, x axis of plots
Ss = 10^seq(0,5, by = 0.1) #spillovers/year, y axis of plots 

# Re-S trajectory parms 
## variable parms
R0_range = c(0.85, 0.95, 1.1, 1.3)
S0_range = 10^c(3,4,5)

## F_immune over time
f_immune = proportion_immune %>% select(Year, prop_immune) 

## calculating c 
Re.1984 = 0.32 #estimated effective reproductive number 
f.1984 = f_immune %>% filter(Year == 1984) %>% pull(prop_immune) # portion immune in year of estimated effective reproductive number

cs = sapply(R0_range, function(R0) calc_c(f = f.1984, Re = Re.1984, R0 = R0)) # calculating the cross-protection corresponding to each R0
```

```{r calculating-trajectories}
# calculating Re and S over time for each c and S0 combination
trajectories = lapply(R0_range, function(R0){ #iterate over each possible R0
    dfs = lapply(S0_range, function(S0){ #iterate over each S0 in a completely susceptible pop to calculate trajectories over time
        Re_S = sapply(f_immune$prop_immune, function(f){ #iterate over every year and calculate the Re and S resulting from that year's proportion immune to smallpox
          # calculate c from R0
          c = calc_c(f = f.1984, Re = Re.1984, R0 = R0)
         
           #calculate Re from f and R0
          Re = Re_calc(f = f, c = c, R0 = R0)
          
          #calculate Se from f and S0
          S = S_calc(f = f, S0 = S0, c = c)
          
          #return a vector of 2 values: Re and S
          return(c("Re" = Re, "S" = S))
        })  %>% #returns a matrix of 2 rows (Re and S) and length(prop_immune) columns corresponding to year
        t(.) %>% cbind("S0" = S0, "Year" = f_immune$Year, .) %>% as.data.frame()  #transposes array so columns are Re and S, adds column of years, and converts to DF
      return(Re_S)
    })
  #combine all dataframes of Re and S corresponding to one S0 as a long form df
  trajectories.dfs = do.call(rbind, dfs) 
})

# naming dfs by assumed R0
names(trajectories) = R0_range
```

lets create the heatmap df:

```{r df-for-heatmap-with-trajectories}
## set simulation parms for fig
mu = mu # set mu for this plot
sigma = sigma # set sigma for this plot

## vector of ps
ps = sapply(R_wts, function(R_spill){ # iterates over all spillover R for a given mu and sigma 
        p = pEmergence_supercrit_deltaR(delta_R = sigma, mu = mu, R_wt = R_spill, R_adapted = R_adapted) }) # generates a vector containing probabilities of emergence per spillover, withhighest R set to 3 as default

## calculate total P for each ps and S combination
Ps = sapply(Ss, function(S){ #outside is columns
  sapply(ps, function(p){ #inside is rows
    P = total_P(p, S, years=1) #probability of emergence over a single year
  })
}) %>% cbind(R_wts, .) %>% as.data.frame()

# format df for plotting
colnames(Ps) = c("Rspill", Ss) #name columns
Ps_long = Ps %>% pivot_longer(cols = !Rspill,
                         names_to = "S",
                         values_to = "P") %>% as.data.frame()

```

and plot the trajectories:

```{r plot-trajectories}
#plot trajectories 
plots_list = lapply(trajectories, function(trajectory){
        plot = ggplot() + geom_point(data = trajectory, aes(x = Re, y = S, group = S0)) 
        for(i in 1:length(S0_range)){
          plot = plot + 
          geom_segment(data = trajectory %>% filter(S0 == S0_range[i]), aes(x = Re, y = S, xend = c(tail(Re, n = -1),NA), yend = c(tail(S, n = -1),NA)), arrow = arrow(length = unit(0.2,"cm")))
        }
        return(plot)})

do.call(grid.arrange, plots_list)

# trajectories by 10 year intervals
plots_list = lapply(trajectories, function(trajectory){
        plot = ggplot() + geom_line(data = trajectory, aes(x = Re, y = S, group = S0)) 
        for(i in 1:length(S0_range)){
          plot = plot + 
          geom_segment(data = trajectory %>% filter(S0 == S0_range[i], Year %% 10 == 0), aes(x = Re, y = S, xend = c(tail(Re, n = -1),NA), yend = c(tail(S, n = -1),NA)), arrow = arrow(length = unit(0.2,"cm")))
        }
        return(plot)})
do.call(grid.arrange, plots_list)
```

let's put it together, and plot the heatmap together with trajectories:

```{r heatmap-with-trajectories}
#generate heatmap
heatmap_base =  
    ## Plot
    plot = ggplot() + geom_tile(data = Ps_long, aes(x = Rspill, y = log10(as.numeric(S)), fill = log10(P))) +
      scale_fill_gradient2(low = ("blue"), mid = "white", high = ("red"), midpoint = -5, limits = c(-10, 0), na.value = "lightgrey") +
      geom_contour(data = Ps_long, aes(x = Rspill, y = log10(as.numeric(S)), z = log10(P)), breaks = c(-3,-1), col = 'black', linewidth = 1.5) +
      scale_x_continuous(breaks = c(0.1, 0.5, 1.0)) +
      #aesthetics
      theme_classic() +
      theme(axis.text=element_text(size=24),
            axis.title=element_text(size=24),
            legend.title = element_text(size = 24), # Legend title font size
            legend.text = element_text(size = 16)  # Legend key labels font size
            )    

#plot trajectories 
plots_list = lapply(trajectories, function(trajectory){
      #colors for plotting 
      cols = brewer.pal(n = length(S0_range), name = "Greys")
      names(cols) = S0_range
      plot = heatmap_base  
        for(i in 1:length(S0_range)){
          plot = plot + 
          geom_segment(data = trajectory %>% filter(S0 == S0_range[i], Year %% 10 == 0), 
                       aes(x = Re, y = log10(S), xend = c(tail(Re, n = -1),NA), yend = c(tail(log10(S), n = -1),NA)), colour = cols[i], 
                       arrow = arrow(length = unit(0.2,"cm"))) +
          geom_text(data = trajectory %>% filter(S0 == S0_range[i], Year %% 10 == 0), aes(x = Re, y = log10(S), label = Year),
                     nudge_x = 0.02,
                     nudge_y = .1,
                    size = 2)
        }
      plot = plot + 
        scale_color_manual(values = cols, labels = S0_range) +
        labs(x = expression(paste(R[e], " of introduced pathogen")), y = expression(paste("Log spillover rate ", year^-1)), title = paste0(), fill = "log P", color = "S0")
      return(plot)})

#add titles

do.call(grid.arrange, plots_list)
```

There is a time scale question here, where a probability of emergence being 1 in 100 per year suggests that within the next 100 years the emergence will occur. But over the next 100 years, immunity may continue to decay, and so the true probability of emergence should be calculated by considering the integral(in some way?) of the rate per year for 1/P(Emergence) years of the given Re/S0 trajectory. That is a *time to emergence* problem and we should consider it. 


# Fraction to vaccinate to prevent disease emergence

Here we calculate the fraction of the population which must be vaccinated to prevent infectious disease emergence given a spillover $R_{spill}$ and some $R_{eff}$ below which the pathogen is unlikely to evolve sustained transmission after spillover. 

Parallel to the classic heuristic giving the fraction of the population that must be vaccinated to eradicate an endemic disease, $F_{eradication} = 1 - \frac{1}{R_0}$, the fraction of the population that must be vaccinated to prevent the evolutionary emergence of a disease can be given by:

$$
F_{prevention} = 1 - \frac{R_{prevent}}{R_{spill}}
$$

```{r calculate-f-prevention}
Rps = seq(0.8, 0.9, by = 0.01) # define R preventions
Rwts = c(seq(0.91, 1.2, by = 0.01),3)  # define R0s note the last is smallpox R

f_vaccinate = function(Rp, Rwt) { 1- Rp/Rwt} #define fraction to vaccinate function

# calculate f_vaccinate
F_prevent = sapply(Rps, function(Rp){
              sapply(Rwts, function(Rwt){
                f_vaccinate(Rp, Rwt)
              })
            })

# format for heatmap
F_prevent = cbind(Rwts, F_prevent) %>% as.data.frame() # add column to indicate Rwt for each row
colnames(F_prevent) = c("Rspill", Rps) #rename columns with Rp for referencing

# pivot to long form 
F_prevent_long = pivot_longer(F_prevent, #long form input for geom_tile
                              cols = !Rspill,
                              names_to = "Rp",
                              values_to = "Fp")



```

plot:

```{r plot-f-prevention}
# heatmap
plot = ggplot() + geom_tile(data = F_prevent_long, aes(x = Rspill, y = Rp, fill = Fp)) +
   labs(x = "R spill", y = "R prevention", fill = "Fraction to vaccinate", title = "Fraction to vaccinate to prevent evolutionary emergence") + 
   scale_x_continuous(breaks = c(seq(0.9, 1.2, by = 0.1))) +
   scale_fill_gradientn(colours = my.colors, na.value = "cadetblue") +
  theme_classic()

print(plot)
```

