---
title: "Manuscript Figures"
subtitle: "Preventing Disease Emergence Following Eradication: Application to Mpox"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(nleqslv)
library(ggstance)
library(scales)

# colors for plotting
num_cols = 20
my.colors = brewer.pal(n = 8, name = "Set1")
my.colors.many = colorRampPalette(my.colors)(num_cols)
```

# Objective

The goal of this document is to combine work developed in `code/data analysis` and `code/project` to generate a cohesive understanding of the factors that most greatly contribute to the probability of virus evolution out of zoonotic reservoirs and into new species populations. In particular, we are interested in the emergence of zoonoses closely related to viruses that have been eradicated from a population and for which population-level immunity is decreasing.

This document will be divided into three sections: 

1) Visualizing the loss of population-level immunity to smallpox over time and consequences for R
2) Demonstrating and contextualizing how viruses with low transmissibility can evolve high transmissibility, and will given a significant rate of spillover into the new population. 
3) Showing the sensitivity of the probability of emergence of highly transmissible disease to reductions in the transmissibility of spillover infections, particularly due to vaccination

All functions for branching processes and their non-extinction probabilities can be found in the file `code/functions/functions.R` and are loaded below. 

```{r load-analytical-functions}
source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
```

All functions required for data analysis and interpretation can be found in the file `code/functions/analysis_functions.R` and are loaded below. 

```{r load-data-analysis-functions}
source(here("code", "functions", "analysis_functions.R"))
```


NOTE: Code-chunks for generating figures that appear in the manuscript for SI will be labeled with their figure number. 

# Visualizing the loss of population-level immunity to Smallpox and its consequences 

After the eradication of Smallpox and the end of smallpox vaccinations in 1980, the proportion of the population with immunity to smallpox has been slowly decaying. We used a matrix model parameterized by human demographics data reported by the World Health Organization to capture the loss of smallpox immunity at the population level in the Democratic Republic of the Congo from 1980-2023. We then observed how the effective reproductive number of Mpox could be expected to increase as the cross-protection conferred by smallpox immunity is lost. 

## Figure 2: Loss of population level immunity to Smallpox

We observed that the demographics of the Congo have shown little change since 1980, and the country is still undergoing exponential growth. We predicted the continued loss of population level immunity to Smallpox using the matrix model parameterized by the demographics from 2023. 

In the code chunk below, we load the model estimates and projections. These were originally developed in `code/data analysis/congo/matrix_model_projections_2.qmd` and are saved to `results/data analysis/congo/proportion_immune.csv`

```{r load-decaying-immunity}
proportion_immune = read.csv(here("results", "data analysis", "congo", "proportion_immune.csv"))
```

Plotted below are the model estimates of the fraction of the population of the DRC with smallpox immunity from 1980-2023 (black) and projected until 2080 (red.)

```{r plot-decaying-immunity}
plot = ggplot() + 
            geom_line(data = proportion_immune %>% filter(Year <= 2024), aes(x = Year, y = (prop_immune)), col = "black", linewidth = 2) +
            geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = (prop_immune)), col = "black", alpha = 0.5, linewidth = 2) +
            labs(x = "Year", y = "Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023 \nand projected to 2080") +
            theme(axis.text = element_text(size = 20),
                  axis.title = element_text(size = 20), 
                  legend.title = element_text(size = 20),
                  legend.text = element_text(size = 20)) +
            theme_bw()  
```

One concern that arises from this loss of population immunity is the potential for increased transmission in closely related zoonoses that have spilled into the human population. One such related zoonotic virus is Mpox. While the basic reproductive number of Mpox is unknown (difficulty in estimating it arises from obstacles such as heterogeneity in transmission between subpopulations or between outbreaks, remaining population immunity from Smallpox history, and unknown cross-protection against transmission conferred by mpox), we can plot the effective reproductive number of mpox over time relative to it's basic reproductive number, that is, $\frac{R_e}{R_0}$. 

The increase in $\frac{R_e}{R_0}$ as the fraction of the population with smallpox immunity $F_i$ decays is given as $\frac{R_e}{R_0} = 1-cF_I$, where $c$ represents the cross-protection against mpox transmission conferred by smallpox immunity. 

Below, we calculate $\frac{R_e}{R_0}$ for Mpox considering different degrees of cross-protection conferred by Smallpox immunity. 

```{r calculate-relative-Re}
#cross-protection
cs = seq(0.5, 1, by = 0.1) #half protected to completely protected

#calculate
##create a matrix with year and predicted Re_r0 for each c. 
relative.effective.R = sapply(cs, FUN = function(j) { #columns will be cs, rows will be year
  sapply(proportion_immune$prop_immune, FUN = function(i) {
    #set cross-reactive fraction immune equal to c*f
    Re_R0_ratio(j, i)})}) #applying re_r0_ratio function with each c to proportion immune

#add year to resulting matrix, format as dataframe, and rename features
relative.effective.R = cbind(proportion_immune$Year, relative.effective.R) %>% as.data.frame()
colnames(relative.effective.R) = c("Year", cs)

#make longform for ease of plotting 
relative.effective.R = relative.effective.R %>% 
  pivot_longer(cols = !Year, 
               names_to = "cross.immunity", 
               values_to = "relative.Re")
```

Now, we plot $\frac{R_e}{R_0}$ as a function of time for different degrees of cross-protection conferred by Smallpox immunity. 

```{r plot-relative-Re}
plot = ggplot() + 
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
  scale_color_manual(values = my.colors) +
  labs(x = "Year", y = "relative effective R, Re/R0", title = "increase in Mpox reproductive number") +
  theme_bw()

```

We plot the loss of smallpox immunity on the left axis and related increase in reproductive number on the right axis. This figure will be saved in `/results/paper figs`.

```{r plot-Fig-2}
#plotting altogether
plot = ggplot() + 
  #plotting relative effective R
  geom_line(data = relative.effective.R %>% filter(Year <= 2024), aes(x = Year-1980, y = relative.Re, group = cross.immunity, col = cross.immunity), linewidth = 2) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year-1980, y = relative.Re, group = cross.immunity, col = cross.immunity), linewidth = 2, alpha = 0.5) +
  #plotting proportion immune
  geom_line(data = proportion_immune %>% filter(Year <= 2024), aes(x = Year-1980, y = (prop_immune)), col = "black", linewidth = 2) +
  geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year-1980, y = (prop_immune)), col = "black", alpha = 0.5, linewidth = 2) +
  #geom_vline(xintercept = 2023-1980, linetype = "dashed", linewidth = 2) +
  #adding labels
  scale_y_continuous(name = expression(paste("Fraction immune to Smallpox ", italic("(F)"))), sec.axis = sec_axis(~., name = expression(paste("Relative effective R ", italic((R["e"]/R["0"])))), breaks = c(0, 0.5, 1)), breaks = c(0, 0.5, 1)) +
  scale_x_continuous(breaks = c(1980,2000,2020,2040,2060,2080)-1980) +
  labs(x = "Years post eradication", color = "Cross immunity") +
  # plot aesthetics
  theme_bw() + 
  theme(axis.text = element_text(size = 25),
        axis.text.y.right = element_text(size = 25, color = "royalblue4"),
        axis.title = element_text(size = 25),
        axis.title.y.right = element_text(size = 25, color = "royalblue4"),
        legend.title = element_text(size = 25),
        legend.text = element_text(size = 25)) +
  scale_color_manual(values = brewer.pal(n = length(cs), name = "Blues")) 

#print to plot viewer
print(plot)
#save
file = here("results", "paper figs", "fig_increase_R_with_f.png")
ggsave(filename = file, plot = plot)

```

######### examples of plots to add ##########

# plotting the probability that a given $R_e$ evolves supercriticality $R_e > 1$. 

This does *not* take into consideration any constraint on how large $R_0$ can be of the evolved variant. 

## Probability of evolving $Re>1$ from subcritical $R_e$ of spillover 

Below, we plot the probability that the pathogen evolves an $R > 1$ after spilloverfor different mus: 

```{r plot-subcritical-to-supercritical-mus}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype such that it contributes negligibly to the extinction probability - once R = 1 is achieved P(non-extinction) is one, so we are essentially calculating the probability the process does not go extinct before R = 1 is achieved
delta_Rs = c(1, 0.25, 0.1, 0.05)
mus = c(10^-3, 10^-4) #mutation rate

emergence_curves = list() #storage list

for(k in 1:length(mus)){ #creating a seperate dataframe, probs.emergence, for each mu
  mu = mus[k] #set mu
  ## matrix to store emergence probabilities 
  probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

  ## simulate
  for(j in 1:length(delta_Rs)){
    for(i in 1:length(R_wts)){
      probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
    }
  }
  
  ## format matrix for plotting
  probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
  colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
  emergence_curves[[k]] = probs.emergence.deltaR %>% #pivot to long-form
    pivot_longer(cols = !R0, #keep R_wt column
                 names_to = "delta_R", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

}

## plot
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[1]))) + #plot p emergence as function of R_wt for each delta R
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(sigma), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    coord_cartesian(ylim = c(-8, 0), xlim = c(0.2,1))+
    #scale_y_continuous(limits = c(-13,0)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))

for(i in 2:length(emergence_curves)){
  plot = plot +  geom_line(data = emergence_curves[[i]], aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R, linetype = as.character(mus[i])), lwd = 1, alpha = 0.75)  #plot p emergence as function of R_wt for each delta r and mu
}

#print plot and save to results 
print(plot)

# save 
file = here("results", "paper figs", "fig_subcritical_to_critical_mus.jpg")
ggsave(filename = file, plot = plot)

```

and below we plot the curves for a single mu: 

```{r plot-subcritical-to-supercritical}
## parameters
R_wts = seq(0.01, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(.02,.05,.1,.25, 1) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3 #mutation rate

## matrix to store emergence probabilities 
probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

## simulate
for(j in 1:length(delta_Rs)){
  for(i in 1:length(R_wts)){
    probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
  }
}

## format matrix for plotting
probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
probs.emergence.deltaR = probs.emergence.deltaR %>% #pivot to long-form
  pivot_longer(cols = !R0, #keep R_wt column
               names_to = "delta_R", #add feature denoting "m" used row-wise
               values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

# plot
plot = ggplot() + 
          geom_line(data = probs.emergence.deltaR, aes(x = R0, y = log10(P.Emergence), group = delta_R, col = delta_R), linewidth = 2) + #plot p emergence as function of R_wt for each m
          theme_bw() + #white background with grey gridlines
          labs(x = expression("R"[e] ~ " of introduced pathogen"), y = "Probability of Emergence", title = "Non-extinction probability of process before Re > 1 (no limit on R0 of the evolved variant)", col = expression(sigma)) + #axis and title labels
          scale_x_continuous(breaks = seq(0.5, 1, by = 0.1)) +
          coord_cartesian(ylim = c(-8,0), xlim = c(0.6, 1)) +
          scale_color_manual(values = my.colors, breaks = as.character(round(delta_Rs, digits = 2))) + #because pretty
  # aesthetics
          theme_bw() + 
          theme(axis.text = element_text(size = 25),
                axis.title = element_text(size = 25),
                legend.title = element_text(size = 25),
                legend.text = element_text(size = 25))
# print plot
print(plot)

# save 
file = here("results", "paper figs", "fig_subcritical_to_critical.jpg")
ggsave(filename = file, plot = plot)

```

We might want to know, given some $R_0$ of the wildtype variant, what the probability of the whole process evolving $R_0 > 3$ is. 
One way to plot this is to plot the probability that a process starting with some $R_e$ and a given $R_0$ (where $R_{e,initial} = R_0(1-cF)$) evolves to $R_{e, evolved} = 3(1-cF)$, where $cF = 1-\frac{R_{e,initial}}{R_0}$ and we only plot for the range of $R_e$ where $cF>2/3$ (thus $R_{e, evolved} > 1$). $cF$ is the effective immunity  in the population. 

```{r plot-R0-to-high-R0}
# four plots in figure
R0_wt_vec = c(0.75, 0.85,0.95,1.05)
R0_evolved = 3 #like smallpox

## parameters
sigma_vec = c(.02, .05, 0.1, 0.25, 1) #proportional increase in Re with mutation
mu = 10^-3 #mutation rate

# list of plot dfs
dfs = lapply(R0_wt_vec, function(R0_wt){
  Re_spill_vec = seq(0.5, R0_wt, by = 0.01) #wt R0s to test
  ## simulate
  probs.emergence = 
    sapply(sigma_vec, function(sigma){
      sapply(Re_spill_vec, function(Re_spill){
        # Calculate end Re for final type in branching process
        cF = 1-(Re_spill/R0_wt) # effective amount of immunity in the population
        Re_evolved = R0_evolved*(1-cF) # effective transmission when R0 = 3 in the population
        # Caluclate non-extinction probability 
        return(pEmergence(sigma = sigma, mu = mu, Re_spill = Re_spill, Re_evolved = Re_evolved)) #run simulation for each sigma,Re_spill combination
      })
    })
  
  ## format matrix for plotting
  probs.emergence = cbind(Re_spill_vec, probs.emergence) %>% as.data.frame() #format as dataframe with column denoting the Re_spill used to generate the probabilities row-wise and column for each sigma used to generate the probabilities column-wise
  colnames(probs.emergence) = c("Re_spill", as.character(sigma_vec)) #name the columns for easy referencing
  probs.emergence = probs.emergence %>% #pivot to long-form
    pivot_longer(cols = !Re_spill, #keep R_wt column
                 names_to = "sigma", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column
  # add column indicating R0 of wildtype
  probs.emergence = cbind(probs.emergence, "R0_wt" = R0_wt)
})


# create a list of plots for grid.arrange
### plot 
plots = lapply(dfs, function(df){
  plot = ggplot() + 
  geom_line(data = df, aes(x = Re_spill, y = log10(P.Emergence), group = sigma, col = sigma), linewidth = 2) + #plot p emergence as function of Re_spill for each sigma
          theme_bw() + #white background with grey gridlines
          labs(x = expression("R"[e] ~ " of introduced pathogen"), y = "Probability of Emergence", title = bquote("R"[0] ~ "of introduced pathogen is" ~ .(unique(df$R0_wt))), col = expression(sigma)) + #axis and title labels
          scale_x_continuous(breaks = seq(0.5, 1, by = 0.1)) +
          coord_cartesian(ylim = c(-8,0), xlim = c(0.5, 1)) +
          scale_color_manual(values = my.colors, breaks = as.character(sigma_vec)) + #because pretty
  # aesthetics
          theme_bw() + 
          theme(axis.text = element_text(size = 25),
                axis.title = element_text(size = 25),
                legend.title = element_text(size = 25),
                legend.text = element_text(size = 25))
})
# print plots
do.call(grid.arrange, plots)

```
 
Nearly identical, except for the range of plotting. 

I can do a heatmap with F(t) on the x axis, R0 on the y axis, and p(R0 -> 3) filled in. 

```{r p-r-to-3}
# define vector of R0 and Frac Immune
R0_wt_vec = seq(0.75, 1.05, by = 0.001)
Frac_immune_vec = seq(0, 0.5, by = 0.01)

# define unchanging parms
sigma = 0.05 # proportional increase in fitness with mutation
mu = 10^-3 # rate of beneficial mutations per infection
cp = 0.8 #cross-reactivity
R0_evolved = 3 # max R0 evolvable

#define vector of years post eradication 
proportion_immune$time = proportion_immune$Year - 1980

# calculate
plot.df = sapply(R0_wt_vec, function(R0_wt){
  sapply(Frac_immune_vec, function(Frac_immune){
    # calculate R_spill and R_final
    Re_spill = R0_wt*(1-cp*Frac_immune) 
    Re_evolved = R0_evolved*(1-cp*Frac_immune)
    
    # Caluclate non-extinction probability 
    return(pEmergence(sigma = sigma, mu = mu, Re_spill = Re_spill, Re_evolved = Re_evolved)) #run simulation for each R0, Immunity combination
  })
})

#plot
filled.contour(x = Frac_immune_vec, y = R0_wt_vec, z=log10(plot.df), nlevels = 20, lwd = 2,
               xlab = "Fraction Immune", ylab = expression("R"[0] ~ "of Spillover"), 
               key.par = list(mar = c(4, 1, 4, 3)),
               key.title = title(main = expression("P"[E])),
               plot.axes = {
                 axis(1)
                 axis(2)
                 #arrow pointing left 
                 arrows(x1 = proportion_immune$prop_immune, 
                        x0 = c(lag(proportion_immune$prop_immune, 1)), 
                        y0 = rep(0.8, nrow(proportion_immune)), y1 = rep(0.8, nrow(proportion_immune)), length = 0.1, lwd = 2)
                 text(x = 0.4, y = 0.81, labels = "Decay over time")
               })


```


# Figure showing $P_p$ vs $F(t)$ or time for different $R_0$ 

```{r plot-Pp-over-time}
# define parms 
c = 0.8 # cross reactivity of eradicated disease with zoonotic disease
R0_evolved = 3 # R0 of most evolved variant - perhaps the R0 of the eradicated disease

R0_wt_vec = c(0.75, 0.85, 0.95, 1.05) # plotting 4 trajectories

S0 = 10^4

# Add Se over time, Re evolved, and years post eradication 
proportion_immune = proportion_immune %>%
  mutate("Se" = S0*(1-c*prop_immune)) %>% # spillover rate increases with time towards Se like Re increases towards R0 with mutation. 
  mutate("time" = Year - 1980) %>% # Add time, time post eradication in years 
  mutate("Re_evolved" = R0_evolved*(1-c*prop_immune)) # Add Re_Evolved, the effective reproductive number of the fully evolved variant 

# at which year is Re of the evolved > 1? 
min_year = proportion_immune %>%
  filter(Re_evolved > 1) %>% #all rows where Re evolved is > 1
  select(time) %>% # only want the times post vaccination the condition is true
  slice(1) %>% pull(time) # first entry will be the minimum, pull reports variable as numeric rather than df

# calculate Pp over time for each R0 
df = lapply(R0_wt_vec, function(R0_wt){
  # calculate Re of spillover from proportion immune, calculate Pe frp, proportion immune and 
  pandemic_potential = proportion_immune %>% 
    mutate("Re_spill" = R0_wt*(1 - c*prop_immune)) %>% # calculate Re spillover
    mutate("Pe" = mapply(pEmergence, sigma = sigma, mu = mu, Re_spill = Re_spill, Re_evolved = Re_evolved)) %>% #calculate PE
    mutate("Pp" = 1-(1-Pe)^Se) %>% # calculate pandemic potential
    mutate("R0" = R0_wt) # add column distinguishing R0
  # return
  return(pandemic_potential)
}) %>% do.call(rbind, .) %>% as.data.frame() #combine dfs to be longform wrt R0 and format as df


# plot
plot = ggplot() + geom_point(data = df, aes(x = time, y = log10(Pp), group = factor(R0), col = factor(R0)), size = 2) +
  geom_line(data = df, aes(x = time, y = log10(Pp), group = factor(R0), col = factor(R0)), lwd = 1.5) +
  labs(x = "Time post vaccination, years", y = bquote("log"[10] ~ "probability of emergence per year, P"[P]), col = bquote("R"[0] ~ "at spillover")) +
  theme_bw() +
  scale_color_manual(values = my.colors) +
  coord_cartesian(ylim = c(-11, 0)) +
  geom_hline(yintercept = -5, linetype = "dashed") +
  geom_vline(xintercept = min_year, linetype = "dashed")

#print to viewer
print(plot)
```

##########################

I can do a heatmap with F(t) on the x axis, R0 on the y axis, and p(R0 -> 3) filled in. 

```{r p-r-to-3}
# define vector of R0 and Frac Immune
R0_wt_vec = seq(0.75, 1.05, by = 0.001)
Frac_immune_vec = seq(0, 0.5, by = 0.01)

# define unchanging parms
sigma = 0.05 # proportional increase in fitness with mutation
mu = 10^-3 # rate of beneficial mutations per infection
cp = 0.8 #cross-reactivity
R0_evolved = 3 # max R0 evolvable

# calculate
plot.df = sapply(R0_wt_vec, function(R0_wt){
  sapply(Frac_immune_vec, function(Frac_immune){
    # calculate R_spill and R_final
    Re_spill = R0_wt*(1-cp*Frac_immune) 
    Re_evolved = R0_evolved*(1-cp*Frac_immune)
    
    # Caluclate non-extinction probability 
    return(pEmergence(sigma = sigma, mu = mu, Re_spill = Re_spill, Re_evolved = Re_evolved)) #run simulation for each R0, Immunity combination
  })
})

#plot
filled.contour(x = Frac_immune_vec, y = R0_wt_vec, z=log10(plot.df), nlevels = 20, lwd = 2,
               xlab = "Fraction Immune", ylab = expression("R"[0] ~ "of Spillover"), 
               key.par = list(mar = c(4, 1, 4, 3)),
               key.title = title(main = expression("P"[E])),
               plot.axes = {
                 axis(1)
                 axis(2)
                 #arrow pointing left 
                 arrows(x0 = 0.45, x1 = 0.2, y0 = 0.8, y1 = 0.8, length = 0.1, lwd = 2) #not the same place
                 text(x = 0.4, y = 0.81, labels = "Decay over time") 
               })

# greate long form grid
plot_long <- expand.grid(FracImmune = Frac_immune_vec,
                         R0 = R0_wt_vec) %>%
  mutate(z = as.vector(log10(plot.df)))

# Build arrow segments from proportion_immune
times_to_plot = c(25,50,75,100)
arrow_df <- proportion_immune %>%
  filter(time %in% times_to_plot) %>%
  mutate(x0 = lag(prop_immune, 1),
         x1 = prop_immune,
         y = 0.8) %>%
  filter(!is.na(x0))

ggplot(plot_long, aes(x = FracImmune, y = R0, fill = z)) +
  geom_tile() +
  scale_fill_viridis_c() +
  geom_segment(data = arrow_df,
               aes(x = x0, xend = x1, y = y, yend = y),
               arrow = arrow(length = unit(0.1, "inches")),
               linewidth = 1, inherit.aes = FALSE) +
  annotate("text", x = proportion_immune %>%
  filter(time %in% times_to_plot) %>% pull(prop_immune), y = c(0.805, 0.785, 0.805, 0.785), label = times_to_plot,
           hjust = 0, vjust = 0, size = 5, fontface = "bold") +
  labs(x = "Fraction Immune",
       y = expression("R"[0] ~ "of Spillover"),
       fill = expression("P"[E])) +
  theme_bw()
```
