---
title: "Waning population immunity projections for Smallpox in the Congo, 2023-2050"
subtitle: "Modeling the decay of immunity to smallpox at the population level after ceasing vaccination"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure

#load functions
source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function
source(here("code", "functions", "matrix_model.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to predict the birth and death rates in the DRC from 2024-2050 and predict the change in portion of individuals born during or before 1980.

Goals: 

1) Predict fecunditities from 2024-2050
2) predict moratlity from 2024-2050
3) use predictions in matrix model to predict waning immunity from 2024-2050

# Trends in birth rates and death rates from 1980-2023

First, we load the fecundities and mortalities from `data\congo\formatted data\`. 

```{r load-data}
#birth rates
births = as.matrix(read_csv(here("data", "congo", "formatted data", "birth_rates.csv"))[,-1])

#death rates
deaths = as.matrix(read_csv(here("data", "congo","formatted data", "death_rates.csv"))[,-1])
```

For easier manipulation, I want to reformat the matrices as long form data frames: 

```{r format-demographics}
#format birth rates
births_lf = as.data.frame(births) %>% #add column for year, then pivot to long form with column indicating age and column indicating birth rates
  mutate("Year" = seq(1980:2023)) %>%
  pivot_longer(
    cols = !Year,
    names_to = "Age",
    values_to = "birth_rate"
  )

#format death rates
deaths_lf = as.data.frame(deaths) %>% #add column for year, then pivot to long form with column indicating age and column indicating death rates
  mutate("Year" = seq(1980:2023)) %>%
  pivot_longer(
    cols = !Year,
    names_to = "Age",
    values_to = "death_rate"
  )

```

Now, lets plot by age:

```{r plot-demographics}
#births
fecundity_trend = ggplot() +
  geom_line(data = births_lf, aes(x = Year, y = (birth_rate), group = as.factor(as.numeric(Age)), col = as.factor(as.numeric(Age)))) +
  labs(x = "Year", y = "log10 Birth Rate", title = "Fecundity by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

#deaths
mortality_trend = ggplot() +
  geom_line(data = deaths_lf, aes(x = Year, y = log10(death_rate), group = as.factor(as.numeric(Age)), col = as.factor(as.numeric(Age)))) +
  labs(x = "Year", y = "log10 Death Rate", title = "Death by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

#plot together
grid.arrange(fecundity_trend, mortality_trend)


```

A naive approach (but the approach I'm taking for now) is to fit a linear and an exponential model to these guys and use the parameters of these models to predict csx