---
title: "Effect of partial vaccination on disease emergence"
subtitle: "To what extenct does partial vaccination hinder the adaptive evolution of pathogens with subcritical R_0? "
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors

source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to model the decay of immunity to smallpox and related pox viruses at the population level after vaccination ceased (1980) in the Democratic Republic of the Congo. 

# Approach 

We use a Leslie-matrix age-structured model to capture the birth of susceptible individuals and loss of immune individuals in the DRC from 1980-2023. 

We assume that all individuals in all age classes were vaccinated against or immune to smallpox by the time of smallpox eradication (and the subsequent cessation of vaccination campaigns.)  

Fecundity and survivorship estimates by single-year age are taken from the WHO [Department of Economic and Social Affairs, Population Division](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=Population).

# Loading data

The datasets for this analysis can be found in the excel file `data/congo/congo_demographics_1980-2023.xlsx`. This file contains three sheets: 

1) the number of individuals of each age in 1980 (census),
2) mortality by age from 1980-2023, and
3) fecundity by age from 1980-2023.

Note that fecundity is given as the average number of children per 1000 women. We will necessarily need to transform these estimates to return the number of fertile offspring (female) per 1 woman. We will return to this. 

The following code chunk imports these three sheets as 3 independent data frames in a single list: 

```{r load-data}
congo_demographics = multiplesheets(here("data", "congo", "congo_demographics_1980-2023.xlsx"))

```

Here, we take two processing steps. 

First, mortality is reported in the number of deaths (in thousands) per age. For death rate in each age group, I need to divide the number of deaths by the total number in that age group in the population. As both the number of individuals at each and the number of deaths at each age are given per 1000, I can simply take the ratio of dead:population. 

```{r death-rates}
# extract mortality dataframe from congo_demographics
deaths= congo_demographics$mortality %>% #longform for easier manipulation
  pivot_longer(
    cols = !Year,
    names_to = "Age", 
    values_to = "Deaths"
  ) %>%
  mutate("Age" = case_when(Age == "100+" ~ "100", .default = Age)) #rename 100+ age to 100 

# extract total number of individuals at each age from population dataframe in congo_demographics
population = congo_demographics$population %>%
  pivot_longer(
    cols = !Year, 
    names_to = "Age",
    values_to = "Size"
  ) 

# combine data frames by Age and Year 
deaths = inner_join(deaths,population, by = c("Year", "Age"))

# use nrows to check all rows are kept
deaths = deaths %>%
  mutate("mortality" = Deaths/Size)

ggplot(data = deaths) +
  geom_line(aes(x = Year, y = log10(mortality), color = as.factor(as.numeric(Age)), group = as.factor(as.numeric(Age)))) + 
  labs(x = "Year", y = "log10 Mortality Rate", title = "Mortality by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()
```

We see some decline in mortality over time, which is a good sign that we've got these right. For late projections, we might use this information to make predictions about mortality rates in the future. 

Fecundity is given as the average number of children per 1000 women. We will necessarily need to transform these estimates to return the number of fertile offspring (female) per 1 woman; we can accomplish this by dividing by $2 \times 1000$.


```{r birth-rates}
# extract fecundity dataframe from congo_demographics
births = congo_demographics$fertility
# find the minimum and maximum ages 
min_age = min(as.numeric(colnames(births)), na.rm = T)
max_age = max(as.numeric(colnames(births)), na.rm = T)
add_ages = as.character(c(0:(min_age-1), (max_age+1):100))

# add columns of zero fecundity for non-fertile ages
births[add_ages] = 0


# format df to lonform
births = births %>% pivot_longer(
    cols = !Year,
    names_to = "Age", 
    values_to = "Births"
  )

# mutate births to obtain fertile offspring per woman 
births = births %>%
  mutate("Births" = Births/(2*1000))

#plot
ggplot(data = births) +
  geom_line(aes(x = Year, y = (Births), color = as.factor(as.numeric(Age)), group = as.factor(as.numeric(Age)))) + 
  labs(x = "Year", y = "log10 Birth Rate", title = "Fecundity by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

```

Note that birth rate will be considered 0 for ages 0-14 and 50-100, as is assumed in the report. 
Also note that both the population size, death rate, and fecundity dataframes have 4444 rows, corresponding to each Age-Year combination. 

For our leslie matrix, we will sadly want to collapse these back to long form. 

Two ways: 
- first, model decay forward using 2023 birth/death rates
- second, model decay forward using projected birth/death rates from trends in both 

Finally, we will need to account for the effect of increasing population density on the success of transmission. Let's consider this moving forward. 





Sojourn:

```{r formatting-population-dataframe}
# extracting population size through time from congo_demographics list and formatting in long-form
population = congo_demographics$population %>% 
  pivot_longer( #longform for easier manipulation 
    cols = !Year,
    names_to = "Age", 
    values_to = "Size"
    ) %>% 
  as.data.frame() # make data frame, not tibble

summary(population) #see variable types and information

# mutate age feature to numeric
population = population %>% mutate("Age" = as.numeric(Age))

```

```{r reported-vaccinated-through-time}
# adding new feature to population dataframe - vacced
# binary, true if older than age-1980 and thus had received vaccine or natural infection; false if younger and not immune.
# indicated whether immune
population = population %>%
  mutate("Vacced" = case_when(Age >= Year - 1980 ~ TRUE, .default = FALSE))

# sum number category for all vacced within each year
vacc_frac = population %>%
  group_by(Year) %>% #perform functions within groups
  filter(Vacced == T) %>% #filter for only vaccinated ages
  summarise(num_immune = sum(Size)) #sum across all ages 

#add total population size each year to data frame
vacc_frac$total = population %>%
  group_by(Year) %>%
  summarise(total = sum(Size)) %>%
  pull("total")

#calculate fraction immune
vacc_frac = vacc_frac %>%
  mutate("prop_immune" = num_immune/total) #yahoo
```

```{r plot-declining-immunity}
ggplot(data = vacc_frac) + 
  geom_point(aes(x = Year, y = log10(prop_immune))) + geom_line(aes(x = Year, y = log10(prop_immune))) + 
  labs(x = "Year", y = "log10 Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023") +
  theme_bw()  

```

Looks more or less like an exponential drop in immunity, though with slightly great decay (greater concavity). 

Lets see if we can recreate this behavior with the leslie matrix model. 

