---
title: "Exploring waning population immunity to Smallpox in the Congo"
subtitle: "Doc 1. How to best model the decay of immunity to smallpox at the population level after ceasing vaccination"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure

source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The goal of this document is to explore models of the decay of immunity to smallpox and related pox viruses at the population level after vaccination ceased (1980) in the Democratic Republic of the Congo. 

Two methods will be 1) predicting fecudities and mortality from trends in demographics and using an age-structured matrix model to estimate the decay in the proportion of individuals immune, and 2) fitting a decay model to population size through time. 

# Approach 

We use a Leslie-matrix age-structured model to capture the birth of susceptible individuals and loss of immune individuals in the DRC from 1980-2023. 

We assume that all individuals in all age classes were vaccinated against or immune to smallpox by the time of smallpox eradication (and the subsequent cessation of vaccination campaigns.)  

Fecundity and survivorship estimates by single-year age are taken from the WHO [Department of Economic and Social Affairs, Population Division](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=Population).

# Loading data

The datasets for this analysis can be found in the excel file `data/congo/congo_demographics_1980-2023.xlsx`. This file contains three sheets: 

1) the number of individuals of each age in 1980 (census),
2) mortality by age from 1980-2023, and
3) fecundity by age from 1980-2023.

Note that fecundity is given as the average number of children per 1000 women. We will necessarily need to transform these estimates to return the number of fertile offspring (female) per 1 woman. We will return to this. 

The following code chunk imports these three sheets as 3 independent data frames in a single list: 

```{r load-data}
congo_demographics = multiplesheets(here("data", "congo", "congo_demographics_1980-2023.xlsx"))

```

Here, we take two processing steps. 

First, mortality is reported in the number of deaths (in thousands) per age. For death rate in each age group, I need to divide the number of deaths by the total number in that age group in the population. As both the number of individuals at each and the number of deaths at each age are given per 1000, I can simply take the ratio of dead:population. 

```{r death-rates}
# extract mortality dataframe from congo_demographics
deaths = congo_demographics$mortality %>% #longform for easier manipulation
  pivot_longer(
    cols = !Year,
    names_to = "Age", 
    values_to = "Deaths"
  ) %>%
  mutate("Age" = case_when(Age == "100+" ~ "100", .default = Age)) #rename 100+ age to 100 

# extract total number of individuals at each age from population dataframe in congo_demographics
population = congo_demographics$population %>%
  pivot_longer(
    cols = !Year, 
    names_to = "Age",
    values_to = "Size"
  ) 

# combine data frames by Age and Year 
deaths = inner_join(deaths,population, by = c("Year", "Age"))

# use nrows to check all rows are kept
deaths = deaths %>%
  mutate("mortality" = Deaths/Size)

ggplot(data = deaths) +
  geom_line(aes(x = Year, y = log10(mortality), color = as.factor(as.numeric(Age)), group = as.factor(as.numeric(Age)))) + 
  labs(x = "Year", y = "log10 Mortality Rate", title = "Mortality by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()
```

We see some decline in mortality over time, which is a good sign that we've got these right. For late projections, we might use this information to make predictions about mortality rates in the future. 

Fecundity is given as the average number of children per 1000 women. We will necessarily need to transform these estimates to return the number of fertile offspring (female) per 1 woman; we can accomplish this by dividing by $2 \times 1000$.


```{r birth-rates}
# extract fecundity dataframe from congo_demographics
births = congo_demographics$fertility
# find the minimum and maximum ages 
min_age = min(as.numeric(colnames(births)), na.rm = T)
max_age = max(as.numeric(colnames(births)), na.rm = T)
add_ages = as.character(c(0:(min_age-1), (max_age+1):100))

# add columns of zero fecundity for non-fertile ages
births[add_ages] = 0


# format df to lonform
births = births %>% pivot_longer(
    cols = !Year,
    names_to = "Age", 
    values_to = "Births"
  )

# mutate births to obtain fertile offspring per woman 
births = births %>%
  mutate("Births" = Births/(2*1000))

#plot
ggplot(data = births) +
  geom_line(aes(x = Year, y = (Births), color = as.factor(as.numeric(Age)), group = as.factor(as.numeric(Age)))) + 
  labs(x = "Year", y = "log10 Birth Rate", title = "Fecundity by Age, 1980 - 2023", colour = "Age" ) + 
  scale_color_manual(values = my.colors) + 
  theme_bw()

```

Note that birth rate will be considered 0 for ages 0-14 and 50-100, as is assumed in the report. 
Also note that both the population size, death rate, and fecundity dataframes have 4444 rows, corresponding to each Age-Year combination. 

For our leslie matrix, we will sadly want to collapse these back to long form. 

```{r formating-parameter-matrices}
# matrix of birth rates. Rows are year, columns are age
births = births %>% pivot_wider(
  names_from = Age,
  values_from = Births
)
#order columns by increasing age and save as matrix
births = births[,as.character(0:100)] %>% as.matrix() #ensures the correct order of columns

# matrix of death rates. Rows are year, columns are age
deaths = deaths %>% 
  select(Year, Age, mortality) %>% pivot_wider(
  names_from = Age, 
  values_from = mortality
)
#order columns by increasing age and save as matrix
deaths = deaths[, as.character(0:100)] %>% as.matrix() #ensures the correct order of columns by age

# isolate initial population structure in 1980 as vector
init_pop = population %>% filter(Year == 1980) %>% pull(Size) #be sure that pop is in order of ages. Sizes are in thousands. 


#save these to formatted data
write.csv(births, file = here("data", "congo", "formatted data", "birth_rates.csv"))
write.csv(deaths, file = here("data", "congo", "formatted data", "death_rates.csv"))
write.csv(init_pop, file = here("data", "congo", "formatted data","pop_1980.csv"))
```

Setting up leslie matrix: 

```{r matrix-model-setup}
#initialize leslie matrix with parameters from 1980
L = diag(x = 1 - deaths[1, 1:100]) #firts year survivorship 1-death rate on diagonal
L = rbind(births[1, 1:100], L) #add first year fecundities to top row
L = cbind(L, c(births[1,101], rep(0,99), 1-deaths[1,101])) # add column ensures last row includes final death rate 
tail(L) #looks good!

# matrix to store population densities by age each year. rows will be age, columns will be year. 
m = matrix(nrow = length(init_pop), ncol = 2023-1979) #44 columns for 44 years
m[,1] = init_pop #initialize first column with populations structure in 1980

#how many generations/years to run matrix? 
years = 2023-1980

```

```{r simulate}
#simulate population structure since 1980
for(year in 1:years){
  m[,year+1] = L %*% m[,year] #update population distribution
  
  #adjust leslie matrix with parameters from year
  L = diag(x = 1 - deaths[year+1, 1:100]) #firts year survivorship 1-death rate on diagonal
  L = rbind(births[year+1, 1:100], L) #add first year fecundities to top row
  L = cbind(L, c(births[year+1,101], rep(0,99), 1-deaths[year+1,101])) # add column ensures last row includes final survivorship
  
}

```

```{r plot-total}
population_size = apply(m, 2, sum)
plot(x = 0:years, y = population_size)
```

Looks good!

```{r }
#formatting population matrix as age dataframe
sim_pop = cbind(1980:2023, t(m)) %>% as.data.frame() 
colnames(sim_pop) = c("Year", as.character(0:100))


# extracting population size through time and formatting in long-form
sim_population = sim_pop %>% 
  pivot_longer( #longform for easier manipulation 
    cols = !Year,
    names_to = "Age", 
    values_to = "Size"
    ) %>% 
  as.data.frame() # make data frame, not tibble

summary(sim_population) #see variable types and information

# mutate age feature to numeric
sim_population = sim_population %>% mutate("Age" = as.numeric(Age))
```

```{r}
# adding new feature to population dataframe - vacced
# binary, true if older than age-1980 and thus had received vaccine or natural infection; false if younger and not immune.
# indicated whether immune
sim_population = sim_population %>%
  mutate("Vacced" = case_when(Age >= Year - 1980 ~ TRUE, .default = FALSE))

# sum number category for all vacced within each year
sim_vacc_frac = sim_population %>%
  group_by(Year) %>% #perform functions within groups
  filter(Vacced == T) %>% #filter for only vaccinated ages
  summarise(num_immune = sum(Size)) #sum across all ages 

#add total population size each year to data frame
sim_vacc_frac$total = sim_population %>%
  group_by(Year) %>%
  summarise(total = sum(Size)) %>%
  pull("total")

#calculate fraction immune
sim_vacc_frac = sim_vacc_frac %>%
  mutate("prop_immune" = num_immune/total) #yahoo

```


Simulated fraction immune in red, actual reported fraction in black. Cool!
# Age structure in 1980, 2000, and 2023

```{r age-structure-plots}
### create df of population age distribution in 1980, 2000, and 2023
age_structure = data.frame("Age" = 0:100, "pop_1980" = m[, (1980-1980)+1], "pop_2000" = m[,(2000-1980)+1], "pop_2023" = m[,(2023-1980)+1])
### note that I don't have male and female ratios here, so I'll assume 50/50 
p_1980 = ggplot() + # 1980
  geom_col(data = age_structure, aes(x = pop_1980/2, y = factor(Age), fill = "male")) +
  geom_col(data = age_structure, aes(x = -pop_1980/2, y = factor(Age), fill = "female")) +
  scale_fill_manual(values = c("red", "blue")) + 
  labs(x = "Population size (in thousands)", y = "Age", fill = "Sex", title = "Age Structure, 1980") +
  scale_y_discrete(breaks = seq(0,100, 10), labels = seq(0, 100, 10)) + 
  scale_x_continuous(breaks=c(-600,-300, 0 , 300, 600), labels=c("600", "300", "0", "300", "600")) +
  theme_bw()

p_2000 = ggplot() + # 2000
  geom_col(data = age_structure, aes(x = pop_2000/2, y = factor(Age), fill = "male")) +
  geom_col(data = age_structure, aes(x = -pop_2000/2, y = factor(Age), fill = "female")) +
  scale_fill_manual(values = c("red", "blue")) + 
  labs(x = "Population size (in thousands)", y = "Age", fill = "Sex", title = "Age Structure, 2000") +
  scale_y_discrete(breaks = seq(0,100, 10), labels = seq(0, 100, 10)) + 
  scale_x_continuous(breaks=c(-900, -600,-300, 0 , 300, 600, 900), labels=c("900", "600", "300", "0", "300", "600", "900")) +
  theme_bw()

p_2023 = ggplot() + # 2023
  geom_col(data = age_structure, aes(x = pop_2023/2, y = factor(Age), fill = "male")) +
  geom_col(data = age_structure, aes(x = -pop_2023/2, y = factor(Age), fill = "female")) +
  scale_fill_manual(values = c("red", "blue")) + 
  labs(x = "Population size (in thousands)", y = "Age", fill = "Sex", title = "Age Structure, 2000") +
  scale_y_discrete(breaks = seq(0,100, 10), labels = seq(0, 100, 10)) + 
  scale_x_continuous(breaks=c(-2000, -1000, -500, -100, 0 , 100, 500, 1000, 2000), labels=c("2000", "1000", "500", "100", "0", "100", "500", "1000", "2000")) +
  theme_bw()

# plot together
grid.arrange(p_1980, p_2000, p_2023, nrow = 3)

```

From the age structure, it looks like the Congo has maintained rapid growth from 1980 - present, suggesting that the population has maintained high fecundity. 

Probably not good considering that the density of people is increasing and making it more likely that disease can be maintained in the population. 


Two ways: 
- first, model decay forward using 2023 birth/death rates
- second, model decay forward using projected birth/death rates from trends in both 

Finally, we will need to account for the effect of increasing population density on the success of transmission. Let's consider this moving forward. 


To do: 

1) Projectsions: 
  a) lets give the matrix model and the true data model their own scripts
  b) lets extrapolate from an exponential waning model with the true data
  c) lets extrapolate the birth and death rates from the true data
  d) if we want we can plot population structure in 1980 and 2023 and make an inference on future structure from that? DONE


Sojourn:

```{r formatting-population-dataframe}
# extracting population size through time from congo_demographics list and formatting in long-form
population = congo_demographics$population %>% 
  pivot_longer( #longform for easier manipulation 
    cols = !Year,
    names_to = "Age", 
    values_to = "Size"
    ) %>% 
  as.data.frame() # make data frame, not tibble

summary(population) #see variable types and information

# mutate age feature to numeric
population = population %>% mutate("Age" = as.numeric(Age))

```

```{r reported-vaccinated-through-time}
# adding new feature to population dataframe - vacced
# binary, true if older than age-1980 and thus had received vaccine or natural infection; false if younger and not immune.
# indicated whether immune
population = population %>%
  mutate("Vacced" = case_when(Age >= Year - 1980 ~ TRUE, .default = FALSE))

# sum number category for all vacced within each year
vacc_frac = population %>%
  group_by(Year) %>% #perform functions within groups
  filter(Vacced == T) %>% #filter for only vaccinated ages
  summarise(num_immune = sum(Size)) #sum across all ages 

#add total population size each year to data frame
vacc_frac$total = population %>%
  group_by(Year) %>%
  summarise(total = sum(Size)) %>%
  pull("total")

#calculate fraction immune
vacc_frac = vacc_frac %>%
  mutate("prop_immune" = num_immune/total) #yahoo
```

```{r plot-declining-immunity}
ggplot(data = vacc_frac) + 
  geom_point(aes(x = Year, y = log10(prop_immune))) + geom_line(aes(x = Year, y = log10(prop_immune))) + 
  labs(x = "Year", y = "log10 Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023") +
  theme_bw()  

```

Looks more or less like an exponential drop in immunity, though with slightly great decay (greater concavity). 

Lets see if we can recreate this behavior with the leslie matrix model. 

```{r plot-declining-immunity-sim}
ggplot() + 
  geom_point(data = sim_vacc_frac, aes(x = Year, y = log10(prop_immune)), col = "red") + geom_line(data = sim_vacc_frac, aes(x = Year, y = log10(prop_immune)), col = "red") + 
  geom_point(data = vacc_frac, aes(x = Year, y = log10(prop_immune))) + geom_line(data = vacc_frac, aes(x = Year, y = log10(prop_immune))) +
  labs(x = "Year", y = "log10 Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023") +
  theme_bw()  

ggplot() + 
  geom_point(data = sim_vacc_frac, aes(x = Year, y = (prop_immune)), col = "red") + geom_line(data = sim_vacc_frac, aes(x = Year, y = (prop_immune)), col = "red") + 
  geom_point(data = vacc_frac, aes(x = Year, y = (prop_immune))) + geom_line(data = vacc_frac, aes(x = Year, y = (prop_immune))) +
  labs(x = "Year", y = "Prop Immune", title = "Decline in portion of population immune to Smallpox, 1980-2023") +
  theme_bw()  

```

