---
title: "Exploring the rate of increase of effective reproductive number of Mpox in the Congo due to loss of Smallpox immunity"
subtitle: "Plotting the change in Re through time"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(gridExtra) #for arrangeing plotsx
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure

source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 10
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

We have modeled the waning of immunity at the population level in the Congo from 1980-2023 and projected the waning of immunity at the population level forward to 2080, 100 years after the eradication of and ceasing vaccination against smallpox. 

The goal of this document is to show how the effective reproductive number of a related zoonotic virus - mpox - is expected to decrease solely as a function of the loss of population immunity. We will consider the case where cross-immunity from smallpox infection or vaccination is 100%, as well as when smallpox immunity is imperfectly protective against mpox. 

# Load waning estimates

The projections for the waning of population immunity can be found in `results\data analysis\congo\proportion_immune.csv`. I will load this below:

```{r load-pop-immunity}
proportion_immune = read.csv(here("results", "data analysis", "congo", "proportion_immune.csv"))

```

# Decay in Smallpox immunity at the population

We estimated the loss in the proportion of individuals born during or before the year 1980 through time using a matrix model approach parameterized by annual demographics reports from the Congo (see `code\data analysis\congo\matrix_model_projections_2.qmd`). The model projects the loss of individuals in the future assuming the exponential growth observed from 1980-2023 continues to 2080. 

We can plot the proportion immune over time as a reminder of what these projections look like: 

```{r plot-immune}
plot = ggplot() +
  geom_line(data = proportion_immune %>% filter(Year <= 2023), aes(x = Year, y = prop_immune)) +
  geom_point(data = proportion_immune %>% filter(Year <= 2023), aes(x = Year, y = prop_immune)) +
  geom_line(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = prop_immune), col = "red") +
  geom_point(data = proportion_immune %>% filter(Year > 2023), aes(x = Year, y = prop_immune), col = "red") +
  labs(x = "Year", y = "proportion immune, f", title = "decay in Smallpox immunity at the population level") +
  theme_bw()

#save 
#file = here("results", "data analysis", "congo", "fig_Smallpox_immunity_decay.png")
#ggsave(filename = file, plot = plot)
```

# Increase in effective reproductive number of Mpox

Below, I will plot the change in the effective reproductive number over time as a function of the loss of smallpox immunity at the population level and the degree of cross-protection conferred by smallpox immunity. This is based on the notion that the effective reproductive number is limited by the density of susceptibles, such that $R_e = R_0(1-f_I(t))$, where $R_0$ is the reproductive number of mpox in a completely vaccinated population and $f_I$ is the portion of individuals immune to mpox as a function of time post smallpox eradication. Note that this assumes perfect cross-protection. 

Because we don't know the effective reproductive number of mpox nor the degree of cross-protection conferred by immunity at time $t$, we can plot the relative change in $R_e$ over time as $\frac{R_e}{R_0} = 1-cf_I$, where $c$ represents the degree of cross-protection conferred by smallpox immunity. Note that $\frac{R_e}{R_0}$ gives a measure of the proximity of the current $R_e$ to mpox's $R_0$. 

```{r calculate-re-r0}
#cross-protection
cs = seq(0.5, 1, by = 0.1) #half protected to completely protected. 

#define function for cross immunity
Re_R0_ratio = function(c, f) {1-c*f}


##create a matrix with year and predicted Re_r0 for each c. 
relative.effective.R = sapply(cs, FUN = function(j) { #columns will be cs, rows will be year
  sapply(proportion_immune$prop_immune, FUN = function(i) {
    #set cross-reactive fraction immune equal to c*f
    Re_R0_ratio(j, i)})}) #applying re_r0_ratio function with each c to proportion immune


#add year to resulting matrix, format as dataframe, and rename features
relative.effective.R = cbind(proportion_immune$Year, relative.effective.R) %>% as.data.frame()
colnames(relative.effective.R) = c("Year", cs)

#make longform for ease of plotting 
relative.effective.R = relative.effective.R %>% 
  pivot_longer(cols = !Year, 
               names_to = "cross.immunity", 
               values_to = "relative.Re")

```

```{r plot-re-r0}
plot = ggplot() + 
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
  scale_color_manual(values = my.colors) +
  labs(x = "Year", y = "relative effective R, Re/R0", title = "increase in Mpox reproductive number") +
  theme_bw()

#save 
#file = here("results", "data analysis", "congo", "fig_increase_Mpox_relative_R.png")
#ggsave(filename = file, plot = plot)
```

and finally, we can plot our estimate $R_e$ given different potential basic reproductive numbers of mpox: 

```{r plot-re}
#define R0s
R0s = c(0.95, 1.05,1.1,2)

plot = do.call(grid.arrange, lapply(R0s, FUN = function(R0){
  ggplot() + 
  geom_line(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = R0*relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year <= 2023), aes(x = Year, y = R0*relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_line(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = R0*relative.Re, group = cross.immunity, col = cross.immunity)) +
  geom_point(data = relative.effective.R %>% filter(Year > 2023), aes(x = Year, y = R0*relative.Re, group = cross.immunity, col = cross.immunity), shape = 21) +
    geom_vline(xintercept = 2023, linetype = "dashed") +
  scale_color_manual(values = my.colors) +
  labs(x = "Year", y = "effective R, Re", title = paste("increase in Mpox reproductive number when R0 =", R0)) +
  theme_bw()
}))

#save 
#file = here("results", "data analysis", "congo", "fig_increase_Mpox_effective_R.png")
#ggsave(filename = file, plot = plot)
```