---
title: "Exploring probability of evolution to high transmissibility after emergence"
subtitle: "Doc.1 How to best model the probability of evolution to high transmission when R > 1"
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(RColorBrewer) #for colors
library(gridExtra) #for plotting multiple plots in one figure
library(parallel)

source(here("code", "functions", "analysis_functions.R")) #for current version of adaptive mutation function

# colors for plotting
num_cols = 101
my.colors = colorRampPalette(brewer.pal(n = 8, name = "Set2"))(num_cols)
```

# Objective

The objective of this document is to develop a function to simulate the multi-step branching process of the adaptation of a virus with a near critical $R_0$ out of a zoonotic source. 

# Model 


```{r bp-testfunction}
### define function
bp_testdeltaR = function(Rwt, deltaR, Rfinal, mu, gens){
  ##### calculate parameters
  # caluclate number of types
   # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    print("extinct")
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    print("emergent") 
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if emerges at high R
  }
  return(X)
}


```

```{r run-sim}
# parms
Rwt = 1.5
deltaR = 0.3 
Rfinal = 3
mu = 10^-3
gens = 1000


##### calculate parameters
  # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    print("extinct")
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    print("emergent") 
    print(paste("At gen", gen, "with last type", max(rows_greaterthan_0), "of", m, "and total", sum(X, na.rm = T)))
    break} #break if emerges at high R
  }

### cool 
```


```{r run-testfunction}
# parms
Rwt = 1.3
deltaR = 0.3
Rfinal = 3
mu = 10^-3
gens = 100

#run sim
N = bp_deltaR(Rwt = Rwt, deltaR = deltaR, Rfinal = Rfinal, mu = mu, gens = 100)


```

# Probably and time to emergence *per intro*

I will run the simulation 100 and calculate the fraction of emergent versus extinct spillover events. 
I will keep track of the total size of the outrbeak (sum of all infections) and the number of generations before the emergence of the highly-transmissible variant. 

```{r }
### define function
bp_deltaR = function(Rwt, deltaR, Rfinal, mu, gens){
  ##### calculate parameters
  # caluclate number of types
   # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    extinct = extinct + 1
    extinction_size = c(extinction_size, sum(X))
    extinction_gen = c(extinction_gen, gen)
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    emerged = emerged + 1
    emergence_size = c(emergence_size, sum(X))
    emergence_gen = c(emergence_gen, gen)
    break} #break if emerges at high R
  }
  return(X)
}
```

```{r sim-setup}
# Parms
spills = 10 #number of spillovers 

Rwt = 1.3 #wt r
deltaR = 0.3 #change in r with mut  
Rfinal = 3 #r of final type 
mu = 10^-3 #mutation rate
gens = 100 #number of generations to run each sim

# Storage
## emergence
emerged = 0 #counter to keep track of emergence events
emergence_size = c(0) # vector to keep track of final sizes of emergence events
emergence_gens = c(0) # vector to keep track of generation at which final variant emerges
## extinction
extinct = 0 #counter to keep track of extinction events
extinction_size = c(0) # vector to keep track of final sizes of extinction events
extinction_gens = c(0) # vector to keep track of generation at which extinction occurs


```

```{r run-sim-perintro}
c = 1
while(c<=spills){
  #parms
   # caluclate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  #### report
    # last type 
    #rows_greaterthan_0 <- which(apply(X > 0, 1, any))
    
    
  if(all(X[,gen] == 0, na.rm = T)){
    extinct = extinct + 1
    extinction_size = c(extinction_size, sum(X, na.rm = T))
    extinction_gens = c(extinction_gens, gen)
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    emerged = emerged + 1
    emergence_size = c(emergence_size, sum(X, na.rm = T))
    emergence_gens = c(emergence_gens, gen)
    break} #break if emerges at high R
  }
  if(c %% 10 == 0){
  print(paste("done", c))
  }
  c = c+1
  
}

```

```{r analysis}
#### fraction emerged and extinct
p_emerge = emerged/spills
p_extinct = extinct/spills

#average number of gens to and size at emergence 
mean_emergence_gens = mean(emergence_gens)
mean_emergence_size = mean(emergence_size)

#average number of gens to and size after extinction
mean_extinction_gens = mean(extinction_gens)
mean_extinction_size = mean(extinction_size)
```


Next steps - 
run simulation for varying wild type and delta R. 

Then add 1 or 2 other mus. 

Figuring out how to do this timely is not trivial! I bet I could get away with 100 spillovers...

# Parallelization

## 
Each simulation takes quite a bit of time. I would like to run them in parallel. First, I'd like to change my function to use `apply` instead of a `while` loop. 

I am going to create a function that returns 3 numbers: 

1) `type`, binary 1 or 0, depending on whether the disease has emerged (1) or gone extinct (0),
2) `gens`, the number of generations before emergence or extinction, and 
3) `size` the total number of infections before emergence or extinction.

```{r iterable-function}
#function for a single iteration of bp 
iterable_bp_deltaR = function(Rwt, deltaR, Rfinal, mu, gens){
  #parms
   # calculate number of types
  m = ceiling((Rfinal-Rwt)/deltaR) + 1 #divide interval into m-1 segments of delta R length and add 1 for initial type
  # define vector of R0 for each type 
  R0 = Rwt + (0:(m))*deltaR # add the changes from wtR0 to give R0 of each variant
  # define empty matrix for storing numbers of individuals of each type over gens generations
  X = matrix(nrow = m, ncol = gens+1) #rows are types over time, columns are individuals of each type at given time
 
  ##### initial conditions
  init = c(1, rep(0, m-1)) # start with one individual of wt and no individuals of any other type
  X[,1] = init # add to matrix for gen 0
 
  #### simulation
  # stochastic simulation summing random variables each gen
  for(gen in 2:gens){
    X[1,gen] = sum(rpois(n = X[1,gen-1], lambda = (1-mu)*R0[1])) # add X1,gen-1 random samples from pois dist with lambda = (1-u)R0(i)
    for(i in 2:(m-1)){
      X[i,gen] = sum(rpois(n = X[i,gen-1], lambda = (1-mu)*R0[i])) + # contributions from type i
        sum(rpois(n = X[i-1,gen-1], mu*R0[i-1])) #mutants from type i-1
    }
    X[m,gen] = sum(rpois(n = X[m,gen-1], lambda = R0[m])) + #contribution from type m
      sum(rpois(n=X[m-1,gen-1],lambda = mu*R0[m-1])) #contribution from type m-1
  
  if(gen == gens){
    type = 2
    size = sum(X, na.rm = T)
    generation = gen
  }  
    
  #### report
  if(all(X[,gen] == 0, na.rm = T)){
    type = 0
    size = sum(X, na.rm = T)
    generation = gen
    break} #break if extinct
    
  else if(any(X[m,] > 0, na.rm = T)){
    type = 1
    size = sum(X, na.rm = T)
    generation = gen
    break} #break if emerges at high R
    
    else if(sum(X, na.rm = T) > 10^9){
    type = 1
    size = NaN
    generation = NaN #break if too big
    break}
    
  }
  
  #return a vector with type, gens, size. these will be concatenated into a single dataframe row-wise after all iterations
  iteration_result = c(type, size, generation)
  return(iteration_result)
}



```

```{r sim-setup}
# Parms
spills = 10 #number of spillovers 

Rwt = 1.3 #wt r
deltaR = 0.1 #change in r with mut  
Rfinal = 3 #r of final type 
mu = 10^-3 #mutation rate
gens = 100 #number of generations to run each sim

```

Lets test out 10 iterations for a single `Rwt` and `deltaR` combination.

```{r test-single-regime}
#run simulation spills times using replicate and save output in a dataframe
system.time(sim1 <- mclapply(1:spills, function(i) {iterable_bp_deltaR(Rwt=Rwt, 
                                          deltaR=deltaR, 
                                          Rfinal=Rfinal, 
                                          mu=mu,
                                          gens=gens)}))
sim1 = do.call(rbind,sim1) %>% as.data.frame()
 
#rename columns
colnames(sim1) = c("type", "size", "generation")
```
                   

For each parameter combination, we would like to reduce this data frame into a vector of six numbers - probability of emergence or extinction, average outbreak size if emerged or if extinct, and number of generations until emergence or extinction. We would also like to know the values of the changing parameters, $Rwt$ and $deltaR$, so we will add those

```{r metrics}
#output vector 
p_emerge = sum(sim1$type)/spills
p_extinct = 1 - p_emerge

#outbreak sizes
mean_size_emerge = sim1 %>%
  filter(type == 1) %>%
  pull(size) %>%
  mean(na.rm = T)

mean_size_extinct = sim1 %>%
  filter(type == 0) %>%
  pull(size) %>%
  mean(na.rm = T)

#number of gens to final
mean_gens_emerge = sim1%>% 
  filter(type == 1) %>%
   pull(generation) %>%
  mean(na.rm = T)

mean_gens_extinct = sim1 %>%
  filter(type == 0) %>%
  pull(generation) %>%
  mean(na.rm = T)

# dataframe
slightly_supercritical = data.frame("Rwt" = Rwt, 
                                    "deltaR" = deltaR,
                                    "Pemerge" = p_emerge,
                                    "Pextinct" = p_extinct,
                                    "mean_size_emerge" = mean_size_emerge,
                                    "mean_size_extinct" = mean_size_extinct,
                                    "mean_gens_emerge" = mean_gens_emerge,
                                    "mean_gens_extinct" = mean_gens_extinct)


```

Putting these chunks together into a single function that takes in a matrix of Rwt and deltaR combinations and outputs the 6 metrics: 

```{r}
#define simulation for parallelization with mclapply
bp_deltaR = function(Rwt, deltaR, Rfinal, mu, gens, spills){
  sim1 = mclapply(1:spills,  function(i) { #mclapply returns a list of vectors 
                 iterable_bp_deltaR(Rwt=Rwt, 
                              deltaR=deltaR, 
                              Rfinal=Rfinal, 
                              mu=mu,
                              gens=gens)}) 
sim1 = do.call(rbind,sim1) %>% as.data.frame() #here we collapse that list into a dataframe where each observation is one iteration
#rename columns
colnames(sim1) = c("type", "size", "generation") #rename Df columns for easy access

# dataframe
slightly_supercritical = c("Rwt" = Rwt, # report averages across all iterations such that each regime gets a single observation
                                    "deltaR" = deltaR,
                                    "Pemerge" = sum(sim1$type)/spills,
                                    "Pextinct" = 1 - sum(sim1$type)/spills,
                                    "mean_size_emerge" = sim1 %>%
                                                          filter(type == 1) %>%
                                                          pull(size) %>%
                                                          mean(na.rm = T),
                                    "mean_size_extinct" = sim1 %>%
                                                          filter(type == 0) %>%
                                                          pull(size) %>%
                                                          mean(na.rm = T),
                                    "mean_gens_emerge" = sim1 %>%
                                                          filter(type == 1) %>%
                                                          pull(generation) %>%
                                                          mean(na.rm = T),
                                    "mean_gens_extinct" = sim1 %>%
                                                          filter(type == 0) %>%
                                                          pull(generation) %>%
                                                          mean(na.rm = T))
return(slightly_supercritical)

}

```


# Running across parameter combinations

We would like to see how emergence probability, time to emergence, and total size changes with the wildtype R of the spilled over variant when each infection gives rise to at least one but no more than two secondary infections on average. That is, $1 \le R < 2.$

```{r params}
# Parms
spills = 1000 #number of spillovers 

#Rwt = seq(1.3,1.5,by = 0.2) #wt r
#deltaR = seq(0.1, 0.3, by = 0.2) #change in r with mut  
Rwt = seq(1.2, 2, by = 0.05)
deltaR = seq(0.25, 1, by = 0.25)
Rfinal = 3 #r of final type 
mu = 10^-3 #mutation rate
gens = 100 #number of generations to run each sim

#parameter regime matrix
regimes = expand.grid(Rwt, deltaR)

regimes
```

```{r simulate}
#run 
system.time(run <- apply(regimes, 1, FUN = function(x){bp_deltaR(Rwt=as.numeric(x[1]), deltaR=as.numeric(x[2]), Rfinal=Rfinal, mu=mu, gens=gens, spills=spills)}) %>% t())

run = run %>% as.data.frame()


# Save output 
file = here("results", "analysis", "R greater than 1", "archive", "Rsims", "R_sim_output.csv")
write.csv(run, file, row.names = F)

```

```{r plot-results}
plot = ggplot() +
  geom_line(data = run, aes(x = Rwt, y = Pemerge, group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = palette("set2")) +
  labs(x = "initial R", y = "Probability of evolving high transmissibility", col = "Increase with Mutation", title = "Probability of evolving high transmissibility, spillover R > 1")

file = here("results", "analysis", "R greater than 1", "archive", "Rsims", "fig_P_high_transmission.png")
ggsave(filename = file, plot)

```

- looks like evolution to high transmissibility does not depend on number of mutations, just on how much greater than 1 R is.

- really I should simulate Rs realy close to 1 because thats where the biggest drops are

Lets look at size at emergence. 

```{r plot-size}
plot = ggplot() +
  geom_line(data = run, aes(x = Rwt, y = log10(mean_size_emerge), group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = palette("set2")) +
  labs(x = "initial R", y = "log10 Size", col = "Increase with Mutation", title = "Size at emergent high transmissibility, spillover R > 1")

#save plot
file = here("results", "analysis", "R greater than 1", "archive", "Rsims", "fig_size_high_transmission.png")
ggsave(filename = file, plot)

plot = ggplot() +
  geom_line(data = run, aes(x = Rwt, y = log10(mean_size_extinct), group = factor(deltaR), col = factor(deltaR))) +
  theme_bw() +
  scale_color_manual(values = palette("set2")) +
  labs(x = "initial R", y = "log10 Size", col = "Increase with Mutation", title = "Size at extinction before high transmissibility, spillover R > 1")

#save plot
file = here("results", "analysis", "R greater than 1", "archive", "Rsims", "fig_size_extinction.png")
ggsave(filename = file, plot)

```
