---
title: "Effect of partial vaccination on disease emergence"
subtitle: "To what extenct does partial vaccination hinder the adaptive evolution of pathogens with subcritical R_0? "
format: html
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(nleqslv) #for solving nonlinear systems of equations
library(gridExtra) #for arrangeing plotsx

source(here("code", "functions", "functions.R")) #for current version of adaptive mutation function
source(here("code", "functions", "archive", "poisson_functions.R")) #for nature version of additive mutation function

# colors for plotting
my.colors = c("black", "red", "forestgreen", "blue", "orange", "purple") #because pretty
```


# Objective

In this document, we explore the probability that pathogen acquires sufficient mutations to achieve $R_0 >1$ during a stuttering chain of transmission. Importantly, we assume that each mutation confers an increase in the $R_0$ of infections with the mutated variant, such that, though all variants except for the fully-adapted variant have subcritical $R_0$, the number of secondary infections expected of a single case increases with each mutation. 

We will investigate how small changes in the $R_0$ of a spillover pathogen can change the probability that it acquires a sufficient number of adaptive mutations to achieve an $R_0 > 1$ and potentiate an epidemic (we will refer to this as "emergence".)

We will then investigate how the change in the probability of emergence *per spillover* of pathogen with a given $R_0$ effects the expected time to emergence of the pathogen, assuming some constant rate of introduction. 


# Model Outline

It is helpful to see how $P_{Emergence}$ depends on the $R_0$ of the initial spillover infection and the increase in fitness conferred by a mutation. The intuition motivating this model follows. 

@AntiaNature2003 show that the probability of mutation is sharply dependent upon the $R_0$ of initial infection, and that this dependence is particularly evident at low mutation rates and with initial $R_0$ close to 1 (see `code\exploratory\Antia2003Models.qmd`.)

They also explored how the probability of emergence decreases with the total number of mutations required to adapt. In their model, mutations have no effect on the $R_0$ on the infection except for the final mutation, and requiring more mutations leads to much steeper dependencies on the $R_0$ of the infection. The intuition here is that a higher $R_0$ results in a longer transmission chain, and more opportunities to acquire the necessary mutations. 

Here, we want to observe the effect that mutations that do confer an increase in $R_0$, though not exceeding $R_0 > 1$, have on the total emergence probability.

In this model, each mutation increases the $R_0$ by a fixed step, such that 1 fewer mutation is required to achieve $R_0 > 1$. As a result, with each mutation, the probability of emergence increases sharply, as acquiring the next mutation becomes more likely. In essence, the $R_0$ of the intitial infection determines how many mutations are required for emergence, and thus, small decreases in $R_0$ can drop the emergence probability to a "new" curve, where another mutation is required for emergence:


```{r model-intution}
## parameters
R0_1s = seq(0.01, 1, by = 0.01) #wt R0s to test
R0_2 = 1.5 #R0 at which human-human transmission is achieved/end of interval
ms = c(2, 3, 4, 5) #how many types? m-1 mutations required, meaning m-2 intermediate types
mu = 10^-3#mutation rate


  ## matrix to store emergence probabilities 
  probs.emergence = matrix(nrow = length(R0_1s), ncol = length(ms)) # vector to store probabilities of emergence
  
  ## simulate
  for(j in 1:length(ms)){
    for(i in 1:length(R0_1s)){
      probs.emergence[i,j] = pEmergence_additive(m = ms[j], mu = mu, R0_1 = R0_1s[i], R0_2 = R0_2) #run simulation for each m,R0_1 combination
    }
  }
  
  ## format matrix for plotting
  probs.emergence = cbind(R0_1s, probs.emergence) %>% as.data.frame() #format as dataframe with column denoting the R0_1 used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
  colnames(probs.emergence) = c("R0", ms) #name the columns for easy referencing
  probs.emergence = probs.emergence %>% #pivot to long-form
    pivot_longer(cols = !R0, #keep R0_1 column
                 names_to = "m", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R0_1 in R0_1 column
  



## plot
linetypes = c("solid", "dashed", "dotted")

# plot
plot = ggplot() + 
  geom_line(data = probs.emergence, aes(x = R0, y = P.Emergence, group = m)) + #plot p emergence as function of R0_1 for each m
  geom_line(data = probs.emergence %>% filter(m == ms[1], R0 > 1-(0.5/3), R0 <= 1), aes(x = R0, y = P.Emergence, group = m), col = "red", size = 2) +
  geom_line(data = probs.emergence %>% filter(m == ms[2], R0 > 1-2*(0.5/3), R0 <= 1-(0.5/3)), aes(x = R0, y = P.Emergence, group = m), col = "red", size = 2) +
  geom_line(data = probs.emergence %>% filter(m == ms[3], R0 > 1-3*(0.5/3), R0 <= 1-2*(0.5/3)), aes(x = R0, y = P.Emergence, group = m), col = "red", size = 2) +
  geom_line(data = probs.emergence %>% filter(m == ms[4], R0 > 1-4*(0.5/3), R0 <= 1-3*(0.5/3)), aes(x = R0, y = P.Emergence, group = m), col = "red", size = 2) +
  ### adding horizontal dashed lines in powerpoint
  theme_bw() + #white background with grey gridlines
  labs(x = "R0 of first infection", y = "Probability of emergence", title = "probability of emergence with m phenotypes") + #axis and title labels
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(trans = "log10", limits = c(10^-7,1)) 

# show plot
print(plot)
```


Below, we plot the curves for the probability of emergence of a variant spilling over with some initial $R_0$, depending on the increase in fitness gained from a mutation. 

```{r delta-r-model, fig.cap = "Probability of emergence increases sharply with $R_0$ as the number of mutations required for emergence decreases. When mutations confer smaller changes in $R_0$, small changes in wildtype $R_0$ result in greater increases in emergence probability.", results = "hide"}
## parameters
R_wts = seq(0, 1, by = 0.01) #wt R0s to test
R_adapted = 1000 #R0 of final phenotype
delta_Rs = c(1, 0.25, 0.1, 0.05)
mus = c(10^-3, 10^-4) #mutation rate

emergence_curves = list() #storage list

for(k in 1:length(mus)){ #creating a seperate dataframe, probs.emergence, for each mu
  mu = mus[k] #set mu
  ## matrix to store emergence probabilities 
  probs.emergence.deltaR = matrix(nrow = length(R_wts), ncol = length(delta_Rs)) # vector to store probabilities of emergence

  ## simulate
  for(j in 1:length(delta_Rs)){
    for(i in 1:length(R_wts)){
      probs.emergence.deltaR[i,j] = pEmergence_deltaR(delta_R = delta_Rs[j], mu = mu, R_wt = R_wts[i], R_adapted = R_adapted) #run simulation for each delta_R,R_wt combination
    }
  }
  
  ## format matrix for plotting
  probs.emergence.deltaR = cbind(R_wts, probs.emergence.deltaR) %>% as.data.frame() #format as dataframe with column denoting the R_wt used to generate the probabilities row-wise and column for each m used to generate the probabilities column-wise
  colnames(probs.emergence.deltaR) = c("R0", as.character(round(delta_Rs, digits = 2))) #name the columns for easy referencing
  emergence_curves[[k]] = probs.emergence.deltaR %>% #pivot to long-form
    pivot_longer(cols = !R0, #keep R_wt column
                 names_to = "delta_R", #add feature denoting "m" used row-wise
                 values_to = "P.Emergence") #add feature denoting the pemergence calculated from m in m column and R_wt in R_wt column

}

## plot
# plot
plot = ggplot() + 
    geom_line(data = emergence_curves[[1]], aes(x = R0, y = P.Emergence, group = delta_R, col = delta_R, linetype = as.character(mus[1]))) + #plot p emergence as function of R_wt for each delta R
    theme_bw() + #white background with grey gridlines
    labs(x = expression("R"[0] ~ " of introduced pathogen"), y = "Probability of Emergence", col = expression(Delta*R), linetype = expression(mu)) + #axis and title labels
    scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
    scale_y_continuous(trans = "log10", limits = c(10^-13,1)) + #log10 y scale
    scale_color_manual(values = my.colors, breaks = as.character(delta_Rs)) + #because pretty +
    scale_linetype_manual(values = c("solid", "dashed"), labels = format(mus, scientific = T))

for(i in 2:length(emergence_curves)){
  plot = plot +  geom_line(data = emergence_curves[[i]], aes(x = R0, y = P.Emergence, group = delta_R, col = delta_R, linetype = as.character(mus[i])), lwd = 1)  #plot p emergence as function of R_wt for each delta r and mu
}

#print plot and save to results 
print(plot)

#save
file = here("results", "analysis", "fig_Emergence_curves_deltaR.jpg") #file path
ggsave(filename = file, plot = plot) #plot save to file path

```

# Probability of emergence decreases with partial vaccination

We assume that vaccination reduces the spillover $R_0$ of a pathogen by the fraction of the population vaccinated, $f_v$, such that $R_{0eff} = (1-f_v)R_{0wt}$. We calculate the fold change in $P_{emergence}$ for a given spillover $R_0$ from emergence in an unvaccinated population to emergence in a population with proportion $f_v$ individuals immune to infection. 

```{r Pemergence-heatmaps, results = "hide"}
# parameters
mus = 10^seq(-6,-3, by = 0.1) #mutation rate
R_wts = seq(0.75,1.0, by = 0.01) #wt R0s to test - we only care about large R_0s at spillover because these actually give rise to stuttering chains of transmission
R_adapted = 1000 #emergence R_0
delta_Rs = c(0.25, 0.1, 0.05) #step increase in R_0 with mutation
f_vs = c(0,.05,.10,.2) #portion of population vaccinated

# combinations of parameters fv and delta R
parms.grid = expand.grid("fv" = f_vs, "delta_R" = delta_Rs) %>% as.data.frame() 

#plot for each m and r combo
myplots <- lapply(1:nrow(parms.grid), function(plot) {
  # set delta_R and portion vaccinated
  delta_R = parms.grid[plot,"delta_R"] # increase in R_0 with mutation
  fv = parms.grid[plot, "fv"] # fraction vaccinated (reduces R0 to R0(1-fv))
  
  # storage matrix 
  P.Es = matrix(nrow = length(R_wts), ncol = length(mus)) #store times to emergence for each R0/mu pair
  
  # calculate delta ps
  for(i in 1:length(R_wts)){
    for(j in 1:length(mus)){
    ## wild-type probability
      P.E = pEmergence_deltaR(delta_R = delta_R , mu = mus[j], R_wt = R_wts[i]*(1-fv), R_adapted = R_adapted) #calculated emergence probability for each combinations
    
     # add solution to matrix
      P.Es[i,j] = P.E #add to matrix for plotting
    }
  }
  
  #format P.Es dataframe
  P.Es = cbind(R_wts, P.Es) %>% as.data.frame() #dataframe with R0s column
  colnames(P.Es) = c("R0", mus) #column names are mutation rates for referencing
  
  #long-form 
  P.Es.long = P.Es %>% #pivots dataframe to longform with 3 columns: R0 of intoduced pathogen, mutation rate, and probability of emergence
    pivot_longer(
      cols = !R0, 
      names_to = "Mut.Rate",
      values_to = "P.Emergence"
    )
  
  # plot heatmap 
  plot = ggplot() +  #plots heatmap of proability of emergence given mutation rate v spillover R_0
          geom_tile(data = P.Es.long, aes(x = R0, y = as.numeric(Mut.Rate), fill = log10(P.Emergence))) + 
          labs(x = "spillover R0", y = "mutation rate", fill = "log10 PE", title = paste0("Fv = ", fv, ", delta_R = ", delta_R)) + #each heatmap assumes a combination of fv and deltaR
          scale_y_continuous(breaks = 10^seq(-1, -5, -1), trans = "log10") +
          scale_fill_gradient(low = "white", high = "blue", limits = c(-15, 0)) #darker = closer to emergence
  plot
}
)


#plot together on one grid
plot = do.call(grid.arrange, list("grobs" = myplots, "nrow" = 4))

#save
file = here("results", "analysis", "fig_PE_heatmaps.jpg") #file path
ggsave(filename = file, plot = plot) #plot save to file path

```

Below, we plot the fold reduction in the probability of emergence from a completely susceptible population to a population with partial vaccination:

```{r FR-Pemergence-heatmaps, results = "hide"}
# parameters
mus = 10^seq(-6,-3, by = 0.1) #mutation rate
R_wts = seq(0.75,1.0, by = 0.01) #wt R0s to test
R_adapted = 1000 #final phenotype R_0
delta_Rs = c(0.25, 0.1, 0.05) #step increase in R_0 with mutation
f_vs = c(.05,.10,.2) #portion of population vaccinated

# combinations of parameters 
parms.grid = expand.grid("fv" = f_vs, "delta_R" = delta_Rs) %>% as.data.frame() 

#plot for each m and r combo
myplots <- lapply(1:nrow(parms.grid), function(plot) {
  # set deltaR and fv
  delta_R = parms.grid[plot,"delta_R"] # number of possible phenotypes
  fv = parms.grid[plot, "fv"] # fraction vaccinated (reduces R0 to R0(1-fv))
  
  # storage matrix 
  FR.Ps = matrix(nrow = length(R_wts), ncol = length(mus)) #store times to emergence for each R_wt/mu pair
  
  # calculate delta ps
  for(i in 1:length(R_wts)){
    for(j in 1:length(mus)){
    ## wild-type probability
      pWT = pEmergence_deltaR(delta_R = delta_R, mu = mus[j], R_wt = R_wts[i], R_adapted = R_adapted) #calculated emergence probability with spillover wt R0
      
      ## vaccination probability
      pVax = pEmergence_deltaR(delta_R = delta_R, mu = mus[j], R_wt = R_wts[i]*(1-fv), R_adapted = R_adapted) #calculated emergence probability for spillover R0 in vaccinated population
    
     # add solution to matrix
      FR.Ps[i,j] = pWT / pVax #add to matrix for plotting
    }
  }
  
  #format P.Es dataframe
  FR.Ps = cbind(R_wts, FR.Ps) %>% as.data.frame() #dataframe with R0s column
  colnames(FR.Ps) = c("R0", mus) #column names for referencing
  
  #long-form 
  FR.Ps.long = FR.Ps %>%
    pivot_longer(
      cols = !R0, 
      names_to = "Mut.Rate",
      values_to = "FR.P"
    )
  
  # plot heatmap 
  plot = ggplot() + 
          geom_tile(data = FR.Ps.long, aes(x = R0, y = as.numeric(Mut.Rate), fill = log10(FR.P))) + 
          labs(x = "spillover R0", y = "mutation rate", fill = "log10 FR.P", title = paste0("Fv = ", fv, ", delta_R = ", delta_R)) +
         scale_y_continuous(breaks = 10^seq(-1, -5, -1), trans = "log10") +
  scale_fill_gradient(low = "white", high = "blue", limits = c(-1, 10)) 
  plot
}
)


#plot together on one grid
plot = do.call(grid.arrange, list("grobs" = myplots, "nrow" = 3))

#save
file = here("results", "analysis", "fig_FR_PE_heatmaps.jpg") #file path
ggsave(filename = file, plot = plot) #plot save to file path

```

# Times to emergence

We are not concerned with a pathogen whose expected waiting time to emergence is very large (these are unlikely to cause a problem more quickly than we notice a problem is occurring and intervene); conversely, there isn't too much we can do for pathogens whose expected time to emergence is very short (these are essentially "off - the - shelf" pathogens). Thus, we are most interested in observing how partial vaccination increases the emergence time of pathogens with intermediate emergence times in an unvaccinated population. 

As a note, the total probability of emergence of a pathogen is given by the probability of emergence *per introduction* times the number of introductions. Assuming a constant rate of spillovers, $\rho$, the waiting time to emergence for a pathogen with a given probability of emergence per introduction ($P_{Emergence}$) is exponentially distributed with an expected waiting time of $\frac{1}{\rho P_{Emergence}}$.

Below, we plot the expected waiting time for pathogens as a function of the spillover $R_0$, mutation rate, $\mu$, and increase in $R_0$ with mutation, $\Delta R$, as the level of population vaccination $f_v$ increases. The red contours show the range of intermediate $T_{Emergence}$, on the order of months to years. 

```{r Temergence-heatmaps, results = "hide"}
# parameters
mus = 10^seq(-6,-3, by = 0.1) #mutation rate
R_wts = seq(0.75,1.0, by = 0.01) #wt R0s to test
R_adapted = 1000 #final phenotype R_0
delta_Rs = c(0.25, 0.1, 0.05) #step increase in R_0 with mutation
f_vs = c(0,.05,.10,.2) #portion of population vaccinated
rho = 10^3 #spillover rate

# combinations of parameters 
parms.grid = expand.grid("fv" = f_vs, "delta_R" = delta_Rs) %>% as.data.frame() 

#plot for each m and r combo
myplots <- lapply(1:nrow(parms.grid), function(plot) {
  # set m and r
  delta_R = parms.grid[plot,"delta_R"] # number of possible phenotypes
  fv = parms.grid[plot, "fv"] # fraction vaccinated (reduces R0 to R0(1-fv))
  
  # storage matrix 
  T.Es = matrix(nrow = length(R_wts), ncol = length(mus)) #store times to emergence for each R_wt/mu pair
  
  # calculate delta ps
  for(i in 1:length(R_wts)){
    for(j in 1:length(mus)){
    ## wild-type probability
      T.E = 1/(pEmergence_deltaR(delta_R = delta_R, mu = mus[j], R_wt = R_wts[i]*(1-fv), R_adapted = R_adapted)*rho) #calculated emergence probability for each combinations
    
     # add solution to matrix
      T.Es[i,j] = T.E #add to matrix for plotting
    }
  }
  
  #format T.Es dataframe
  T.Es = cbind(R_wts, T.Es) %>% as.data.frame() #dataframe with R0s column
  colnames(T.Es) = c("R0", mus) #column names for referencing
  
  #long-form 
  T.Es.long = T.Es %>%
    pivot_longer(
      cols = !R0, 
      names_to = "Mut.Rate",
      values_to = "T.Emergence"
    )
  
  # plot heatmap 
  plot = ggplot() + 
          geom_tile(data = T.Es.long, aes(x = R0, y = as.numeric(Mut.Rate), fill = log10(T.Emergence))) + 
          labs(x = "R0 of introduced pathogen", y = "mutation rate", fill = "log10 TE", title = paste0("Fv = ", fv, ", delta_R = ", delta_R)) +
          geom_contour(data = T.Es.long, aes(x = R0, y = as.numeric(Mut.Rate), z = log10(T.Emergence)), 
                   breaks = c(-1,1), col = 'maroon') + 
          scale_y_continuous(breaks = 10^seq(-1, -5, -1), trans = "log10") +
          scale_fill_gradient(low = "white", high = "blue") 
  plot
}
)


#plot together on one grid
plot = do.call(grid.arrange, list("grobs" = myplots, "nrow" = 3))

#save
file = here("results", "analysis", "fig_TE_heatmaps.jpg") #file path
ggsave(filename = file, plot = plot) #plot save to file path

```
