---
title: "Exploring Multi-type Branching Processes"
subtitle: "Recapitulating the results of Antia 2003"
format: pdf
execute:
  echo: false
  warning: false
bibliography: ../../assets/bibFiles/dataanalysis-references.bib
csl: ../../assets/bibFiles/apa.csl
---

```{r setup}
# load libraries 
library(ggplot2) #for plotting
library(here) #for easy local path referencing
library(tidyverse) #for easy manipulation of data structures
library(nleqslv) #for solving nonlinear systems of equations

#load functions
source(here("code", "functions", "Nature_2003_functions.R"))

## colors for plotting
my.colors = c("black", "orange", "red", "forestgreen", "blue")
```

# Objective

We are interested in exploring the probability of emergence of zoonotic disease in a population immune to a cross-reactive relative of the pathogen, and, in particular, how the emergence probability changes with the level of immunity in the population.

When the first infection caused by a spillover event from the animal reservoir into humans has an $R_0 < 1$, a short chain of transmission may stochastically occur before the pathogen goes extinct in the human population. 

During these stuttering chains of human-human transmission, selection may act on mutations acquired during infection and the zoonotic pathogen may evolve a higher $R_0$. 

These ideas were initially explored in @AntiaNature2003, wherein the probability of disease emergence as a function of 1) mutation rate, 2) number of mutations required for specialization to the new host, and 3) fitness of intermediate mutations, were modeled using a multi-type branching process framework. 

The goal of this document is to develop and plot the same model in @AntiaNature2003. We will explore the biological motivation for the model, and how it can be modified to explore different questions. 

# Single mutation model (2-type branching process)

Epidemics emerge from zoonoses when the pathogen spills over into the human population and its subsequent spread is maintained in the population. For the latter to occur, the pathogen must have a basic reproductive number, $R_0$, greater than 1. The $R_0$ of the first infection may already be greater than 1, such that the pathogen transmits well in humans immediately after spillover. This may not need to be the case. @AntiaNature2003 explores how pathogens may evolve an $R_0 > 1$ during stochastic transmission in humans even when the $R_0$ of the first infection is less than 1. 

In the simplest case, only 1 mutation is required for the $R_0$ of the pathogen to exceed 1. The stochastic chain of infections after spillover in this scenario can be modeled with a 2-type branching process, with type $1$ corresponding to infections cause by the wild-type (spillover) variant and type $2$ corresponding to infections caused by the human-adapted variant, where: 
  
- the total number of secondary infections arising from an individual infected with variant $1$ is Poisson-distributed with mean $R_0^{(1)}$ (the basic-reproductive number of infections with the wild-type pathogen);
- the number of secondary infections of type $1$ from an infection of type $1$ is poisson-distributed with mean $(1-\mu)R_0^{(1)}$, and
- the number of secondary infections of type $2$, the adapted type, from an infection of type $1$ is poisson-distributed with mean $\mu R_0^{(1)}$, and 
- the total number of secondary infections arising from an individual infected with variant $2$ is Poisson-distributed with mean $R_0^{(2)}$ (the basic-reproductive number of infections with the human-adapted pathogen). Note that infections with the human-adapted pathogen, type $2$, *only* give rise to other type $2$ infections with mean $R_0^{(2)}$.

## Figure 2a

The probability that the pathogen evolves to give rise to a human-adapted infection, type $2$, before extinction depends strongly on the $R_0$ of the introduced pathogen:

```{r fig-2a, results = "hide"}
#single mutation required for evolution 
## parms 
R0_1s = seq(0,1, by = 0.01) # R_0 of infection with introduced pathogen, type 1 infections
Rfinal = 1 #  R_0 of human-adapted type 2 infection
mus = c(10^-1, 10^-2, 10^-3, 10^-4) # mus to test 

# matrix to store emergence probabilities
probs.emergence = matrix(nrow = length(R0_1s), ncol = length(mus)) #rows correspond to R0_1, cols to mu

# solve for emergence probability for each mu, R0_1 combination
for(j in 1:length(mus)){
  #set mu
  mu = mus[j] 
  for(i in 1:length(R0_1s)){
   # set R0_1
    R0_1 = R0_1s[i]
   # solve for emegence probability and add to probs.emergene matrix
   probs.emergence[i,j] = pEmergence_single(mu = mu, R0_1 = R0_1, Rfinal = Rfinal) # loaded from Nature_2003_functions.R
  }
}

### prepare probs.emergence for plotting
# format wide df
probs.emergence = cbind(R0_1s, probs.emergence) %>% as.data.frame() #add R01_s column and format as df
colnames(probs.emergence) = c("R0", as.character(formatC(mus, format = "e", digits = 1))) # add mu names to columns

# pivot to longform 
probs.emergence.long = probs.emergence %>% # columns R0, mu, and Prob.emergence give Prob emergence for each combination of R0_1 and mu
 pivot_longer(cols = !R0,
  names_to = "mu",
  values_to = "Prob.Emergence"
)

# Plot the probability of emergence as a function of the initial R_0, R_01, for each mutation rate
ggplot() + 
  geom_line(data = probs.emergence.long, aes(x = R0, y = Prob.Emergence, group = mu, col = mu)) + #line plot
  theme_bw() + #sets background color to white and grey gridlines
  labs(title = "fig 2a", x = "R0 of introduced pathogen", y = "Probability of emergence") + #axis labels
  scale_y_continuous(trans = "log10") + #log10 y axis
  scale_color_manual(values = my.colors) #use predefined color vector from r setup chunk

```

Note that "probability of evolution" is slightly different from "probability of emergence" - we will get to this at some point. 

While the emergence probability scales almost linearly with $\mu$, the emergence probability is strongly dependent upon the $R_0$ of the first infection, $R_0^{(1)}$ (which is the same as the $R_0$ of all infections of type $1$). The emergence probability increases sharply with high $R_0$ particularly for low $\mu$. 

## Fig 2b

Even with an $R_0 > 1$, a transmission chain can still stochastically go extinct. The probability of emergence - that is, that an epidemic is achieved before the transmission chain goes extinct - is thus dependent upon the $R_0$ of the human-adapted strain, or the $R_0$ of infections of type $2$, $R_0^{(2)}$.

Below, we plot the extinction probabilities of transmission chains as a function of the $R_0^{(1)}$ of the introduced pathogen when the $R_0^{(2)}$ is either $1,000, 1.5,$ or $1.2$, corresponding to the solid, dashed, and dotten lines respectively. We plot these three curves for two mutation rates $(10^{-1})$ and $10^{-3}$.


```{r fig-2b, results = "hide"}
#single mutation required for evolution 
## parms 
R0_1s = seq(0,1, by = 0.01) # R_0 of infection with introduced pathogen, type 1 infections
Rfinals = c(10^3,1.5,1.2) #  R_0 of human-adapted type 2 infection
mus = c(10^-1, 10^-3) # mus to test 

#list to store matrices of emergence probabilities calculated with each mu
probs.emergence.list = list()

# simulate - will result in two dfs of emergence probabiliies v R0_1 for different Rfinals corresponding to the mu assumed while running simulations
c = 1 # count refers to which mu to assume in vector of mus, and thus the element of the list to which we will be saving
while(c <= length(mus)){
  #set mutation rate
    mu = mus[c] 
  # matrix to store emergence probabilities for R_01 and Rfinal combinations
  probs.emergence = matrix(nrow = length(R0_1s), ncol = length(Rfinals)) #rows correspond to R0_1, cols to Rfinal
  # solve for emergence probability for each Rfinal, R0_1 combination
    for(j in 1:length(Rfinals)){
      #set Rfinal
      Rfinal = Rfinals[j] 
      for(i in 1:length(R0_1s)){
       # set R0_1
        R0_1 = R0_1s[i]
       # solve for emegence probability and add to probs.emergene matrix
       probs.emergence[i,j] = pEmergence_single(mu = mu, R0_1 = R0_1, Rfinal = Rfinal) # loaded from Nature_2003_functions.R
      }
    }
  
  ### prepare probs.emergence for plotting
  # format wide df
  probs.emergence = cbind(R0_1s, probs.emergence) %>% as.data.frame() #add R01_s column and format as df
  colnames(probs.emergence) = c("R0", as.character(Rfinals)) # add mu names to columns
  
  # pivot to longform and store in df list
  probs.emergence.list[[c]] = probs.emergence %>% # columns R0, mu, and Prob.emergence give Prob emergence for each combination of R0_1 and mu
   pivot_longer(cols = !R0,
    names_to = "Rfinal",
    values_to = "Prob.Emergence"
  )

c = c+1 #next mu  
}


#plot
ggplot() + 
  geom_line(data = probs.emergence.list[[1]], aes(x = R0, y = Prob.Emergence, group = Rfinal, linetype = Rfinal, col = "10^-1")) +
  geom_line(data = probs.emergence.list[[2]], aes(x = R0, y = Prob.Emergence, group = Rfinal, linetype = Rfinal, col = "10^-3")) +
  theme_bw() + #sets background color to white and grey gridlines
  labs(title = "fig 2b", x = "R0 of introduced pathogen", y = "Probability of emergence", col = "mu", linetype = "R0 of adapted") + #axis labels
  scale_y_continuous(trans = "log10") +
  scale_linetype_manual(values = c("1.2" = "dotted", "1.5" = "dashed", "1000" = "solid")) +
  scale_color_manual(values = c("10^-1" = "black", "10^-3" = "forestgreen")) 

```

# Multipe-step mutation model (m-type branching process)

The single-mutation model can be extended to a model that considers how the probability of emergence depends on the number of evolutionary changes required for the pathogen to attain $R_0 > 1$. 

This process can be captured by a branching process with more than 2 types, corresponding to the number of variants that exist in the evolutionary path from the wild-type variant that causes the first infection to the fully adapted variant with an $R_0 > 1$. 

## fig 3a

In figure 3a of @AntiaNature2003, the emergence probabilities are calculated as a function of the $R_0$ of the spillover variant, $R_0^{(1)}$ as the number of steps required to adapt increases. This model **assumes that the fitness of all intermediate variants is the same as the $R_0$ of the spillover variant**, that is, that $R_0^{(1)} = R_0^{(2)} = ... = R_0^{(m-1)}$**, and only the fully adapted variant has an $R_0$ that increases such that $R_0^{(m)} > 1$.

The model also assumes that infections with variant $i$ only give rise to infections of the same type, $i$, or of the type with one more mutation, $i+1$, and: 
  - the total number of secondary infections arising from an individual infected with variant $i$ is Poisson-distributed with mean $R_0^{(1)}$ (the        basic-reproductive number of infections with the wild-type pathogen);
  - the number of secondary infections of type $i$ from an infection of type $i$ is poisson-distributed with mean $(1-\mu)R_0^{(i)}$, and
  - the number of secondary infections of type $i+1$, with one more mutation, from an infection of type $i$ is poisson-distributed with mean $\mu          R_0^{(i)}$, and 
  - the total number of secondary infections arising from an individual infected with variant $m$ is Poisson-distributed with mean $R_0^{(m)}$ (the        basic-reproductive number of infections with the adapted-type pathogen)
  
It should be noted that the paper refers to $n$ to indicate the number of intermediate variants between type $1$ and type $m$, such that:

- $n=0$ corresponds to a model with m=2 types, the wild type with $R_0^{(1)} < 1$ and the adapted type with $R_0^{(2)} > 1$, and 
- $n = 4$ corresponds to a model with m=6 types, the wild type with $R_0^{(1)}<1$, 4 intermediate types with $R_0^{(i)} = R_0^{(1)}$ for $i \in \{2...5\}$ and the adapted type with $R_0^{(6)} > 1$

I'll use the same notation ($n$) as in the paper for clarity. 

```{r fig-3a, results = "hide"}
## parms
R0_1s = seq(0,1.2, by = 0.01) # R_0 of infection with introduced pathogen and all type 1:(m-1) infections
Rfinal = 1.2 #  R_0 of human-adapted type m infection
mu = 10^-1 # mutation rate
ns = 0:4 #m = 2:6

# matrix to store emergence probabilities
probs.emergence = matrix(nrow = length(R0_1s), ncol = length(ns)) #rows correspond to R0_1, cols to number of intermediats

# solve for emergence probability for each mu, R0_1 combination
for(j in 1:length(ns)){
  #set mu
  n = ns[j] 
  for(i in 1:length(R0_1s)){
   # set R0_1
    R0_1 = R0_1s[i]
   # solve for emegence probability and add to probs.emergene matrix
   probs.emergence[i,j] = pEmergence_jackpot(n = n, mu = mu, R0_1 = R0_1, Rfinal = Rfinal) # loaded from Nature_2003_functions.R
  }
}

### prepare probs.emergence for plotting
# format wide df
probs.emergence = cbind(R0_1s, probs.emergence) %>% as.data.frame() #add R01_s column and format as df
colnames(probs.emergence) = c("R0", as.character(ns)) # add n names to columns

# pivot to longform 
probs.emergence.long = probs.emergence %>% # columns R0, mu, and Prob.emergence give Prob emergence for each combination of R0_1 and mu
 pivot_longer(cols = !R0,
  names_to = "n",
  values_to = "Prob.Emergence"
)

# Plot the probability of emergence as a function of the initial R_0, R_01, for each mutation rate
ggplot() + 
  geom_line(data = probs.emergence.long, aes(x = R0, y = Prob.Emergence, group = n, col = n)) + #line plot
  theme_bw() + #sets background color to white and grey gridlines
  labs(title = "fig 3a", x = "R0 of introduced pathogen", y = "Probability of emergence", col = "# intermediates, n") + #axis labels
  scale_y_continuous(trans = "log10", limits = c(10^-5,1)) + #log10 y axis
  scale_color_manual(values = my.colors) #use predefined color vector from r setup chunk


```

## fig 3b

In figure 3b, the assumption that the $R_0$ of all intermediates is the same as the $R_0$ of the introduced pathogen is relaxed. Instead, three different fitness landscapes are explored: 

1) the jackpot landscape, where only the final mutation offers an increase in $R_0$ from the $R_0$ of the introduced pathogen; 
2) the additive landscape, where the intermediate mutations increase $R_0$ by a fixed amount until the final mutation shifts $R_0 > 1$, and
3) the fitness valley landscape, where the intermedate mutations can decrease $R_0$ 

These models are only explored for the $n=1$ intermediate case, that is, for a 3-type branching process with wildtype, intermediate, and adapted types. 
The function in `\code\functions\Nature_2003_functions.R` extends these to the m-type case. 

```{r fig-3b, results = "hide"}



```


